```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")

library(dplyr)
library(forcats)
library(readr)
library(ggridges) # for geom_density_line
```


# Avoid line drawings


However, the figure still needs to satisfy the principle of proportional ink (Chapter \@ref(proportional-ink)).

(ref:titanic-ages-lines) Histogram of the ages of Titanic passengers.

```{r titanic-ages-lines, message=FALSE, warning=FALSE, fig.cap='(ref:titanic-ages-lines)'}
df <- read_csv("datasets/Titanic.csv")
titanic <- na.omit(data.frame(age = df$Age, sex = df$Sex, survived = df$Survived))
age_hist_3 <- data.frame(age = (1:25)*3-1.5, 
                         count = hist(titanic$age, breaks=(0:25)*3 + .01, plot = FALSE)$counts)

h3_bad <- ggplot(age_hist_3, aes(x = age, y = count)) + geom_col(width = 2., fill = "transparent",
                                                             color = "black")  + 
  scale_y_continuous(limits = c(0, 86), expand = c(0, 0), breaks = 25*(0:5)) +
  scale_x_continuous(limits = c(0, 75), expand = c(0, 0)) +
  theme_dviz() +
  background_grid(major = "y", minor = "none")

stamp_bad(h3_bad)
```

(ref:titanic-ages-filled) Histogram of the ages of Titanic passengers.

```{r titanic-ages-filled, message=FALSE, warning=FALSE, fig.cap='(ref:titanic-ages-filled)'}
h3_good <- ggplot(age_hist_3, aes(x = age, y = count)) + geom_col(width = 2.7, fill = "#56B4E9")  + 
  scale_y_continuous(limits = c(0, 86), expand = c(0, 0), breaks = 25*(0:5)) +
  scale_x_continuous(limits = c(0, 75), expand = c(0, 0)) +
  theme_dviz() +
  background_grid(major = "y", minor = "none")

stamp_good(h3_good)
```

(ref:iris-densities-lines) Density estimates of the sepal lengths of three different iris species.

```{r iris-densities-lines, message=FALSE, warning=FALSE, fig.cap='(ref:iris-densities-lines)'}
# compute densities for sepal lengths
iris_dens <- group_by(iris, Species) %>%
  do(ggplot2:::compute_density(.$Sepal.Length, NULL)) %>%
  rename(Sepal.Length = x)

# get the maximum values
iris_max <- filter(iris_dens, density == max(density)) %>%
  ungroup() %>%
  mutate(hjust = c(0, 0.4, 0),
         vjust = c(1, 0, 1),
         nudge_x = c(0.11, 0, 0.2),
         nudge_y = c(-0.02, 0.02, -0.02) 
        )

iris_lines <- ggplot(iris_dens, aes(x = Sepal.Length, y = density, linetype = Species)) + 
  geom_density_line(stat = "identity", alpha = 0., size = 0.75) +
  geom_text(data = iris_max, aes(label = Species, hjust = hjust, vjust = vjust,
                                 x = Sepal.Length + nudge_x, 
                                 y = density + nudge_y), size = 12/.pt) +
  scale_linetype_manual(values = c(1, 3, 4),
                    breaks = c("virginica", "versicolor", "setosa"),
                    guide = "none") +
  scale_x_continuous(expand = c(0, 0), name = "sepal length") +
  scale_y_continuous(limits = c(0, 1.5), expand = c(0, 0)) +
  theme_dviz()

stamp_bad(iris_lines)
```


(ref:iris-densities-colored-lines) Density estimates of the sepal lengths of three different iris species.

```{r iris-densities-colored-lines, message=FALSE, warning=FALSE, fig.cap='(ref:iris-densities-colored-lines)'}
iris_colored_lines <- ggplot(iris_dens, aes(x = Sepal.Length, y = density, color = Species)) + 
  geom_density_line(stat = "identity", alpha = 0., size = 1) +
  geom_text(data = iris_max, aes(label = Species, hjust = hjust, vjust = vjust,
                                 x = Sepal.Length + nudge_x, 
                                 y = density + nudge_y), size = 12/.pt) +
  scale_color_manual(values = c("#56B4E9", "#E69F00", "#009E73"),
                    breaks = c("virginica", "versicolor", "setosa"),
                    guide = "none") +
  scale_x_continuous(expand = c(0, 0), name = "sepal length") +
  scale_y_continuous(limits = c(0, 1.5), expand = c(0, 0)) +
  theme_dviz()

stamp_ugly(iris_colored_lines)
```


(ref:iris-densities-filled) Density estimates of the sepal lengths of three different iris species.

```{r iris-densities-filled, message=FALSE, warning=FALSE, fig.cap='(ref:iris-densities-filled)'}
iris_filled <- ggplot(iris_dens, aes(x = Sepal.Length, y = density, fill = Species)) + 
  geom_density_line(stat = "identity", alpha = 0.7) +
  geom_text(data = iris_max, aes(label = Species, hjust = hjust, vjust = vjust,
                                 x = Sepal.Length + nudge_x, 
                                 y = density + nudge_y), size = 12/.pt) +
  scale_fill_manual(values = c("#56B4E9", "#E69F00", "#009E73"),
                    breaks = c("virginica", "versicolor", "setosa"),
                    guide = "none") +
  scale_x_continuous(expand = c(0, 0), name = "sepal length") +
  scale_y_continuous(limits = c(0, 1.5), expand = c(0, 0)) +
  theme_dviz()
  
stamp_good(iris_filled)
```

(ref:mpg-linespoints) Fuel economy versus engine displacement, for cars with front-wheel drive (FWD), rear-wheel drive (RWD), and all-wheel drive (4WD).

```{r mpg-linespoints, fig.width=5, fig.asp=.7416, fig.cap='(ref:mpg-linespoints)'}
mpg_linespoints <- ggplot(mpg, aes(y=cty, x=displ, shape=drv)) +
  geom_point(size = 2) + 
  ylab("fuel economy (mpg)") +
  xlab("displacement (l)") +
  scale_shape_manual(values=c(4, 1, 2), 
                     name="drive train",
                     breaks=c("f", "r", "4"),
                     labels=c("FWD", "RWD", "4WD")) +
  theme(legend.position = c(.7, .8))
stamp_bad(mpg_linespoints)
```


(ref:mpg-filledpoints) Fuel economy versus engine displacement.

```{r mpg-filledpoints, fig.width=5, fig.asp=.7416, fig.cap='(ref:mpg-filledpoints)'}
mpg_filledpoints <- ggplot(mpg, aes(y=cty, x=displ, color=drv, shape=drv)) +
  geom_point(size = 2.5) + 
  ylab("fuel economy (mpg)") +
  xlab("displacement (l)") +
  scale_color_manual(values=c("#000000", "#E69F00", "#56B4E9"), 
                     name="drive train",
                     breaks=c("f", "r", "4"),
                     labels=c("FWD", "RWD", "4WD")) +
  scale_shape_manual(values=c(18, 16, 17), 
                     name="drive train",
                     breaks=c("f", "r", "4"),
                     labels=c("FWD", "RWD", "4WD")) +
  theme(legend.position = c(.7, .8))

stamp_good(mpg_filledpoints)
```


(ref:lincoln-weather-box-empty) *Figure caption goes here.*

```{r lincoln-weather-box-empty, fig.cap='(ref:lincoln-weather-box-empty)'}
lincoln_weather %>% mutate(month_short = fct_recode(Month,
                                                    Jan = "January",
                                                    Feb = "February",
                                                    Mar = "March",
                                                    Apr = "April",
                                                    May = "May",
                                                    Jun = "June",
                                                    Jul = "July",
                                                    Aug = "August",
                                                    Sep = "September",
                                                    Oct = "October",
                                                    Nov = "November",
                                                    Dec = "December")) %>%
  mutate(month_short = fct_rev(month_short)) -> lincoln_df

lincoln_box_empty <- ggplot(lincoln_df, aes(x = month_short, y = `Mean Temperature [F]`)) +
  geom_boxplot() + xlab("Month")

lincoln_box_empty
```


(ref:lincoln-weather-box-filled) *Figure caption goes here.*

```{r lincoln-weather-box-filled, fig.cap='(ref:lincoln-weather-box-filled)'}
lincoln_box_filled <- ggplot(lincoln_df, aes(x = month_short, y = `Mean Temperature [F]`)) +
  geom_boxplot(fill = 'grey90') + xlab("Month")

lincoln_box_filled
```
