```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")
```

# Choosing the right plotting software


Throughout much of the 20th century, data visualizations were drawn by hand, mostly by technical illustrators who did so for a living. 

One notable example is Randall Munroe, creator of the web comic XKCD, 

```{block type='rmdtip', echo=TRUE}
The best plotting software is the one that allows you to make the visualizations you need.
```




## Reproducibility and repeatability

In the context of scientific experiments, we refer to work as reproducible if the overarching scientific finding of the work will remain unchanged if a different research group performs the same type of study. For example, if one research group finds that a new pain medication reduces perceived headache pain significantly without causing noticeable side effects and a different group subsequently studies the same medication on a different patient group and has the same findings, then the work is reproducible. By contrast, work is repeatable if very similar or identical measurements can be obtained by the same person repeating the exact same measurement procedure on the same equipment. For example, if I weigh my dog and find she weighs 41 lbs and then I weigh her again on the same scales and find again that she weighs 41 lbs, then this measurement is repeatable.


We can apply these concepts to data visualizations with minor modifications. A visualization is reproducible if the plotted data are available and any data transformations that may have been applied are exactly specified. For example, if you make a figure and then send me the exact data that you plotted, then I can prepare a figure that looks substantially similar. We may be using slightly different fonts or colors or point sizes to display the same data, so the two figures may not be exactly identical, but your figure and mine convey the same message and therefore are reproductions of each other. A visualization is repeatable, on the other hand, if it is possible to recreate the exact same visual appearance, down to the last pixel, from the raw data. Strictly speaking, repeatability requires that even if there are random elements in the figure, such as jitter (Chapter \@ref(overlapping-points)), those elements were specified in a repeatable way and can be regenerated at a future date. For random data, repeatability generally requires that we set and record a seed for the random number generator.

**Make an example of a figure that reproduces but doesn't repeat another figure. Use a different theme, colors, point sizes, etc. Maybe also use altered jitter. And also refer to Chapter \@ref(avoid-line-drawings) which has several more such examples.**


## Separation of content and design

A good plotting software allows you to think separately about content and design. 

**Refer back to the figure from the previous subsection which shows the same content using different designs.**

(ref:unemploy-themes) Number of unemployed persons in the U.S. from 1970 to 2015. The same figure is displayed using four different themes that are part of the R package ggplot2: (a) theme gray (the ggplot2 default); (b) theme linedraw; (c) theme classic; (d) theme minimal.  Data source: U.S. Bureau of Labor Statistics

```{r unemploy-themes, fig.width = 8.5, fig.cap = '(ref:unemploy-themes)'}
unemploy_plot <- ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line(color = "#0072B2", size = .5) +
  scale_y_continuous(name = "unemployed (x1000)",
                     limits = c(0, 17000),
                     breaks = c(0, 5000, 10000, 15000),
                     labels = c("0", "5000", "10,000", "15,000"),
                     expand = c(0, 0)) +
  scale_x_date(name = "year",
               expand = c(0.01, 0))

plot_grid(unemploy_plot + theme_gray(),
          unemploy_plot + theme_linedraw(),
          unemploy_plot + theme_classic(),
          unemploy_plot + theme_minimal(),
          labels = "auto", label_fontface = "plain",
          scale = 1)
```

## Rapid iteration
