```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")
```

# Effective use of color in figures




## Color as a tool to highlight

(ref:Aus-athletes-track) Track athletes are among the shortest and leanest of male professional athletes participating in popular sports.

```{r Aus-athletes-track, fig.width = 7, fig.cap='(ref:Aus-athletes-track)'}
male_Aus <- filter(Aus_athletes, sex=="m") %>%
  filter(sport %in% c("basketball", "field", "swimming", "track (400m)",
                      "track (sprint)", "water polo")) %>%
  mutate(sport = case_when(sport == "track (400m)" ~ "track",
                           sport == "track (sprint)" ~ "track",
                           TRUE ~ sport))

male_Aus$sport <- factor(male_Aus$sport,
                         levels = c("track", "field", "water polo", "basketball", "swimming"))

colors <- c("#BD3828", rep("#808080", 4))
fills <- c("#BD3828D0", rep("#80808080", 4))

ggplot(male_Aus, aes(x=height, y=pcBfat, shape=sport, color = sport, fill = sport)) +
  geom_point(size = 3) +
  scale_shape_manual(values = 21:25) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = fills) +
  xlab("height (cm)") +
  ylab("% bodyfat") +
  theme_minimal_grid()
```

## Color as a tool to distinguish

(ref:popgrowth-US) Population growth in the US from 2000 to 2010. States in the West and South have seen the largest increases, whereas states in the Midwest and Northeast have seen much smaller increases or even, in the case of Michigan, a decrease.

```{r popgrowth-US, fig.width = 6.5, fig.asp = 1.2, fig.cap = '(ref:popgrowth-US)'}
popgrowth_df <- left_join(US_census, US_regions) %>%
    group_by(region, division, state) %>%
    summarize(pop2000 = sum(pop2000, na.rm = TRUE),
              pop2010 = sum(pop2010, na.rm = TRUE),
              popgrowth = (pop2010-pop2000)/pop2000,
              area = sum(area)) %>%
    arrange(popgrowth) %>%
    ungroup() %>%
    mutate(state = factor(state, levels = state),
           region = factor(region, levels = c("West", "South", "Midwest", "Northeast")))

# make color vector in order of the state
region_colors <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442")
region_colors_dark <- darken(region_colors, 0.4)
state_colors <- region_colors_dark[as.numeric(popgrowth_df$region[order(popgrowth_df$state)])]

ggplot(popgrowth_df, aes(x = state, y = 100*popgrowth, fill = region)) + 
  geom_col() + 
  scale_y_continuous(limits = c(-.6, 37.5), expand = c(0, 0),
                     name = "population growth (%)\n2000 to 2010    ") +
  scale_fill_manual(values = region_colors) +
  coord_flip() + 
  theme_minimal_vgrid() +
  theme(axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.length = unit(0, "pt"),
        axis.text.y = element_text(size = 10, color = state_colors),
        legend.position = c(.56, .68),
        #legend.text = element_text(color = region_colors),
        legend.background = element_rect(fill = "#ffffffb0"))
```


(ref:qualitative-scales) Example qualitative color scales. The Okabe Ito scale is the default scale used throughout this book.

```{r qualitative-scales, fig.width=8.5, fig.asp=3*.14, fig.cap = '(ref:qualitative-scales)'}
p1 <- gg_color_swatches(7) + scale_fill_OkabeIto() + ggtitle("Okabe Ito")
p2 <- gg_color_swatches(7) + scale_fill_brewer(type = "qual", palette = "Dark2") + ggtitle("Colorbrewer Dark2")
p3 <- gg_color_swatches(7) + scale_fill_hue() + ggtitle("ggplot2 default")
plot_grid(p1, p2, p3, ncol = 1)
```


## Color to represent of a value

(ref:map-Texas-income) Median income in Texas counties.

```{r map-Texas-income, fig.width = 7, fig.asp = 0.75, fig.cap = '(ref:map-Texas-income)'}
library(sf)

# B19013_001: Median household income in the past 12 months (in 2015 Inflation-adjusted dollars)

texas_income %>%
  ggplot(aes(fill = estimate)) + 
  geom_sf(color = "white") + 
  coord_sf(datum = NA) +
  theme_map() + 
  scale_fill_continuous_sequential(palette = "Teal", rev = TRUE, 
                                   na.value = "grey50", end = 0.9,
                                   name = "income",
                                   limits = c(18000, 90000),
                                   breaks = 20000*c(1:4),
                                   labels = c("$20,000", "$40,000", "$60,000", "$80,000")) +
  theme(legend.title.align = 0.5,
        legend.text.align = 0,
        legend.position = c(0.07, 0.8))
```

(ref:sequential-scales) Example sequential color scales.

```{r sequential-scales, fig.width=8.5, fig.asp=3*.14, fig.cap = '(ref:sequential-scales)'}
p1 <- gg_color_swatches(7) + scale_fill_brewer(type = "seq", palette = "Blues", direction = -1) + ggtitle("Colorbrewer Blues")
p2 <- gg_color_swatches(7) + scale_fill_discrete_sequential("Heat") + ggtitle("Heat")
p3 <- gg_color_swatches(7) + scale_fill_viridis_d() + ggtitle("Viridis")
plot_grid(p1, p2, p3, ncol = 1)
```


(ref:map-Texas-race) Racial composition in Texas counties.

```{r map-Texas-race, fig.width = 7, fig.asp = 0.75, fig.cap = '(ref:map-Texas-race)'}
# we reuse the geometries from texas_income
select(texas_income, -NAME, -variable, -estimate, -moe) %>%
  left_join(select(texas_race, GEOID, variable, pct) %>% filter(variable == "White")) %>%
  ggplot(aes(fill = pct)) +
  geom_sf(color = "white") +
  coord_sf(datum = NA) + 
  theme_map() +
  #scale_fill_continuous_diverging(palette = "Blue-Red 3", # uncomment for alternative color scale
  scale_fill_continuous_carto(palette = "Earth",
                              mid = 50,
                              limits = c(0, 100),
                              breaks = 25*(0:4),
                              name = "% white") +
  theme(legend.title.align = 0.5,
        legend.text.align = 0,
        legend.position = c(0.07, 0.8))
```

(ref:diverging-scales) Example diverging color scales.

```{r diverging-scales, fig.width=8.5, fig.asp=3*.14, fig.cap = '(ref:diverging-scales)'}
p1 <- gg_color_swatches(7) + scale_fill_brewer(type = "div", palette = "BrBG") + ggtitle("Colorbrewer BrBG")
p2 <- gg_color_swatches(7) + scale_fill_discrete_diverging("Blue-Red") + ggtitle("Blue-Red")
p3 <- gg_color_swatches(7) + scale_fill_discrete_carto(palette = "ArmyRose") + ggtitle("Carto ArmyRose")
plot_grid(p1, p2, p3, ncol = 1)
```

## Dos and don'ts of color use 


(ref:popgrowth-vs-popsize-colored) Population growth from 2000 to 2010 versus population size in 2000, for all 50 US states and the Discrict of Columbia. Every state is marked in a different color, but it is not possible to reliably match the colors in the legend to the dots in the scatter plot.

```{r popgrowth-vs-popsize-colored, fig.width = 7, fig.asp = 2*0.618, fig.cap = '(ref:popgrowth-vs-popsize-colored)'}
colors <- c(rainbow_hcl(8, l = 35, c = 25, start = 0, end = 315),
            rainbow_hcl(8, l = 45, c = 34, start = -10, end = 305),
            rainbow_hcl(9, l = 55, c = 42, start = -20, end = 300),
            rainbow_hcl(9, l = 65, c = 50, start = -30, end = 290),
            rainbow_hcl(9, l = 75, c = 55, start = -40, end = 280),
            rainbow_hcl(8, l = 85, c = 32, start = -50, end = 265))

#colors <- sample(colors, 51)

p_base <- ggplot(popgrowth_df, aes(x = pop2000, y = 100*popgrowth, color = as.character(state))) +
  geom_point(size = 4) +
  scale_x_log10(labels = label_log10) +
  scale_color_manual(values = colors, name = "state") +
  xlab("population size in 2000") +
  ylab("population growth (%)\n2000 to 2010    ") +
  theme_minimal_grid() +
  theme(legend.text = element_text(size = 10),
        legend.justification = "center")

p_comb <- plot_grid(p_base + theme(legend.position = "none"), get_legend(p_base), ncol = 1)
stamp_bad(plot_grid(NULL, p_comb, NULL, rel_widths = c(.5, 6, .5), nrow = 1))
```

(ref:popgrowth-vs-popsize-bw) Population growth from 2000 to 2010 versus population size in 2000. Unlike Figure \@ref(fig:popgrowth-vs-popsize-colored), I have now shown all states in black and have directly labeled a subset of states. All states with very large or very small population size in 2000 or with very large population growth are labeled. Among the remaining states, an arbitrary subset is labeled to provide context.

```{r popgrowth-vs-popsize-bw, fig.width = 8.5, fig.asp = 0.8, fig.cap = '(ref:popgrowth-vs-popsize-bw)'}
library(ggrepel)

set.seed(7586)

df_repel <- select(popgrowth_df, x = pop2000, y = popgrowth, label = state) %>%
  mutate(y = 100 * y,
         label = ifelse(runif(n()) < .35 | x > 1e7 | x < 640000 | y > 20, as.character(label), ""))

ggplot(popgrowth_df, aes(x = pop2000, y = 100*popgrowth)) +
  geom_point(size = 2) +
  geom_text_repel(data = df_repel,
                  aes(x, y, label = label),
                  segment.alpha = 0.5,
                  size = 10/.pt) +
  scale_x_log10(labels = label_log10) +
  xlab("population size in 2000") +
  ylab("population growth (%)\n2000 to 2010    ") +
  theme_minimal_grid() +
  theme(legend.text = element_text(size = 10),
        legend.justification = "center")
```

(ref:popgrowth-US-rainbow) Population growth in the US from 2000 to 2010. The rainbow coloring of states serves no purpose and is distracting. Furthermore, the colors are overly saturated.

```{r popgrowth-US-rainbow, fig.width = 6.5, fig.asp = 1.2, fig.cap = '(ref:popgrowth-US-rainbow)'}
popgrowth_bars_rainbow <- ggplot(popgrowth_df, aes(x = state, y = 100*popgrowth, fill = state)) + 
  geom_col() + 
  scale_y_continuous(limits = c(-.6, 37.5), expand = c(0, 0),
                     name = "population growth (%)\n2000 to 2010    ") +
  scale_fill_hue(c = 140, l = 55) +
  coord_flip() + 
  theme_minimal_vgrid() +
  theme(axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.length = unit(0, "pt"),
        axis.text.y = element_text(size = 10),
        legend.position = "none")

stamp_ugly(popgrowth_bars_rainbow)
```

This figure violates two principles: (i) don't use spurious colors; (ii) don't use overly saturated colors.
