```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")

library(dplyr)
```

# Effective use of color in figures

(ref:popgrowth-US) Population growth in the US from 2000 to 2010. States in the West and South have seen the largest increases, whereas states in the Midwest and Northeast have seen much smaller increases or even, in the case of Michigan, a decrease.

```{r popgrowth-US, fig.width = 6.5, fig.asp = 1.2, fig.cap = '(ref:popgrowth-US)'}
popgrowth_df <- left_join(US_census, US_regions) %>%
    group_by(region, division, state) %>%
    summarize(pop2000 = sum(pop2000, na.rm = TRUE),
              pop2010 = sum(pop2010, na.rm = TRUE),
              popgrowth = (pop2010-pop2000)/pop2000,
              area = sum(area)) %>%
    arrange(popgrowth) %>%
    ungroup() %>%
    mutate(state = factor(state, levels = state),
           region = factor(region, levels = c("West", "South", "Midwest", "Northeast")))

# make color vector in order of the state
region_colors <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442")
region_colors_dark <- darken(region_colors, 0.4)
state_colors <- region_colors_dark[as.numeric(popgrowth_df$region[order(popgrowth_df$state)])]

ggplot(popgrowth_df, aes(x = state, y = 100*popgrowth, fill = region)) + 
  geom_col() + 
  scale_y_continuous(limits = c(-.6, 37.5), expand = c(0, 0),
                     name = "population growth (%)\n2000 to 2010    ") +
  scale_fill_manual(values = region_colors) +
  coord_flip() + 
  theme_minimal_vgrid() +
  theme(axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.length = unit(0, "pt"),
        axis.text.y = element_text(size = 10, color = state_colors),
        legend.position = c(.56, .68),
        #legend.text = element_text(color = region_colors),
        legend.background = element_rect(fill = "#ffffffb0"))
```


(ref:Aus-athletes-track) Track athletes are among the shortest and leanest of male professional athletes participating in popular sports.

```{r Aus-athletes-track, fig.cap='(ref:Aus-athletes-track)'}
male_Aus <- filter(Aus_athletes, sex=="m") %>%
  filter(sport %in% c("basketball", "field", "swimming", "track (400m)",
                      "track (sprint)", "water polo")) %>%
  mutate(sport = case_when(sport == "track (400m)" ~ "track",
                           sport == "track (sprint)" ~ "track",
                           TRUE ~ sport))

male_Aus$sport <- factor(male_Aus$sport,
                         levels = c("track", "field", "water polo", "basketball", "swimming"))

colors <- c("#BD3828", rep("#808080", 4))
fills <- c("#BD3828D0", rep("#80808080", 4))

ggplot(male_Aus, aes(x=height, y=pcBfat, shape=sport, color = sport, fill = sport)) +
  geom_point(size = 3) +
  scale_shape_manual(values = 21:25, name = "height (cm)" ) +
  scale_color_manual(values = colors, name = "height (cm)") +
  scale_fill_manual(values = fills, name = "height (cm)") +
  ylab("% bodyfat") +
  theme_minimal()
```


## Color scales

```{r fig.width=8.5, fig.asp=3*.14}
p1 <- gg_color_swatches(7) + scale_fill_OkabeIto() + ggtitle("Okabe Ito")
p2 <- gg_color_swatches(7) + scale_fill_brewer(type = "qual", palette = "Accent") + ggtitle("Colorbrewer Accent")
p3 <- gg_color_swatches(7) + scale_fill_hue() + ggtitle("ggplot2 default")
plot_grid(p1, p2, p3, ncol = 1)
```


```{r fig.width=8.5, fig.asp=3*.14}
p1 <- gg_color_swatches(7) + scale_fill_brewer(type = "seq", palette = "Blues", direction = -1) + ggtitle("Colorbrewer Blues")
p2 <- gg_color_swatches(7) + scale_fill_discrete_sequential("Heat") + ggtitle("Heat")
p3 <- gg_color_swatches(7) + scale_fill_viridis_d() + ggtitle("Viridis")
plot_grid(p1, p2, p3, ncol = 1)
```

```{r fig.width=8.5, fig.asp=3*.14}
p1 <- gg_color_swatches(7) + scale_fill_brewer(type = "div", "Spectral") + ggtitle("Colorbrewer Spectral")
p2 <- gg_color_swatches(7) + scale_fill_discrete_diverging("Heat") + ggtitle("heat")
p3 <- gg_color_swatches(7) + scale_fill_discrete_carto(palette = "Fall") + ggtitle("Carto Fall")
plot_grid(p1, p2, p3, ncol = 1)
```

## Dos and Don'ts of color use 


(ref:popgrowth-vs-popsize-colored) Population growth from 2000 to 2010 versus population size in 2000, for all 50 US states and the Discrict of Columbia. Every state is marked in a different color, but it is not possible to reliably match the colors in the legend to the dots in the scatter plot.

```{r popgrowth-vs-popsize-colored, fig.width = 7, fig.asp = 2*0.618, fig.cap = '(ref:popgrowth-vs-popsize-colored)'}
colors <- c(rainbow_hcl(8, l = 35, c = 25, start = 0, end = 315),
            rainbow_hcl(8, l = 45, c = 34, start = -10, end = 305),
            rainbow_hcl(9, l = 55, c = 42, start = -20, end = 300),
            rainbow_hcl(9, l = 65, c = 50, start = -30, end = 290),
            rainbow_hcl(9, l = 75, c = 55, start = -40, end = 280),
            rainbow_hcl(8, l = 85, c = 32, start = -50, end = 265))

#colors <- sample(colors, 51)

p_base <- ggplot(popgrowth_df, aes(x = pop2000, y = 100*popgrowth, color = as.character(state))) +
  geom_point(size = 4) +
  scale_x_log10(labels = label_log10) +
  scale_color_manual(values = colors, name = "state") +
  xlab("population size in 2000") +
  ylab("population growth (%)\n2000 to 2010    ") +
  theme_minimal_grid() +
  theme(legend.text = element_text(size = 10),
        legend.justification = "center")

p_comb <- plot_grid(p_base + theme(legend.position = "none"), get_legend(p_base), ncol = 1)
stamp_bad(plot_grid(NULL, p_comb, NULL, rel_widths = c(.5, 6, .5), nrow = 1))
```

```{r fig.width = 6.5, fig.asp = 1.2}
popgrowth_bars_rainbow <- ggplot(popgrowth_df, aes(x = state, y = 100*popgrowth, fill = state)) + 
  geom_col() + 
  scale_y_continuous(limits = c(-.6, 37.5), expand = c(0, 0),
                     name = "population growth (%)\n2000 to 2010    ") +
  scale_fill_hue(c = 140, l = 55) +
  coord_flip() + 
  theme_minimal_vgrid() +
  theme(axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.length = unit(0, "pt"),
        axis.text.y = element_text(size = 10),
        legend.position = "none")

stamp_ugly(popgrowth_bars_rainbow)
```
This figure violates two principles: (i) don't use spurious colors; (ii) don't use overly saturated colors.


Notes:

- Color as quantification: colored map
