```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")

library(dplyr)
library(ggforce)
library(ggridges)
library(colorspace)
```

# (PART\*) Part II: A visualization for every occasion {-}

# Directory of visualizations

```{r message = FALSE, warning = FALSE}
## general setup code

# theme
theme_plot_icon <- function(bg_color = "#F5F8EA", line_color = "#243400") {
  line_size <- .8
  font_size <- 10
  theme_dviz_half_open() %+replace% theme(
    axis.text.x       = element_blank(),
    axis.text.y       = element_blank(),
    axis.title.x      = element_blank(),
    axis.title.y      = element_blank(),
    axis.line.x       = element_line(size = line_size, color = line_color),
    axis.line.y       = element_line(size = line_size, color = line_color),
    axis.ticks        = element_line(size = line_size, color = line_color),
    axis.ticks.length = grid::unit(4, "pt"),
    legend.position   = "none",
    plot.margin       = margin(2, font_size, font_size, font_size),
    plot.title        = element_text(hjust = 0.5, face = "bold", color = line_color,
                                     size = font_size,
                                     margin = margin(0, 0, 2, 0)),
    plot.background   = element_rect(fill = bg_color)
  )
}


# data sets
set.seed(5142)

n <- 15
x <- rnorm(n)
y <- .4*x + .6*rnorm(n)
df_scatter_xy <- data.frame(x, y)

df_one_dist <- data.frame(x = c(rnorm(1000, 1., 1.6), rnorm(350, 4, .4)))

df_one_normal <- data.frame(x = rnorm(20))

df_fractions <- data.frame(y = c(.3, .39, .48, .6, .25, .13, .22, .24, .45, .48, .3, .16),
                 x = factor(rep(1:4, 3)),
                 type = rep(c("A", "B", "C"), each = 4))

n <- 70
df_multi_dist <- data.frame(y = c(rnorm(n, 1, .8), rnorm(n, 2, .7), rnorm(n, 0, .5)),
                 type = rep(c("A", "B", "C"), each = n),
                 number = rep(c(2, 1, 3), each = n))

df_props = data.frame(value = c(60, 30, 10),
                      group = c("A", "B", "C"))



# palettes

npal <- 5
# earth-brown
pal_earth_brown <- sequential_hcl(n = npal, h1 = 71, c1 = 80, c2 = 10, l1 = 18, l2 = 97, p1 = 1.5)

# brown-green
pal_brown_green <- sequential_hcl(n = npal, h1 = 86, c1 = 80, c2 = 10, l1 = 18, l2 = 97, p1 = 1.5)

# green-brown
pal_green_brown <- sequential_hcl(n = npal, h1 = -265, c1 = 80, c2 = 10, l1 = 18, l2 = 97, p1 = 1.5)

# burgundy-red
pal_burgundy_red <- sequential_hcl(n = npal, h1 = 17, c1 = 80, c2 = 10, l1 = 18, l2 = 97, p1 = 1.5)

# brown-red
pal_brown_red <- sequential_hcl(n = npal, h1 = 41, c1 = 80, c2 = 10, l1 = 18, l2 = 97, p1 = 1.5)

# ocean-blue
pal_ocean_blue <- sequential_hcl(n = npal, h1 = 241, c1 = 80, c2 = 10, l1 = 18, l2 = 97, p1 = 1.5)

# steel-blue
pal_steel_blue <- sequential_hcl(n = npal, h1 = 257, c1 = 80, c2 = 10, l1 = 18, l2 = 97, p1 = 1.5)
```

## Proportions

```{r proportions, fig.width = 8, fig.asp = 1/2, message = FALSE}
palette <- pal_brown_green

p1_main <- ggplot(df_props, aes(x = 1, y = value, fill = group)) + 
  geom_col(position = "stack", color = palette[1]) + 
  coord_polar(theta = "y") +
  scale_y_continuous(breaks = NULL, name = "") +
  scale_x_continuous(breaks = NULL, name = "") +
  scale_fill_manual(values = palette[2:4]) +
  labs(title = "Pie chart") +
  theme_plot_icon(palette[npal], palette[1]) +
  theme(axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank())

# make sure plot background is fully filled, as in the other plots
p1 <- ggdraw(p1_main) +
  theme(plot.background = element_rect(fill = palette[npal],
                                       color = "transparent"))

p2 <- ggplot(df_props, aes(x = group, y = value, fill = group)) + 
  geom_col(position="identity", color = palette[1],
           width = .8) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = palette[2:4]) +
  labs(title = "Side-by-side bars") +
  theme_plot_icon(palette[npal], palette[1])

p3 <- ggplot(filter(df_fractions, x!=4), aes(x, y,
                                   fill=factor(type, levels = c("A", "C", "B")))) + 
  geom_col(position="dodge", color = palette[1],
           width = .7) +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, .58)) +
  scale_fill_manual(values = palette[2:4]) +
  labs(title = "Side-by-side bars") +
  theme_plot_icon(palette[npal], palette[1])

p4 <- ggplot(df_fractions, aes(x, y, fill=type)) + 
  geom_col(position="stack", color = palette[1]) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = palette[2:4]) +
  labs(title = "Stacked bars") +
  theme_plot_icon(palette[npal], palette[1])

p5 <- ggplot(df_multi_dist, aes(x = y, fill = factor(type, levels = c("C", "A", "B")))) + 
  geom_density(alpha = 0.7, color = palette[1], position = "fill") +
  scale_fill_manual(values = palette[2:4]) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(title = "Stacked densities") +
  theme_plot_icon(palette[npal], palette[1])


plot_grid(p1, p2, p3, p4,
          p5, ncol = 4, scale = .9)
```



## Individual distributions

```{r single-distributions, fig.width = 8, fig.asp = 1/4, message = FALSE, warning=FALSE}

palette <- pal_earth_brown 

p1 <- ggplot(df_one_dist, aes(x)) +
  geom_histogram(fill = palette[4], color = palette[2], bins = 18) +
  scale_x_continuous(limits = c(-4.8, 6.8), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, 290), 
                     expand = c(0, 0)) +
  labs(title = "Histogram") +
  theme_plot_icon(palette[npal], palette[1])


p2 <- ggplot(df_one_dist, aes(x)) +
  geom_density(fill = palette[4], color = palette[2], bw = .35) +
  scale_x_continuous(limits = c(-4.8, 6.8), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, .29), expand = c(0, 0)) +
  labs(title = "Density plot") +
  theme_plot_icon(palette[npal], palette[1])

p3 <- ggplot(df_one_normal, aes(x)) +
  stat_ecdf(color = palette[2], size = .7) +
  scale_x_continuous(expand = c(0.05, 0)) +
  scale_y_continuous(limits = c(0, 1.1), expand = c(0, 0)) +
  labs(title = "Cumulative density") +
  theme_plot_icon(palette[npal], palette[1])

p4 <- ggplot(df_one_normal, aes(sample = x)) +
  geom_abline(intercept = 0, slope = 1, color = palette[4]) +
  geom_qq(color = palette[2]) +
  labs(title = "q-q plot") +
  theme_plot_icon(palette[npal], palette[1])

plot_grid(p1, p2, p3, p4, ncol = 4, scale = .9)
```

Histograms and density plots (Chapter \@ref(histograms-density-plots)) provide the most intuitive visualizations of a distribution, but both require arbitrary parameter choices and can be misleading. Cumulative densities and q-q plots (Chapter \@ref(ecdf-qq)) always represent the data faithfully but can be more difficult to interpret.

## Multiple distributions

```{r multiple-distributions, fig.width = 8, fig.asp = 1/2, message = FALSE}
palette <- pal_ocean_blue

p1 <- ggplot(df_multi_dist, aes(x = type, y = y)) + geom_boxplot(color = palette[1], fill = palette[4]) +
  labs(title = "Boxplots") +
  theme_plot_icon(palette[npal], palette[1])


p2 <- ggplot(df_multi_dist, aes(x = type, y = y)) + geom_violin(color = palette[1], fill = palette[4]) +
  labs(title = "Violin plots") +
  theme_plot_icon(palette[npal], palette[1])

p3 <- ggplot(df_multi_dist, aes(x = type, y = y)) + geom_jitter(color = palette[1], width = 0.3, size = .8) +
  labs(title = "Jittered points") +
  theme_plot_icon(palette[npal], palette[1])

p4 <- ggplot(df_multi_dist, aes(x = type, y = y)) + 
  geom_sina(color = palette[1], size = 0.8) +
  labs(title = "Sina plot") +
  theme_plot_icon(palette[npal], palette[1])

p5 <- ggplot(df_multi_dist, aes(x = y, fill = factor(type, levels = c("C", "A", "B")))) + 
  geom_histogram(color = palette[1], bins = 12) +
  scale_fill_manual(values = palette[2:4]) +
  labs(title = "Stacked histograms") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, 49), expand = c(0, 0)) +
  theme_plot_icon(palette[npal], palette[1])

p6 <- ggplot(df_multi_dist, aes(x = y, fill = factor(type, levels = c("C", "A", "B")))) + 
  geom_density(alpha = 0.7, color = palette[1]) +
  scale_fill_manual(values = palette[2:4]) +
  labs(title = "Overlapping densities") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, .95), expand = c(0, 0)) +
  theme_plot_icon(palette[npal], palette[1])

p7 <- ggplot(df_multi_dist, aes(x = y, y = number, group = number)) + 
  geom_density_ridges(alpha = 0.7, color = palette[1], fill = palette[3], scale = 2.0) +
  labs(title = "Ridgeline plot") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(limits = c(1, 6.5), expand = c(0, 0)) +
  theme_plot_icon(palette[npal], palette[1])


plot_grid(p1, p2, p3, p4, 
          p5, p6, p7, ncol = 4, scale = .9)
```


## Other

```{r fig.width = 8, fig.asp = 1/4, message = FALSE, warning=FALSE}

palette <- pal_steel_blue


p1 <- ggplot(df_scatter_xy, aes(x, y)) + 
  geom_point(color = palette[3], size = 2.4) + 
  scale_x_continuous(expand = c(.2, 0)) +
  scale_y_continuous(expand = c(.2, 0)) +
  labs(title = "Scatterplot") +
  theme_plot_icon(palette[npal], palette[1])

plot_grid(p1, ncol = 4, scale = .9)

## plots to add:

# x-y relationships:
# - 2d bins
# - hexbins
# - smooth line with error region
# - multiple lines

# quantities
# - horizontal bars, ordered

# proportions
# - mosaic plot/stacked bars flipped to horizontal
# - pie chart
```

