```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")
library(ggrepel)
```

# Figure titles and captions {#figure-titles-captions}

Topics to cover in this chapter:

- Figure and table captions.
- Locations of figure and table captions. (Figure captions below, table captions above.)
- Axis and legend titles (when do we need them, when can we leave them out?)


(ref:corruption-development) Corruption and human development: The most developed countries experience the least corruption. Data sources: Transparency International & UN Human Development Report

```{r corruption-development, fig.width = 7, fig.asp = 0.7, fig.cap = '(ref:corruption-development)'}
country_highlight <- c("Germany", "Norway", "United States", "Greece", "Singapore", "Rwanda", "Russia", "Venezuela", "Sudan", "Iraq", "Ghana", "Niger", "Chad", "Kuwait", "Qatar", "Myanmar", "Nepal", "Chile", "Brazil", "Japan", "China")


corruption %>% filter(year == 2015) %>% na.omit() %>%
  mutate(region = case_when(
    region == "Middle East and North Africa" ~ "Middle East\nand North Africa",
    region == "Europe and Central Asia" ~ "Europe and\nCentral Asia",
    region == "Sub Saharan Africa" ~ "Sub Saharan\nAfrica",
    TRUE ~ region),
    label = ifelse(country %in% country_highlight, country, "")
    ) %>%
  ggplot(aes(cpi, hdi)) + 
    geom_smooth(aes(color = "y ~ log(x)", fill = "y ~ log(x)"),
                method = 'lm', formula = y~log(x), se = FALSE, fullrange = TRUE) +
    geom_point(aes(color = region, fill = region),
               size = 2.5, alpha = 0.5) + 
    geom_text_repel(aes(label = label), color = "black", size = 10/.pt,
                    point.padding = 0.1, box.padding = .6, force = 1.,
                    min.segment.length = 0,
                    family = dviz_font_family) +
    scale_color_OkabeIto(name = NULL, order = c(1:5, 8)) +
    scale_fill_OkabeIto(name = NULL, order = c(1:5, 8)) +
    scale_y_continuous(limits = c(0.3, 1.05), breaks = c(0.2, 0.4, 0.6, 0.8, 1.0),
                       expand = c(0, 0),
                       name = "Human Development Index\n(1.0 = most developed)") +
    scale_x_continuous(limits = c(10, 95),
                       breaks = c(20, 40, 60, 80, 100),
                       expand = c(0, 0),
                       name = "Corruption Perceptions Index (100 = least corrupt)") +
    guides(color = guide_legend(nrow = 1,
      override.aes = list(linetype = c(rep("blank", 5), "solid"),
                          shape = c(rep(21, 5), NA)))) +
    theme_dviz_hgrid(12, rel_small = 1) +
    theme(legend.position = "top",
          legend.justification = "right",
          legend.text = element_text(size = 10)) -> plot_corrupt_base

#cur_dev <- grDevices::dev.cur()
#cowplot::png_null_device(width = 7, height = 4.9)
#null_dev <- dev.cur()
#grob_corrupt_base <- grid::forceGrob(ggplotGrob(plot_corrupt_base))
#null <- grDevices::dev.off(null_dev)
#null <- grDevices::dev.set(cur_dev)
ggsave("figures/corruption_plot_base.png", plot_corrupt_base, width = 7, height = 4.9,
       dpi = 600)

ggdraw() + draw_image("figures/corruption_plot_base.png")
```


(ref:corruption-development-infographic) Figure \@ref(fig:corruption-development) reformatted to be posted on the web or to be used as an infographic. The title, subtitle, and data source statements have been incorporated into the figure.

```{r corruption-development-infographic, fig.width = 7, fig.asp = 4.9*(0.12+1+.07)/7, fig.cap = '(ref:corruption-development-infographic)'}
plot_corrupt_title <- ggdraw() +
  labs(title = "Corruption and human development",
       subtitle = "The most developed countries experience the least corruption") +
  theme_dviz_map(12, rel_small = 1) +
  theme(plot.margin = margin(6, 0, 0, 0))

plot_corrupt_caption <- ggplot() +
  labs(caption = "Data sources: Transparency International & UN Human Development Report") +
  theme_dviz_map(12) +
  theme(plot.margin = margin(0, 0, 6, 0))


plot_grid(plot_corrupt_title,
          ggdraw() + draw_image("figures/corruption_plot_base.png"),
          plot_corrupt_caption,
          ncol = 1, rel_heights = c(.12, 1, .07))
```
