```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")
library(ggrepel)
library(grid)
library(gridExtra)
library(gtable)
library(tibble)
library(lubridate)
```

# Putting data into context: Titles, captions, labels, and tables {#figure-titles-captions}

*Need a brief intro paragraph here.*


## Figure titles and captions

Consider Figure \@ref(fig:corruption-development). This figure is inspired by a posting in @Economist-corruption. I have recreated their figure with several subtle but important modifications. First, I have updated the data to a more recent dataset (2015) and adapted the color and symbol styling so they follow the styling used throughout this book. Second, and more importantly, I have drawn the actual figure without title or statement about the data sources used. These pieces of information are instead provided in the caption block underneath the figure display. This is the style I am using throughout this book. I consistently show figures without integrated titles and with separate captions. (One exception are the stylized plot examples in Chapter \@ref(directory-of-visualizations), which instead have titles and no captions.)

(ref:corruption-development) Corruption and human development: The most developed countries experience the least corruption. Data sources: Transparency International & UN Human Development Report

```{r corruption-development, fig.width = 7, fig.asp = 0.7, fig.cap = '(ref:corruption-development)'}
country_highlight <- c("Germany", "Norway", "United States", "Greece", "Singapore", "Rwanda", "Russia", "Venezuela", "Sudan", "Iraq", "Ghana", "Niger", "Chad", "Kuwait", "Qatar", "Myanmar", "Nepal", "Chile", "Argentina", "Japan", "China")


corruption %>% filter(year == 2015) %>% na.omit() %>%
  mutate(region = case_when(
    region == "Middle East and North Africa" ~ "Middle East\nand North Africa",
    region == "Europe and Central Asia" ~ "Europe and\nCentral Asia",
    region == "Sub Saharan Africa" ~ "Sub Saharan\nAfrica",
    TRUE ~ region),
    label = ifelse(country %in% country_highlight, country, "")
    ) %>%
  ggplot(aes(cpi, hdi)) + 
    geom_smooth(aes(color = "y ~ log(x)", fill = "y ~ log(x)"),
                method = 'lm', formula = y~log(x), se = FALSE, fullrange = TRUE) +
    geom_point(aes(color = region, fill = region),
               size = 2.5, alpha = 0.5) + 
    geom_text_repel(aes(label = label), color = "black", size = 10/.pt,
                    point.padding = 0.1, box.padding = .6, force = 1.,
                    min.segment.length = 0, seed = 7654,
                    family = dviz_font_family) +
    scale_color_OkabeIto(name = NULL, order = c(1:5, 8)) +
    scale_fill_OkabeIto(name = NULL, order = c(1:5, 8)) +
    scale_y_continuous(limits = c(0.3, 1.05), breaks = c(0.2, 0.4, 0.6, 0.8, 1.0),
                       expand = c(0, 0),
                       name = "Human Development Index, 2015\n(1.0 = most developed)") +
    scale_x_continuous(limits = c(10, 95),
                       breaks = c(20, 40, 60, 80, 100),
                       expand = c(0, 0),
                       name = "Corruption Perceptions Index, 2015 (100 = least corrupt)") +
    guides(color = guide_legend(nrow = 1,
      override.aes = list(linetype = c(rep("blank", 5), "solid"),
                          shape = c(rep(21, 5), NA)))) +
    theme_dviz_hgrid(12, rel_small = 1) +
    theme(legend.position = "top",
          legend.justification = "right",
          legend.text = element_text(size = 10)) -> plot_corrupt_base

## for some reason grid::forceGrob creates an empty plot, not sure why
#cur_dev <- grDevices::dev.cur()
#cowplot::png_null_device(width = 7, height = 4.9)
#null_dev <- dev.cur()
#grob_corrupt_base <- grid::forceGrob(ggplotGrob(plot_corrupt_base))
#null <- grDevices::dev.off(null_dev)
#if (cur_dev > 1 ) null <- grDevices::dev.set(cur_dev)
#ggdraw(grob_corrupt_base)


ggsave("figures/corruption_plot_base.png", plot_corrupt_base, width = 7, height = 4.9,
       dpi = 600)

ggdraw() + draw_image("figures/corruption_plot_base.png")
```

For comparison, here I also provide a version of the figure that has these elements incorporated into the main display (Figure \@ref(fig:corruption-development-infographic)). In a direct comparison, you may find Figure \@ref(fig:corruption-development-infographic) more attractive than Figure \@ref(fig:corruption-development), and you may wonder why I am choosing the latter style throughout this book. I do so because the two styles have different application areas, and figures with integrated titles are not appropriate for conventional book layouts. The underlying principle is that a figure can have only one title. Either the title is integrated into the actual figure display or it is provided as the first element of the caption underneath the figure. And, if a publication is laid out such that each figure has a regular caption block underneath the display item, then the title *must* be provided in that block of text. For this reason, in the context of conventional book or article publishing, we do not normally integrate titles into figures. Figures with integrated titles, subtitles, and data source statements are appropriate, however, if they are meant to be used as stand-alone infographics or to be posted on social media or on a web page without accompanying caption text.

(ref:corruption-development-infographic) Figure \@ref(fig:corruption-development) reformatted to be posted on the web or to be used as an infographic. The title, subtitle, and data source statements have been incorporated into the figure.

```{r corruption-development-infographic, fig.width = 7, fig.asp = 4.9*(0.12+1+.07)/7, fig.cap = '(ref:corruption-development-infographic)'}
plot_corrupt_title <- ggdraw() +
  labs(title = "Corruption and human development",
       subtitle = "The most developed countries experience the least corruption") +
  theme_dviz_map(12, rel_small = 1) +
  theme(plot.margin = margin(6, 0, 0, 0))

plot_corrupt_caption <- ggplot() +
  labs(caption = "Data sources: Transparency International & UN Human Development Report") +
  theme_dviz_map(12) +
  theme(plot.margin = margin(0, 0, 6, 0))


plot_grid(plot_corrupt_title,
          ggdraw() + draw_image("figures/corruption_plot_base.png"),
          plot_corrupt_caption,
          ncol = 1, rel_heights = c(.12, 1, .07))
```

```{block type='rmdtip', echo=TRUE}
If your document layout uses caption blocks underneath each figure, then place the figure titles as the first element of each caption block, not on top of the figures.
```


## Axis and legend titles

*Discuss when we need axis and legend titles and when we can leave them out. Maybe a plot showing a timeline, with an absurd x axis title listing date with units.*

Refer to Figure \@ref(ref:tech-stocks-good-legend).

(ref:tech-stocks-minimal-labeling) Stock price over time for four major tech companies. The stock price for each company has been normalized to equal 100 in June 2012. *Refer to Figure \@ref(ref:tech-stocks-good-legend).*

```{r tech-stocks-minimal-labeling, fig.cap = '(ref:tech-stocks-minimal-labeling)'}
price_plot_base <- ggplot(tech_stocks, aes(x = date, y = price_indexed, color = ticker)) +
  geom_line(na.rm = TRUE) +
  scale_color_manual(
    values = c("#000000", "#E69F00", "#56B4E9", "#009E73"),
    name = "",
    breaks = c("FB", "GOOG", "MSFT", "AAPL"),
    labels = c("Facebook", "Alphabet", "Microsoft", "Apple")
  ) +
  scale_x_date(
    limits = c(ymd("2012-06-01"), ymd("2017-05-31")),
    expand = c(0,0)
  ) + 
  scale_y_continuous(
    limits = c(0, 560),
    expand = c(0,0)
  ) +
  theme_dviz_hgrid() + 
  theme(plot.margin = margin(3, 7, 3, 0))

price_plot_base + xlab(NULL) + ylab("stock price, indexed")
```


```{r tech-stocks-minimal-labeling-bad, fig.cap = '(ref:tech-stocks-minimal-labeling-bad)'}
stamp_bad(
  price_plot_base + xlab(NULL) + ylab(NULL)
)
```

```{r tech-stocks-labeling-ugly, fig.cap = '(ref:tech-stocks-labeling-ugly)'}
stamp_ugly(
  price_plot_base + xlab("time (years AD)") + ylab("stock price, indexed\n(100/share price on Jun 1, 2012)") +
    scale_color_manual(
      values = c("#000000", "#E69F00", "#56B4E9", "#009E73"),
      name = "company",
      breaks = c("FB", "GOOG", "MSFT", "AAPL"),
      labels = c("Facebook", "Alphabet", "Microsoft", "Apple")
    ) +
    theme(legend.title.align = 0.5)

)
```

## Tables

Tables are an important tool for visualizing data. Yet because of their apparent simplicity, they may not always receive the attention they need. I have shown a handful of tables throughout this book, for example Tables \@ref(tab:boxoffice-gross), \@ref(tab:titanic-ages), and \@ref(tab:color-codes). Take a moment and locate these tables, look how they are formatted, and compare them to a table you or a colleague has recently made. In all likelihood, there are important differences. In my experience, absent proper training in table formatting few people will instinctively make the right formatting choices, and poorly formatted tables are even more prevalent than poorly designed figures in self-published documents. In addition, most software commonly used to create tables provides defaults that are not recommended. For example, my version of Microsoft Word provides 105 pre-defined table styles, and of these at least 70--80 violate some of the table rules I'm going to discuss here. So if you pick a Microsoft Word table layout at random, you have an 80% chance of picking one that has issues. And if you pick the default, you will end up with poorly formatted tables every time.

Some key rules for table layout are the following:

1. Do not use vertical lines.
2. Do not use horizontal lines between data rows. (Horizontal lines as separator between the title row and the first data row or as frame for the entire table are fine.) 
3. Text columns should be left aligned.
4. Number columns should be right aligned and should use the same number of decimal digits throughout.
5. Columns containing single characters are centered.
6. The header fields are aligned with their data, i.e., the heading for a text column will be left aligned and the heading for a number column will be right aligned. 

Figure \@ref(fig:table-examples) reproduces Table \@ref(tab:boxoffice-gross) from Chapter \@ref(visualizing-amounts) in four diferent ways, two of which (a, b) violate several of these rules and two of which (c, d) do not.

(ref:table-examples) Examples of poorly and appropriately formatted tables, using the data from Table \@ref(tab:boxoffice-gross) in Chapter \@ref(visualizing-amounts). (a) This table violates numerous conventions of proper table formatting, including using vertical lines, using horizontal lines between data rows, and using centered data columns. (b) This table suffers from all problems of Table (a), and in addition it creates additional visual noise by alternating between very dark and very light rows. Also, the table header is not strongly visually separated from the table body. (c) This is an appropriately formatted table with a minimal design. (d) Colors can be used effectively to group data into rows, but the color differences should be subtle. The table header can be set off by using a stronger color. Data source: Box Office Mojo (http://www.boxofficemojo.com/). Used with permission

```{r table-examples, fig.asp = 0.65, fig.cap = '(ref:table-examples)'}
boxoffice <- tibble(
  Rank = 1:5,
  Title = c("Star Wars", "Jumanji", "Pitch Perfect 3", "Greatest Showman", "Ferdinand"),
  Amount = c("$71,565,498", "$36,169,328", "$19,928,525", "$8,805,843", "$7,316,746")
)

table_base_size = 11
zgrob <- function(...) ggplot2::zeroGrob()

tt1 <- ttheme_minimal(
  base_size = table_base_size,
  base_family = dviz_font_family,
  core = list(
    bg_params = list(
      col = "black",
      lwd = 1
    )
  ),
  colhead = list(
    fg_params = list(
      fontface = 1L,
      fontfamily = dviz_font_family_bold
    ),
    bg_params = list(
      col = "black",
      lwd = 1
    )
  ),
  rowhead = list(fg_fun = zgrob, bg_fun = zgrob)
)

tt2 <- ttheme_default(
  base_size = table_base_size,
  base_family = dviz_font_family,
  core = list(
    fg_params = list(
      col = c("white", "black")
    ),
    bg_params = list(
      col = "black",
      lwd = 0.5,
      fill = c("grey45", "grey85")
    )
  ),
  colhead = list(
    fg_params = list(
      fontface = 1L,
      fontfamily = dviz_font_family_bold
    ),
    bg_params = list(
      col = "black",
      lwd = 0.5,
      fill = "grey85"
    )
  ),
  rowhead = list(fg_fun = zgrob, bg_fun = zgrob)
)

tt3 <- ttheme_minimal(
  base_size = table_base_size,
  base_family = dviz_font_family,
  padding = unit(c(4, 3.2), "mm"),
  core = list(
    fg_params = list(
      hjust = rep(c(0.5, 0, 1), each = 5),
      x = rep(c(0.5, 0.1, 0.9), each = 5)
    ),
    bg_params = list(
      col = NA
    )
  ),
  colhead = list(
    fg_params = list(
      hjust = c(0.5, 0, 1),
      x = c(0.5, 0.1, 0.9),
      fontface = 1L,
      fontfamily = dviz_font_family_bold
    ),
    bg_params = list(
      col = NA
    )
  ),
  rowhead = list(fg_fun = zgrob, bg_fun = zgrob)
)

tt4 <- ttheme_default(
  base_size = table_base_size,
  base_family = dviz_font_family,
  core = list(
    fg_params = list(
      col = "black",
      hjust = rep(c(0.5, 0, 1), each = 5),
      x = rep(c(0.5, 0.1, 0.9), each = 5)
    ),
    bg_params = list(
      col = NA,
      fill = c('#D9E0EF', '#C2CCE3') #c("grey95", "grey85")
    )
  ),
  colhead = list(
    fg_params = list(
      col = "white",
      hjust = c(0.5, 0, 1),
      x = c(0.5, 0.1, 0.9),
      fontface = 1L,
      fontfamily = dviz_font_family_bold
    ),
    bg_params = list(
      col = NA,
      fill = "#4069A6"#"grey65"
    )
  ),
  rowhead = list(fg_fun = zgrob, bg_fun = zgrob)
)

# horizontal line to be used as separator
hline_top <- segmentsGrob(
  x0 = unit(0,"npc"),
  y0 = unit(1,"npc"),
  x1 = unit(1,"npc"),
  y1 = unit(1,"npc"),
  gp = gpar(lwd = 0.75, col = "black")
)
hline_bottom <- segmentsGrob(
  x0 = unit(0,"npc"),
  y0 = unit(0,"npc"),
  x1 = unit(1,"npc"),
  y1 = unit(0,"npc"),
  gp = gpar(lwd = 0.75, col = "black")
)

t1 <- tableGrob(boxoffice, rows = rep("", nrow(boxoffice)), theme = tt1)
t1$layout$clip <- "off"
t1 <- gtable_add_padding(t1, margin(14, 16, 0, -2))

t2 <- tableGrob(boxoffice, rows = rep("", nrow(boxoffice)), theme = tt2)
t2$layout$clip <- "off"
t2 <- gtable_add_padding(t2, margin(14, 16, 0, -2))

t3 <- tableGrob(boxoffice, rows = rep("", nrow(boxoffice)), theme = tt3)
t3 <- gtable_add_grob(t3,
  grobs = grobTree(hline_top, hline_bottom),
  t = 1, b = 1, l = 2, r = 4)
t3 <- gtable_add_grob(t3,
  grobs = hline_bottom,
  t = 6, b = 6, l = 2, r = 4)
t3$layout$clip <- "off"
t3 <- gtable_add_padding(t3, margin(14, 14, 0, -7))

t4 <- tableGrob(boxoffice, rows = rep("", nrow(boxoffice)), theme = tt4)
t4$layout$clip <- "off"
t4 <- gtable_add_padding(t4, margin(14, 16, 0, -2))

plot_grid(
  stamp_ugly(t1), NULL, stamp_ugly(t2),
  NULL, NULL, NULL,
  t3, NULL, t4,
  rel_widths = c(1, 0.06, 1),
  rel_heights = c(1, 0.08, 1),
  labels = c("a", "", "b", "", "", "", "c", "", "d")
)

```

*talk a bit about horizontal lines and colored rows*

Finally, there is a key distinction between figures and tables in where the caption is located relative to the display item. For figures, it is customary to place the caption underneath, whereas for tables it is customary to place it above. This caption placement is guided by the way in which readers process figures and tables. For figures, readers tend to first look at the graphical display and then read the caption for context, hence the caption makes sense below the figure. By contrast, tables tend to be processed like text, from top to bottom, and reading the table contents before reading the caption will frequently not be useful. Hence, captions are placed above the table.
