```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")
library(ggrepel)
library(grid)
library(gridExtra)
library(gtable)
library(tibble)
```

# Putting data into context: Titles, captions, labels, and tables {#figure-titles-captions}

Topics to cover in this chapter:

- Figure and table captions.
- Locations of figure and table captions. (Figure captions below, table captions above.)
- Axis and legend titles (when do we need them, when can we leave them out?)


(ref:corruption-development) Corruption and human development: The most developed countries experience the least corruption. Data sources: Transparency International & UN Human Development Report

```{r corruption-development, fig.width = 7, fig.asp = 0.7, fig.cap = '(ref:corruption-development)'}
country_highlight <- c("Germany", "Norway", "United States", "Greece", "Singapore", "Rwanda", "Russia", "Venezuela", "Sudan", "Iraq", "Ghana", "Niger", "Chad", "Kuwait", "Qatar", "Myanmar", "Nepal", "Chile", "Brazil", "Japan", "China")


corruption %>% filter(year == 2015) %>% na.omit() %>%
  mutate(region = case_when(
    region == "Middle East and North Africa" ~ "Middle East\nand North Africa",
    region == "Europe and Central Asia" ~ "Europe and\nCentral Asia",
    region == "Sub Saharan Africa" ~ "Sub Saharan\nAfrica",
    TRUE ~ region),
    label = ifelse(country %in% country_highlight, country, "")
    ) %>%
  ggplot(aes(cpi, hdi)) + 
    geom_smooth(aes(color = "y ~ log(x)", fill = "y ~ log(x)"),
                method = 'lm', formula = y~log(x), se = FALSE, fullrange = TRUE) +
    geom_point(aes(color = region, fill = region),
               size = 2.5, alpha = 0.5) + 
    geom_text_repel(aes(label = label), color = "black", size = 10/.pt,
                    point.padding = 0.1, box.padding = .6, force = 1.,
                    min.segment.length = 0,
                    family = dviz_font_family) +
    scale_color_OkabeIto(name = NULL, order = c(1:5, 8)) +
    scale_fill_OkabeIto(name = NULL, order = c(1:5, 8)) +
    scale_y_continuous(limits = c(0.3, 1.05), breaks = c(0.2, 0.4, 0.6, 0.8, 1.0),
                       expand = c(0, 0),
                       name = "Human Development Index\n(1.0 = most developed)") +
    scale_x_continuous(limits = c(10, 95),
                       breaks = c(20, 40, 60, 80, 100),
                       expand = c(0, 0),
                       name = "Corruption Perceptions Index (100 = least corrupt)") +
    guides(color = guide_legend(nrow = 1,
      override.aes = list(linetype = c(rep("blank", 5), "solid"),
                          shape = c(rep(21, 5), NA)))) +
    theme_dviz_hgrid(12, rel_small = 1) +
    theme(legend.position = "top",
          legend.justification = "right",
          legend.text = element_text(size = 10)) -> plot_corrupt_base

## for some reason grid::forceGrob creates an empty plot, not sure why
#cur_dev <- grDevices::dev.cur()
#cowplot::png_null_device(width = 7, height = 4.9)
#null_dev <- dev.cur()
#grob_corrupt_base <- grid::forceGrob(ggplotGrob(plot_corrupt_base))
#null <- grDevices::dev.off(null_dev)
#if (cur_dev > 1 ) null <- grDevices::dev.set(cur_dev)
#ggdraw(grob_corrupt_base)


ggsave("figures/corruption_plot_base.png", plot_corrupt_base, width = 7, height = 4.9,
       dpi = 600)

ggdraw() + draw_image("figures/corruption_plot_base.png")
```


(ref:corruption-development-infographic) Figure \@ref(fig:corruption-development) reformatted to be posted on the web or to be used as an infographic. The title, subtitle, and data source statements have been incorporated into the figure.

```{r corruption-development-infographic, fig.width = 7, fig.asp = 4.9*(0.12+1+.07)/7, fig.cap = '(ref:corruption-development-infographic)'}
plot_corrupt_title <- ggdraw() +
  labs(title = "Corruption and human development",
       subtitle = "The most developed countries experience the least corruption") +
  theme_dviz_map(12, rel_small = 1) +
  theme(plot.margin = margin(6, 0, 0, 0))

plot_corrupt_caption <- ggplot() +
  labs(caption = "Data sources: Transparency International & UN Human Development Report") +
  theme_dviz_map(12) +
  theme(plot.margin = margin(0, 0, 6, 0))


plot_grid(plot_corrupt_title,
          ggdraw() + draw_image("figures/corruption_plot_base.png"),
          plot_corrupt_caption,
          ncol = 1, rel_heights = c(.12, 1, .07))
```


(ref:table-examples) Examples of poorly and appropriately formatted tables, using the data from Table \@ref(tab:boxoffice-gross) in Chapter \@ref(visualizing-amounts). (a) Tables should not have any vertical lines and have no horizontal lines between data rows. In addition, text columns should be left aligned and number columns right aligned. (b) In addition to all issues present in Part (a), this version of the table creates additional visual noise by alternating between very dark and very light rows. Moreover, the table header is not strongly visually separated from the table body. (c) This is an appropriately formatted table with a minimal design. (d) Colors can be used effectively to group data into rows, but the color differences should be subtle. The table header can be set off by using a stronger color.

```{r table-examples, fig.asp = 0.65, fig.cap = '(ref:table-examples)'}
boxoffice <- tibble(
  Rank = 1:5,
  Title = c("Star Wars", "Jumanji", "Pitch Perfect 3", "Greatest Showman", "Ferdinand"),
  Amount = c("$71,565,498", "$36,169,328", "$19,928,525", "$8,805,843", "$7,316,746")
)

table_base_size = 11
zgrob <- function(...) ggplot2::zeroGrob()

tt1 <- ttheme_minimal(
  base_size = table_base_size,
  base_family = dviz_font_family,
  core = list(
    bg_params = list(
      col = "black",
      lwd = 1
    )
  ),
  colhead = list(
    fg_params = list(
      fontface = 1L,
      fontfamily = dviz_font_family_bold
    ),
    bg_params = list(
      col = "black",
      lwd = 1
    )
  ),
  rowhead = list(fg_fun = zgrob, bg_fun = zgrob)
)

tt2 <- ttheme_default(
  base_size = table_base_size,
  base_family = dviz_font_family,
  core = list(
    fg_params = list(
      col = c("white", "black")
    ),
    bg_params = list(
      col = "black",
      lwd = 0.5,
      fill = c("grey45", "grey85")
    )
  ),
  colhead = list(
    fg_params = list(
      fontface = 1L,
      fontfamily = dviz_font_family_bold
    ),
    bg_params = list(
      col = "black",
      lwd = 0.5,
      fill = "grey85"
    )
  ),
  rowhead = list(fg_fun = zgrob, bg_fun = zgrob)
)

tt3 <- ttheme_minimal(
  base_size = table_base_size,
  base_family = dviz_font_family,
  padding = unit(c(4, 3.2), "mm"),
  core = list(
    fg_params = list(
      hjust = rep(c(0.5, 0, 1), each = 5),
      x = rep(c(0.5, 0.1, 0.9), each = 5)
    ),
    bg_params = list(
      col = NA
    )
  ),
  colhead = list(
    fg_params = list(
      hjust = c(0.5, 0, 1),
      x = c(0.5, 0.1, 0.9),
      fontface = 1L,
      fontfamily = dviz_font_family_bold
    ),
    bg_params = list(
      col = NA
    )
  ),
  rowhead = list(fg_fun = zgrob, bg_fun = zgrob)
)

tt4 <- ttheme_default(
  base_size = table_base_size,
  base_family = dviz_font_family,
  core = list(
    fg_params = list(
      col = "black",
      hjust = rep(c(0.5, 0, 1), each = 5),
      x = rep(c(0.5, 0.1, 0.9), each = 5)
    ),
    bg_params = list(
      col = NA,
      fill = c('#D9E0EF', '#C2CCE3') #c("grey95", "grey85")
    )
  ),
  colhead = list(
    fg_params = list(
      col = "white",
      hjust = c(0.5, 0, 1),
      x = c(0.5, 0.1, 0.9),
      fontface = 1L,
      fontfamily = dviz_font_family_bold
    ),
    bg_params = list(
      col = NA,
      fill = "#4069A6"#"grey65"
    )
  ),
  rowhead = list(fg_fun = zgrob, bg_fun = zgrob)
)

# horizontal line to be used as separator
hline_top <- segmentsGrob(
  x0 = unit(0,"npc"),
  y0 = unit(1,"npc"),
  x1 = unit(1,"npc"),
  y1 = unit(1,"npc"),
  gp = gpar(lwd = 0.75, col = "black")
)
hline_bottom <- segmentsGrob(
  x0 = unit(0,"npc"),
  y0 = unit(0,"npc"),
  x1 = unit(1,"npc"),
  y1 = unit(0,"npc"),
  gp = gpar(lwd = 0.75, col = "black")
)

t1 <- tableGrob(boxoffice, rows = rep("", nrow(boxoffice)), theme = tt1)
t1$layout$clip <- "off"
t1 <- gtable_add_padding(t1, margin(14, 16, 0, -2))

t2 <- tableGrob(boxoffice, rows = rep("", nrow(boxoffice)), theme = tt2)
t2$layout$clip <- "off"
t2 <- gtable_add_padding(t2, margin(14, 16, 0, -2))

t3 <- tableGrob(boxoffice, rows = rep("", nrow(boxoffice)), theme = tt3)
t3 <- gtable_add_grob(t3,
  grobs = grobTree(hline_top, hline_bottom),
  t = 1, b = 1, l = 2, r = 4)
t3 <- gtable_add_grob(t3,
  grobs = hline_bottom,
  t = 6, b = 6, l = 2, r = 4)
t3$layout$clip <- "off"
t3 <- gtable_add_padding(t3, margin(14, 14, 0, -7))

t4 <- tableGrob(boxoffice, rows = rep("", nrow(boxoffice)), theme = tt4)
t4$layout$clip <- "off"
t4 <- gtable_add_padding(t4, margin(14, 16, 0, -2))

plot_grid(
  stamp_ugly(t1), NULL, stamp_ugly(t2),
  NULL, NULL, NULL,
  t3, NULL, t4,
  rel_widths = c(1, 0.06, 1),
  rel_heights = c(1, 0.08, 1),
  labels = c("a", "", "b", "", "", "", "c", "", "d")
)

```

Rules for table layout:

- no vertical lines
- no horizontal lines between data columns
- text is left aligned
- numbers are right aligned and use the same number of decimal digits
- single characters are centered
- headers are aligned with their data
