```{r echo = FALSE, message = FALSE, warning = FALSE}
# run setup script
source("_common.R")

library(forcats)
library(ggmap)
library(statebins)
library(sf)
```

# Visualizing geospatial data {#geospatial-data}

Topics to cover:

- Explain basic concepts, latitude, longitude, meridians, parallels, etc.
- Projections (give Alaska example)
- Layers: outline, streets, topography, etc.
- Chloropleth maps, cartograms, etc.

## Projections

(ref:world-orthographic) *caption goes here.*

```{r world-orthographic, fig.width = 4.5, fig.asp = 1, fig.cap = '(ref:world-orthographic)'}
cenlat <- 40
cenlong <- 15

ocean_col <- "#56B4E950"
land_col <- "#E69F00B0"
graticule_col <- "grey30"
line_col <- "black"

draw_ocean(cenlat, cenlong, col = ocean_col, line_col = graticule_col, lwd = 0.25)
draw_land(map_polys$world, cenlat, cenlong, col = land_col, line_col = line_col, lwd = 0.5)
```


(ref:world-longlat) Map of the world, shown in a simple latidue--longitude projection. This projection introduces severe distortions at the poles. For example, Greenland appears to be much bigger than Europe and comparable in size to Africa. In reality, Greenland is less than a quarter the size of Europe and less than a tenth the size of Africa (Figure \@ref(fig:world-orthographic)).

```{r world-longlat, fig.asp = 0.6, fig.cap = '(ref:world-longlat)'}
world_sf <- sf::st_as_sf(rworldmap::getMap(resolution = "low"))

crs_longlat <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"

ggplot(world_sf) + 
  geom_sf(fill = "#E69F00B0", color = "black", size = 0.5/.pt) + 
  coord_sf(expand = FALSE, crs = crs_longlat) + 
  scale_x_continuous(name = "longitude", breaks = seq(-120, 120, by = 60)) +
  scale_y_continuous(name = "latitude", breaks = seq(-60, 60, by = 30)) +
#  scale_x_continuous(
#    name = "longitude",
#    breaks = seq(-150, 150, by = 30),
#    labels = parse(text = c("NA", "120 * degree * W", "NA", "60 * degree * W", "NA", "0 * degree",
#                            "NA", "60 * degree * E", "NA", "120 * degree * E", "NA"))
#  ) +
#  scale_y_continuous(
#    name = "latitude",
#    breaks = seq(-75, 75, by = 15),
#    labels = parse(text = c("NA", "60 * degree * S", "NA", "30 * degree * S", "NA", "0 * degree",
#                            "NA", "30 * degree * N", "NA", "60 * degree * N", "NA"))
#  ) +
  theme_dviz_grid(font_size = 12, rel_small = 1) +
  theme(
    panel.background = element_rect(fill = "#56B4E950"),
    panel.grid.major = element_line(color = "gray30", size = 0.25),
    axis.ticks = element_line(color = "gray30", size = 0.5/.pt)
  )
```

(ref:world-robin) Map of the world, shown in the Robinson projection. This projection **a few sentences here.**

```{r world-robin, fig.asp = 0.6, fig.cap = '(ref:world-robin)'}
crs_robin <- "+proj=robin +lat_0=0 +lon_0=0 +x0=0 +y0=0"

xlim <- c(-18494733, 18613795)
ylim <- c(-9473396, 9188587)

lats <- c(90:-90, -90:90, 90)
longs <- c(rep(c(180, -180), each = 181), 180)

earth_boundary <- sf::st_sfc(
    sf::st_linestring(
      cbind(longs, lats)
    ),
    crs = "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
  )
earth_boundary <- sf::st_transform(earth_boundary, crs = crs_robin)
whiteout <- data.frame(
  x = c(xlim[1], xlim[2], xlim[2], xlim[1], xlim[1], earth_boundary[[1]][, 1]),
  y = c(ylim[1], ylim[1], ylim[2], ylim[2], ylim[1], earth_boundary[[1]][, 2])
)
earth_outline <- data.frame(
  x = earth_boundary[[1]][, 1],
  y = earth_boundary[[1]][, 2]
)

ggplot(world_sf) + 
  geom_sf(fill = "#E69F00B0", color = "black", size = 0.5/.pt) + 
  geom_polygon(
    data = whiteout, aes(x, y),
    fill = "white", color = NA
  ) +
  geom_polygon(
    data = earth_outline, aes(x, y),
    fill = NA, color = "gray30",
    size = 0.5/.pt
  ) +
  scale_x_continuous(name = NULL, breaks = seq(-120, 120, by = 60)) +
  scale_y_continuous(name = NULL, breaks = seq(-60, 60, by = 30)) +
  coord_sf(xlim = 0.95*xlim, ylim = 0.95*ylim, expand = FALSE, crs = crs_robin, ndiscr = 1000) + 
  theme_dviz_grid(font_size = 12, rel_small = 1) +
  theme(
    panel.background = element_rect(fill = "#56B4E950", color = "white", size = 1),
    panel.grid.major = element_line(color = "gray30", size = 0.25),
    axis.ticks = element_line(color = "gray30", size = 0.5/.pt)
  )
```

48 of the 50 states (also referred to as the "lower 48") are contiguous and easy to visualize at once. But two states (Alaska and Hawaii) are located a substantial distance away from the lower 48 states (Figure \@ref(fig:usa-orthographic)).


(ref:usa-orthographic) Relative locations of Alaska, Hawaii, and the lower 48 states shown on a globe.

```{r usa-orthographic, fig.width = 5.5, fig.asp = 1, fig.cap = '(ref:usa-orthographic)'}
cenlat <- 35
cenlong <- -130

draw_ocean(cenlat, cenlong, lwd = 0.25)
draw_land(map_polys$usa, cenlat, cenlong, col = "#D00000D0") 
draw_land(map_polys$world_no_usa, cenlat, cenlong, col = "#C0C0C0B0")
par(family = dviz_font_family_condensed, ps = 12)
text(
#  x = c(0.38, 0.05, -0.4),
#  y = c(0.15, 0.49, -0.1),
  x = c(0.36, -0.17, -0.4),
  y = c(0.13, 0.49, -0.1),
  labels = c("lower 48", "Alaska", "Hawaii"),
  col = c("white", "white", "black")
)
```

(ref:usa-true-albers) Visualization of the United States, using an area-preserving Albers projection (ESRI:102003, commonly used to project the lower 48 states). Alaska and Hawaii are shown in their true locations.

```{r usa-true-albers, fig.asp = 0.72, fig.cap = '(ref:usa-true-albers)'}
longs <- -180:-20
lats <- rep(89.9, length(longs))
earth_boundary <- sf::st_sfc(
    sf::st_linestring(
      cbind(longs, lats)
    ),
    crs = "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
  )
earth_boundary <- sf::st_transform(earth_boundary, crs = "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")
whiteout <- data.frame(
  x = earth_boundary[[1]][, 1],
  y = earth_boundary[[1]][, 2]
)

p <- ggplot(US_states_geoms$true_albers) + 
  geom_sf(fill = "#E69F00B0", color = "black", size = 0.5/.pt) +
  geom_polygon(
    data = whiteout, aes(x, y),
    fill = "white", color = "gray30",
    size = 0.5/.pt
  ) +
  coord_sf(xlim = c(-6721002, 2685733), ylim = c(-1634610, 4888053), expand = FALSE, ndiscr = 1000) +
  scale_x_continuous(name = "longitude", breaks = -20*c(3:10)) +
  scale_y_continuous(name = "latitude", breaks = (1:9)*10) +
  theme_dviz_grid(font_size = 12, rel_small = 1) +
  theme(
    #plot.background = element_rect(fill = "cornsilk"),
    panel.background = element_rect(fill = "#56B4E950"),
    panel.grid.major = element_line(color = "gray30", size = 0.25),
    axis.ticks = element_line(color = "gray30", size = 0.5/.pt)
  )

# work around bug in sf graticule code
b <- ggplot_build(p)
b$layout$panel_params[[1]]$graticule$x_start[11] <- 0
b$layout$panel_params[[1]]$graticule$y_start[11] <- 0.849
ggdraw(ggplot_gtable(b))

```
(ref:usa-albers) Visualization of the United States, with the states of Alaska and Hawaii moved to lie underneath the lower 48 states. Alaska also has been scaled so its linear extent is only 35% of the state's true size. (In other words, the state's area has reduced to approximately 12% of its true size.) Such a scaling is frequently applied to Alaska, to make it visually appear to be of similar size as typical midwestern or western states. However, the scaling is highly misleading, and therefore the figure has been labeled as "bad".

```{r usa-albers, fig.asp = 0.65, fig.cap = '(ref:usa-albers)'}
# standard US Albers map, with AK artificially small

# mimick color of transparent orange on top of transparent blue,
# as in previous maps drawn
# color was obtained by extraction from rendered png
brown <- "#deb664"

p <- ggplot(US_states_geoms$us_albers) + 
  geom_sf(fill = brown, color = "black", size = 0.5/.pt) +
  coord_sf(datum = NA, expand = FALSE) +
  theme_dviz_map() +
  theme(
    #plot.background = element_rect(fill = "cornsilk"),
    plot.margin = margin(6, 6, 1, 0) 
  )

stamp_bad(p)
```


(ref:usa-albers-revised) Visualization of the United States, with the states of Alaska and Hawaii moved to lie underneath the lower 48 states.

```{r usa-albers-revised, fig.asp = 0.75, fig.cap = '(ref:usa-albers-revised)'}
# revised US Albers map, with AK at its original size
ggplot(US_states_geoms$albers_revised) + 
  geom_sf(fill = brown, color = "black", size = 0.5/.pt) +
  coord_sf(datum = NA, expand = FALSE) +
  theme_dviz_map() +
  theme(
    #plot.background = element_rect(fill = "cornsilk")
  )
```

## Layers


(ref:sfbay-overview) Wind turbines in the San Francisco Bay Area. Individual wind turbines are shown as magenta-colored dots. A large wind farm, the Shiloh Wind Farm near Rio Vista, is highlighted with a black rectangle. Map tiles by Stamen Design, under CC BY 3.0. Map data by OpenStreetMap, under ODbL. Wind turbine data source: United States Wind Turbine Database

```{r sfbay-overview, fig.width = 8.5, fig.asp = 0.75, fig.cap = '(ref:sfbay-overview)'}
# From http://www.csgnetwork.com/degreelenllavcalc.html
# Length Of A Degree Of Longitude In Meters at 38deg lat
m_per_deg <- 87832.42967867786

sfbay_scale = data.frame(
  x = -122.83,
  xend = -122.83 + 10000/m_per_deg,
  y = 37.24,
  yend = 37.24,
  label = "10km"
)

sfbay_bbox <- c(left = -122.88, bottom = 37.20, right = -120.88, top = 38.31)

wind_sfbay <- wind_turbines %>%
  filter(
    xlong < sfbay_bbox["right"],
    xlong > sfbay_bbox["left"],
    ylat > sfbay_bbox["bottom"],
    ylat < sfbay_bbox["top"]
  )

shiloh_bbox <- c(left = -121.9, bottom = 38.06, right = -121.71, top = 38.20)

p1 <- ggmap(sfbay_maps$sfbay_bg)  + 
  inset_ggmap(sfbay_maps$sfbay_lines) +
  geom_point(
    data = wind_sfbay,
    aes(x = xlong, y = ylat),
    size = 0.1,
    color = "magenta",
    alpha = 1/3
  ) +
  geom_rect(
    data = data.frame(t(shiloh_bbox)),
    aes(xmin = left, xmax = right, ymin = bottom, ymax = top),
    size = 0.5,
    color = "black",
    fill = NA,
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data.frame(x = 0.5*(shiloh_bbox['left'] + shiloh_bbox['right']), y = shiloh_bbox['top'], label = "Shiloh Wind Farm"),
    aes(x, y, label = label),
    hjust = 0.512,
    vjust = -0.51,
    family = dviz_font_family,
    color = "white",
    size = 11/.pt
  ) +
  geom_text(
    data = data.frame(x = 0.5*(shiloh_bbox['left'] + shiloh_bbox['right']),
                      y = shiloh_bbox['top'], label = "Shiloh Wind Farm"),
    aes(x, y, label = label),
    hjust = 0.5,
    vjust = -0.5,
    family = dviz_font_family,
    size = 11/.pt
  ) +
  inset_ggmap(sfbay_maps$sfbay_labels) +
  geom_segment(
    data = sfbay_scale,
    aes(x, y, xend = xend, yend = yend),
    size = 1
  ) +
  geom_text(
    data = sfbay_scale,
    aes(0.5*(x+xend), y, label = label),
    hjust = 0.5,
    vjust = -0.5,
    family = dviz_font_family,
    size = 10/.pt
  ) +
  theme_dviz_map()

p1
```

(ref:sfbay-layers) The individual layers of Figure \@ref(fig:sfbay-overview). From bottom to top, the figure consists of a terrain layer, a roads layer, a layer showning the wind turbines, and a layer labeling cities and adding a scale bar. Map tiles by Stamen Design, under CC BY 3.0. Map data by OpenStreetMap, under ODbL. Wind turbine data source: United States Wind Turbine Database

```{r sfbay-layers, fig.width = 8.5, fig.asp = 0.75, fig.cap = '(ref:sfbay-layers) '}
l1 <- ggmap(sfbay_maps$sfbay_bg) + labs(subtitle = "terrain") + 
  geom_rect(
    data = data.frame(t(sfbay_bbox)),
    aes(xmin = left, xmax = right, ymin = bottom, ymax = top),
    fill = NA, color = "black",
    size = 0.5,
    inherit.aes  = FALSE
  ) +
  theme_dviz_map() +
  theme(plot.subtitle = element_text(margin = margin(0, 0, 3, 0)))

l2 <- ggmap(sfbay_maps$sfbay_lines) + labs(subtitle = "roads") + 
  geom_rect(
    data = data.frame(t(sfbay_bbox)),
    aes(xmin = left, xmax = right, ymin = bottom, ymax = top),
    fill = NA, color = "black",
    size = 0.5,
    inherit.aes  = FALSE
  ) +
  theme_dviz_map() +
  theme(plot.subtitle = element_text(margin = margin(0, 0, 3, 0)))


l3 <- ggmap(sfbay_maps$sfbay_labels) + 
  geom_segment(
    data = sfbay_scale,
    aes(x, y, xend = xend, yend = yend),
    size = .5*1
  ) +
  geom_text(
    data = sfbay_scale,
    aes(0.5*(x+xend), y, label = label),
    hjust = 0.5,
    vjust = -0.5,
    family = dviz_font_family,
    size = .5*10/.pt
  ) +
  geom_rect(
    data = data.frame(t(sfbay_bbox)),
    aes(xmin = left, xmax = right, ymin = bottom, ymax = top),
    fill = NA, color = "black",
    size = 0.5,
    inherit.aes  = FALSE
  ) +
 labs(subtitle = "city labels, scale bar") + 
 theme_dviz_map() +
 theme(plot.subtitle = element_text(margin = margin(0, 0, 3, 0)))


l4 <- ggmap(sfbay_maps$sfbay_bg) +
  geom_rect(
    data = data.frame(t(sfbay_bbox)),
    aes(xmin = left, xmax = right, ymin = bottom, ymax = top),
    fill = "white", color = "black",
    size = 0.5,
    inherit.aes  = FALSE
  ) + 
  geom_point(
    data = wind_sfbay,
    aes(x = xlong, y = ylat),
    size = .5*0.1,
    color = "magenta",
    alpha = 1/3
  ) +
  geom_rect(
    data = data.frame(t(shiloh_bbox)),
    aes(xmin = left, xmax = right, ymin = bottom, ymax = top),
    size = .5*0.5,
    color = "black",
    fill = NA,
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data.frame(x = 0.5*(shiloh_bbox['left'] + shiloh_bbox['right']), y = shiloh_bbox['top'], label = "Shiloh Wind Farm"),
    aes(x, y, label = label),
    hjust = 0.5,
    vjust = -0.5,
    family = dviz_font_family,
    size = .5*11/.pt
  ) +
  labs(subtitle = "wind turbines") +
  theme_dviz_map() +
  theme(plot.subtitle = element_text(margin = margin(0, 0, 3, 0)))

plot_grid(
  l1, NULL, l2,
  l4, NULL, l3,
  rel_widths = c(1, .05, 1)
)

# fig asp: 418/519 ~= 0.805
```


(ref:shiloh-map) Location of individual wind turbines in the Shiloh Wind Farm. Each dot highlights the location of one wind turbine. The map area corresponds to the rectangle in Figure \@ref(fig:sfbay-overview). Dots are colored by when the wind turbine was built, and the shape of the dots represents the various projects to which the individual wind turbines belong. Map tiles by Stamen Design, under CC BY 3.0. Map data by OpenStreetMap, under ODbL. Wind turbine data source: United States Wind Turbine Database

```{r shiloh-map, fig.asp = 0.75, fig.cap = '(ref:shiloh-map)'}
# From http://www.csgnetwork.com/degreelenllavcalc.html
# Length Of A Degree Of Longitude In Meters at 38deg lat
m_per_deg <- 87832.42967867786

shiloh_scale = data.frame(
  x = -121.735,
  xend = -121.735 + 2000/m_per_deg,
  y = 38.064,
  yend = 38.064,
  label = "2000m"
)

#bbox <- c(left = -121.9, bottom = 38.06, right = -121.71, top = 38.20)
wind_shiloh <- wind_turbines %>%
  filter(
    xlong < shiloh_bbox["right"],
    xlong > shiloh_bbox["left"],
    ylat > shiloh_bbox["bottom"],
    ylat < shiloh_bbox["top"]
  ) %>%
  mutate(
    name = fct_relevel(fct_collapse(p_name,
      `EDF Renewables` = "EDF Renewable V",
      `High Winds` = "High Winds",
      `Shiloh` = c("Shiloh Wind Project", "Shiloh II", "Shiloh III", "Shiloh IV"),
      `Solano` = c("Solano Phase 3", "Solano Phase IIA", "Solano Wind Project", "Solano Wind Project, Phase I", "Solano Wind Project, Phase IA"),
      `other` = c("Montezuma", "Montezuma Winds II", "unknown Solano County")
    ), "EDF Renewables", "High Winds", "Shiloh", "Solano", "other"),
    year_range = cut(
      p_year,
      breaks = c(1980, 2000, 2005, 2010, 2015),
      labels = c("before 2000", "2000 to 2004", "2005 to 2009", "2010 to 2014"),
      right = FALSE
    )
  )

p2 <- ggmap(sfbay_maps$shiloh_terrain)  + 
  geom_point(
    data = wind_shiloh,
    aes(x = xlong, y = ylat, fill = year_range, shape = name),
    size = 1.5,
    color = "black", stroke = 0.2
  ) +
  geom_segment(
    data = shiloh_scale,
    aes(x, y, xend = xend, yend = yend),
    size = 1
  ) +
  geom_text(
    data = shiloh_scale,
    aes(0.5*(x+xend), y, label = label),
    hjust = 0.5,
    vjust = -0.5,
    family = dviz_font_family,
    size = 10/.pt
  ) +
  xlab(label = NULL) +
  ylab(label = NULL) +
  scale_fill_viridis_d(
    name = "year built",
    option = "A", end = .95, begin = 0.3, direction = -1,
    guide = guide_legend(
      order = 2,
      reverse = FALSE,
      override.aes = list(shape = 22, size = 4, stroke = 0))
  ) +
  scale_shape_manual(
    name = "project name",
    values = 21:25,
    guide = guide_legend(
      order = 1,
      override.aes = list(
        fill = "grey70",
        size = 2
      )
    )
  ) +
  theme_dviz_map(12) +
  theme(
    legend.key.width = grid::unit(12, "pt")
  )

p2
```

*Also show wind turbines in all of the lower 48, using some sort of bubble-grid approach?*

## Choropleth mapping

(ref:population-density-counties) Population density in every U.S. county, shown as a choropleth map. Population density is reported as persons per square kilometer. Data source: 2015 Five-Year American Community Survey

```{r population-density-counties, fig.asp = 0.73, fig.cap = '(ref:population-density-counties)'}
# x range: -3683715  2258154
# y range: -2839538  1558935

US_counties_income <- mutate(US_counties_income, logdens = log(as.numeric(popdens)*1e6))

p <- ggplot(US_counties_income) + 
  geom_sf(aes(color = logdens, fill = logdens), size = 0.1) + 
  geom_sf(data = US_states_geoms$albers_revised, fill = NA, color = "grey30", size = 0.2) +
  coord_sf(datum = NA, expand = FALSE) +
  scale_x_continuous(limits = c(-4000000, 2300000)) +
  scale_fill_continuous_sequential(
    aesthetics = c("color", "fill"),
    palette = "YlGnBu", rev = TRUE, cmax = 20, c2 = 20, p2 = 1.75,
    name = "population density\n(persons / square km)",
    limits = log(c(0.01, 30000)),
    breaks = log(c(0.01, 1, 100, 10000)),
    labels = c("0.01", "1", "100", "10,000"),
    guide = guide_colorbar(
      frame.colour = "black",
      ticks.colour = "white",
      barwidth = grid::unit(15, "pt"),
      barheight = grid::unit(90, "pt")
    )
  ) +
  theme_dviz_map(12, rel_small = 1) +
  theme(
    #plot.background = element_rect(fill = "cornsilk"),
    legend.position = c(0, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = grid::unit(3, "pt"),
    legend.title = element_text(hjust = 0.5)
  )

ggdraw(align_legend(p))
```

(ref:population-density-counties2) Population density in every U.S. county, shown as a choropleth map. This map is identical to Figure \@ref(fig:population-density-counties) except that now the color scale uses light colors for high population densities and dark colors for low population densities. Data source: 2015 Five-Year American Community Survey

```{r population-density-counties2, fig.asp = 0.73, fig.cap = '(ref:population-density-counties2)'}
# x range: -3683715  2258154
# y range: -2839538  1558935

p <- ggplot(US_counties_income) + 
  geom_sf(aes(color = logdens, fill = logdens), size = 0.1) +
  # state boundaries don't look good with this color scale
  geom_sf(data = US_states_geoms$albers_revised, fill = NA, color = "black", size = 0.2) +
  coord_sf(datum = NA, expand = FALSE) +
  scale_x_continuous(limits = c(-4000000, 2300000)) +
  scale_fill_continuous_sequential(
    aesthetics = c("color", "fill"),
    palette = "Lajolla", rev = TRUE, p1 = 2, p2 = 1.3,
    name = "population density\n(persons / square km)",
    limits = log(c(0.01, 30000)),
    breaks = log(c(0.01, 1, 100, 10000)),
    labels = c("0.01", "1", "100", "10,000"),
    guide = guide_colorbar(
      frame.colour = "black",
      ticks.colour = "white",
      barwidth = grid::unit(15, "pt"),
      barheight = grid::unit(90, "pt")
    )
  ) +
  theme_dviz_map(12, rel_small = 1) +
  theme(
    #plot.background = element_rect(fill = "cornsilk"),
    legend.position = c(0, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = grid::unit(3, "pt"),
    legend.title = element_text(hjust = 0.5)
  )

ggdraw(align_legend(p))
```


(ref:median-income-counties) Median income in every U.S. county, shown as a choropleth map. Data source: 2015 Five-Year American Community Survey


```{r median-income-counties, fig.asp = 0.73, fig.cap = '(ref:median-income-counties)'}
# x range: -3683715  2258154
# y range: -2839538  1558935

p <- ggplot(US_counties_income) + 
  geom_sf(aes(color = median_income, fill = median_income), size = 0.1) + 
  geom_sf(data = US_states_geoms$albers_revised, fill = NA, color = "grey30", size = 0.2) +
  coord_sf(datum = NA, expand = FALSE) +
  scale_x_continuous(limits = c(-4000000, 2300000)) +
  scale_fill_continuous_sequential(
    aesthetics = c("color", "fill"),
    #palette = "Lajolla", rev = TRUE, p1 = 2, p2 = 1.3,
    #palette = "BlueYellow", rev = TRUE, l1 = 15, p2 = 1.7,
    h1 = -83, h2 = 20, c1 = 30, cmax = 40, c2 = 0, l1 = 20, l2 = 100, p1 = 1, p2 = 1.2, rev = TRUE, 
    name = "median income",
    limits = c(0, 130000),
    breaks = c(0, 50000, 100000),
    labels = c("$0", "$50,000", "$100,000"),
    guide = guide_colorbar(
      frame.colour = "black",
      ticks.colour = "white",
      barwidth = grid::unit(15, "pt"),
      barheight = grid::unit(90, "pt")
    )
  ) +
  theme_dviz_map(12, rel_small = 1) +
  theme(
    #plot.background = element_rect(fill = "cornsilk"),
    legend.position = c(0, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = grid::unit(3, "pt"),
    legend.title = element_text(hjust = 0.5)
  )

ggdraw(align_legend(p))
```

```{r median-income-counties-binned, fig.asp = 0.73, fig.cap = '(ref:median-income-counties-binned)'}
# x range: -3683715  2258154
# y range: -2839538  1558935

US_counties_income <- mutate(
  US_counties_income,
  income_bins = cut(
      ifelse(is.na(median_income), 25000, median_income), # hide missing value
      breaks = c(0, 30000, 55000, 80000, 105000, 150000),
      labels = c("< $30k", "$30k to $60k", "$60k to $85k", "$85k to $105k", "> $105k"),
      right = FALSE
    )
  )

p <- ggplot(US_counties_income) + 
  geom_sf(aes(color = income_bins, fill = income_bins), size = 0.1) + 
  geom_sf(data = US_states_geoms$albers_revised, fill = NA, color = "grey30", size = 0.2) +
  coord_sf(datum = NA, expand = FALSE) +
  scale_x_continuous(limits = c(-4000000, 2300000)) +
  scale_fill_discrete_sequential(
    aesthetics = c("color", "fill"),
    h1 = -83, h2 = 20, c1 = 30, cmax = 40, c2 = 0, l1 = 20, l2 = 100, p1 = 1, p2 = 1.2, rev = TRUE, 
    name = "median income",
    nmax = 6,
    order = 2:6,
    guide = guide_legend(
      override.aes = list(colour = "white", size = 1)
    )
  ) +
  theme_dviz_map(12, rel_small = 1) +
  theme(
    #plot.background = element_rect(fill = "cornsilk"),
    legend.position = c(0, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = grid::unit(3, "pt"),
    legend.spacing.y = grid::unit(3, "pt"),
    legend.title = element_text(hjust = 0.5),
    legend.key.width = grid::unit(18, "pt"),
    legend.key.height = grid::unit(15, "pt")
  )

p
```

(ref:median-income-states) Median income in every U.S. state, shown as a choropleth map. **Say why this is bad.** Data source: 2015 Five-Year American Community Survey


```{r median-income-states, fig.asp = 0.73, fig.cap = '(ref:median-income-states)'}
# x range: -3683715  2258154
# y range: -2839538  1558935

US_income <- mutate(
  US_income,
  income_bins = cut(
      ifelse(is.na(median_income), 25000, median_income), # hide missing value
      breaks = c(0, 40000, 50000, 60000, 70000, 80000),
      labels = c("< $40k", "$40k to $50k", "$50k to $60k", "$60k to $70k", "> $70k"),
      right = FALSE
    )
  )

p <- ggplot(US_income, aes(fill = income_bins)) + 
  geom_sf(color = "grey30", size = 0.2) + 
  coord_sf(datum = NA, expand = FALSE) +
  scale_x_continuous(limits = c(-4000000, 2300000)) +
  scale_fill_discrete_sequential(
    h1 = -83, h2 = 20, c1 = 30, cmax = 40, c2 = 0, l1 = 20, l2 = 100, p1 = 1, p2 = 1.2, rev = TRUE, 
    name = "median income",
    nmax = 7,
    order = 2:6,
    guide = guide_legend(
      override.aes = list(colour = "white", size = 1)
    )
  ) +
  theme_dviz_map(12, rel_small = 1) +
  theme(
    #plot.background = element_rect(fill = "cornsilk"),
    legend.position = c(0, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = grid::unit(3, "pt"),
    legend.spacing.y = grid::unit(3, "pt"),
    legend.title = element_text(hjust = 0.5),
    legend.key.width = grid::unit(18, "pt"),
    legend.key.height = grid::unit(15, "pt")
  )

stamp_bad(p)
```


## Cartograms

(ref:median-income-cartogram) Median income in every U.S. state, shown as a cartogram. The shapes of individual states have been modified such that their area is proportional to their number of inhabitants. Data source: 2015 Five-Year American Community Survey

```{r median-income-cartogram, fig.asp = 0.73, fig.cap = '(ref:median-income-cartogram)'}
# copy binnned data over, order of both data frames is the same
US_income_cartogram$income_bins <- US_income$income_bins

p <- ggplot(US_income_cartogram, aes(fill = income_bins)) + 
  geom_sf(color = "grey30", size = 0.2) + coord_sf(datum = NA, expand = FALSE) +
  scale_x_continuous(limits = c(-3900000, 2500000)) +
  scale_fill_discrete_sequential(
    h1 = -83, h2 = 20, c1 = 30, cmax = 40, c2 = 0, l1 = 20, l2 = 100, p1 = 1, p2 = 1.2, rev = TRUE, 
    name = "median income",
    nmax = 7,
    order = 2:6,
    guide = guide_legend(
      override.aes = list(colour = "white", size = 1)
    )
  ) +
  theme_dviz_map(12, rel_small = 1) +
  theme(
    #plot.background = element_rect(fill = "cornsilk"),
    legend.position = c(0, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = grid::unit(3, "pt"),
    legend.spacing.y = grid::unit(3, "pt"),
    legend.title = element_text(hjust = 0.5),
    legend.key.width = grid::unit(18, "pt"),
    legend.key.height = grid::unit(15, "pt")
  )

p
```

(ref:median-income-statebins) Median income in every U.S. state, shown as a cartogram heatmap. Each state is represented by an equally sized square, and the squares are arranged according to the approximate position of each state relative to the other states. This representation gives the same visual weight to each state. Data source: 2015 Five-Year American Community Survey

```{r median-income-statebins, fig.asp = 0.62, fig.cap = '(ref:median-income-statebins)'}
filter(US_income, name != "Puerto Rico", GEOID != "11") %>% # remove Puerto Rico and DC
  ggplot(aes(state = name, fill = income_bins)) +
  geom_statebins(family = dviz.supp::dviz_font_family,
                 lbl_size = 14/.pt) +
  expand_limits(x = -1.3) + # make space for legend
  coord_equal(expand = FALSE) +
   scale_fill_discrete_sequential(
    h1 = -83, h2 = 20, c1 = 30, cmax = 40, c2 = 0, l1 = 20, l2 = 100, p1 = 1, p2 = 1.2, rev = TRUE, 
    name = "median income",
    nmax = 7,
    order = 2:6,
    guide = guide_legend(
      override.aes = list(colour = "white", size = 1)
    )
  ) +
  theme_dviz_map(12, rel_small = 1) +
  theme(
    #plot.background = element_rect(fill = "cornsilk"),
    legend.background = element_blank(),
    legend.position = c(0, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = grid::unit(3, "pt"),
    legend.spacing.y = grid::unit(3, "pt"),
    legend.title = element_text(hjust = 0.5),
    legend.key.width = grid::unit(18, "pt"),
    legend.key.height = grid::unit(15, "pt")
  )
```

*Add a geofacet example?*
