```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")
```


# Introduction {#introduction}

*Write about visualizations vs. figures. This book is specifically about static visualizations presented in print, online, or in presentations. The book does not cover interactive visuals or movies. For this reason, we will use the words "visualization" and "figure" somewhat interchangeably.*


Data visualization is part art and part science. The challenge is to get the art right without getting the science wrong and vice versa. A data visualization first and foremost has to be correct, which means it must not mislead about or distort the data. If one number is twice as large as another, but in the visualization they look to be about the same, then the visualization is incorect. At the same time, a data visualization should be aesthetically pleasing. Good visual presentations tend to enhance the message of the visualization. If a figure contains jarring colors, imbalanced visual elements, or other features that distract, then the viewer will find it harder to inspect the figure and interpret it correctly.

Scientists typically know how to visualize data without being grossly misleading, but they may not have a well developed sense of visual aesthetics and they may inadvertantly make visual choices that detract from their desired message. Designers, on the other hand, may prepare visualizations that look beautiful but play fast and lose with the data.

This book consists of three parts. The first, "From data to visualization," describes different types of visualization, such as bar graphs, scatter plots, or pie charts. With this part, I am not attempting to provide encyclopedic coverage of every conceivable visualization approach. Instead, I discuss a core set of approaches that you will likely encounter in publications and/or need in your own work. In organizing this part, I have attempted to group visualizations by the type of message they convey instead of by the type of data being visualized. Statistical texts often describe data analysis and visualization by type of data, organizing the material by number and type of variables (one continuous variable, one discrete variable, two continuous variables, one continuous and one discrete variable, etc.) I believe that only statisticians find this organization helpful. Most other people think in terms of the type of message they want to convey, such as how large something is, how it is composed of parts, how it relates to something else, and so on.

The second part, "Principles of figure design," discusses various design issues that arise when assembling data visualizations. **write a bit more**

The third part, "Miscellaneous topics," covers a few remaining issues that didn't fit into the first two parts.  **complete**.


```{r fig.asp = 3/4}
df <- tibble(
  type = c("A", "B", "C"),
  value = c(3, 5, 4),
  expand = c(4, 5, 4.5)
)


p1 <- ggplot(df, aes(type, value)) + 
  geom_col(fill = "#56B4E9", width = 0.65, alpha = 0.9) +
  scale_y_continuous(limits = c(0, 5.2), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_dviz_hgrid(12) +
  theme(
    axis.line = element_blank(),
    plot.margin = margin(18, 12, 0, 0)
  )

p2 <- ggplot(df, aes(type, value)) + 
  geom_col(fill = c("#CCFF00FF", "#00FFFFFF", "#CC00FFFF"), width = 0.75) +
  scale_y_continuous(limits = c(0, 5.2), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_dviz_hgrid(12) +
  theme(
    axis.line = element_blank(),
    axis.title = element_text(family = "Comic Sans MS", size = 15),
    axis.text.x = element_text(family = "Times", size = 10),
    axis.text.y = element_text(family = "Arial", size = 13),
    panel.grid = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.margin = margin(18, 12, 0, 0)
  )


p3 <- ggplot(df, aes(type, value)) + 
  geom_col(fill = "#56B4E9", width = 0.65, alpha = 0.9) +
  geom_point(aes(y = expand), shape = NA) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(~type, scales = "free") +
  coord_cartesian(clip = "off") +
  theme_dviz_hgrid(12) +
  theme(
    axis.line = element_blank(),
    strip.text = element_blank(),
    plot.margin = margin(18, 12, 0, 0)
  )

p3a <- ggplot(df, aes(type, value)) + 
  geom_col(color = "black", fill = NA, width = .5) +
  scale_y_continuous(limits = c(0, 5.2), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_dviz_grid(12) +
  background_grid(
    major = "y", minor = "none",
    colour.major = "grey30", colour.minor = "black",
    size.major = 0.5,
    size.minor = 0.2
  ) +
  theme(
    axis.ticks = element_line(color = "grey30"),
    plot.margin = margin(18, 12, 0, 0)
  )

p4 <- ggplot(df, aes(type, value)) + 
  geom_col(fill = "#56B4E9", width = 0.65, alpha = 0.9) +
  coord_cartesian(xlim = c(0.4, 3.6), ylim = c(2, 6.2), expand = FALSE, clip = "on") +
  theme_dviz_hgrid(12) +
  theme(plot.margin = margin(18, 12, 0, 0))

plot_grid(
  p1, NULL, stamp_ugly(p2),
  NULL, NULL, NULL,
  stamp_sketchy(p3), NULL, stamp_bad(p4),
  rel_widths = c(1, .04, 1),
  rel_heights = c(1, .04, 1),
  labels = c("a", "", "b", "", "", "", "c", "", "d")
)
```
