```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")

library(readr)
library(lubridate)
library(ggrepel)
```

# Optimize the data--ink ratio {#maximize-data-signal}

*The order of chapters should be: (i) optimize data--ink ratio; (ii) small axis labels; (iii) avoid line drawings.*

We can broadly subdivide the graphical components in any visualization into components that represent data and components that do not. The former are things such as the points in a scatter plot, the bars in a histogram or barplot, and the shaded areas in a heatmap. The latter are things such as plot axes, axis ticks and labels, axis titles, legends, and plot annotations. As a general rule, most of the ink (Chapter \@ref(proportional-ink)) in a plot should be devoted to displaying data. Remove unnecessary frames, lines, or other adornments. For non-data elements that can't be removed (such as, for example, a legend), make sure they aren't overly prominent and stand back relative to the data.

## Finding the appropriate data--ink ratio

The idea that one should remove non-data visual elements from graphs was popularized by Edward Tufte in his book "The Visual Display of Quantitative Information" [@TufteQuantDispl]. Tufte introduces the concept of the "data--ink ratio", which he defines as the "proportion of a graphic's ink devoted to the non-redundant display of data-information." He then writes:

> Maximize the data--ink ratio, *within reason.*

(emphasis mine) I have emphasized the phrase "within reason" because it is critical and frequently forgotten. Removing non-data ink is valuable up to a point. Beyond that point, making a graph more minimal will make it less compelling.

First, let's consider a figure that clearly has too much non-data ink (Figure \@ref(fig:Aus-athletes-grid-bad)). In this figure, the colored points in the plot panel (the framed center area containing data points) are data ink. Everything else is non-data ink. The non-data ink includes a frame around the entire figure, a frame around the plot panel, and a frame around the legend. None of these frames are needed. We also see a prominent and dense background grid that draws attention away from the actual data points. By removing the frames and minor grid lines and by drawing the major grid lines in a light gray, we arrive at Figure \@ref(fig:Aus-athletes-grid-good). In this version of the figure, the actual data points stand out much more clearly, and they are perceived as the most important component of the figure.

(ref:Aus-athletes-grid-bad) Percent body fat versus height in professional male Australian athletes. Each point represents one athlete. This figure devotes way too much ink to non-data. There are unnecessary frames around the entire figure, around the plot panel, and around the legend. The coordinate grid is very prominent, and its presence draws attention away from the data points.

```{r Aus-athletes-grid-bad, fig.cap='(ref:Aus-athletes-grid-bad)'}
male_Aus <- filter(Aus_athletes, sex=="m") %>%
    filter(sport %in% c("basketball", "field", "swimming", "track (400m)",
                        "track (sprint)", "water polo")) %>%
    mutate(sport = case_when(sport == "track (400m)" ~ "track",
                             sport == "track (sprint)" ~ "track",
                             TRUE ~ sport))

male_Aus$sport <- factor(male_Aus$sport,
                         levels = c("field", "water polo", "basketball", "swimming", "track"))

p_Aus_base <- ggplot(male_Aus, aes(x=height, y=pcBfat, color=sport, fill=sport, shape=sport)) +
  geom_point(size = 2.5) +
  scale_shape_manual(values = 21:25) +
  scale_color_OkabeIto(order=c(2, 1, 3, 4, 5), darken = 0.3) +
  scale_fill_OkabeIto(order=c(2, 1, 3, 4, 5), darken = 0.1, alpha = 0.7) +
  scale_x_continuous(limits = c(169, 210)) +
  scale_y_continuous(limits = c(5, 20), name = "% body fat")

p <- p_Aus_base +
  theme_dviz_open() +
  theme(
    axis.line = element_blank(),
    panel.grid.major = element_line(color = "black", size = 0.3),
    panel.grid.minor = element_line(color = "black", size = 0.15),
    panel.border = element_rect(color = "black", size = 1),
    legend.background = element_rect(color = "black", size = 0.5),
    legend.margin = margin(7, 7, 7, 7),
    plot.background = element_rect(color = "black", size = 1),
    plot.margin = margin(7, 7, 7, 7)
  )

# width: 6 in = 433.62pt
# height: 0.618 * 6 in = 3.71 in = 268.12

stamp_bad(ggdraw() + draw_plot(p, x = 1/434, y = 1/268, width = 426/434, height = 252/268))
```

(ref:Aus-athletes-grid-good) Percent body fat versus height in professional male Australian athletes. This figure is a cleaned-up version of Figure \@ref(fig:Aus-athletes-grid-bad). Unnecessary frames have been removed, minor grid lines have been removed, and majore grid lines have been drawn in light gray to stand back relative to the data points.

```{r Aus-athletes-grid-good, fig.cap='(ref:Aus-athletes-grid-good)'}
p_Aus_base + theme_dviz_grid()
```

However, we can take the removal of non-data ink too far. Figure \@ref(fig:Aus-athletes-min-bad) is a minimalist version of Figure \@ref(fig:Aus-athletes-grid-good), and a clear regression. Most importantly, the axis tick labels and titles have been made so faint that they are hard to see. If we just glance at the figure we will not immediately perceive what data is actually shown, we only see points floating in space. Second, the legend annotations are so faint that the points in the legend could be mistaken for data points. This effect is amplified because there is no clear visual separation between the plot area and the legend. Notice how the background grid in Figure \@ref(fig:Aus-athletes-grid-good) both anchors the points in space and sets off the data area from the legend area. Both of these effects have been lost in Figure \@ref(fig:Aus-athletes-min-bad).

(ref:Aus-athletes-min-bad) Percent body fat versus height in professional male Australian athletes. In this example, the concept of maximizing the data--ink ratio has been taken too far. The axis tick labels and title are too faint and are barely visible. The data points seem to float in space. The points in the legend are not sufficiently set off from the data points, and the casual observer might think they are part of the data.

```{r Aus-athletes-min-bad, fig.cap='(ref:Aus-athletes-min-bad)'}
axis_col <- "grey70"
title_col <- "grey70"

p <- p_Aus_base + theme_dviz_open(11) +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_line(color = axis_col),
    axis.text = element_text(color = axis_col),
    axis.title = element_text(color = title_col),
    legend.text = element_text(color = title_col),
    legend.title = element_text(color = title_col),
    plot.margin = margin(3.5, 7, 3.5, 0)
  )

stamp_bad(p)
```

In my experience, the most common problem of plots with too little non-data ink are figure elements that appear to float in space, without clear connection or reference to anything. This problem tends to be particularly severe in small multiples plots. Figure \@ref(fig:titanic-survival-by-gender-class-bad) shows a small-multiples plot comparing six different bar plots, but it looks more like a piece of modern art than a useful data visualization. The bars appear to be floating, they are not anchored to a clear baseline, and the individual plot facets are not clearly delineated. We can resolve all these issues by adding a light gray background and thin horizontal grid lines to each facet (Figure \@ref(fig:titanic-survival-by-gender-class)).

(ref:titanic-survival-by-gender-class-bad) Survival of passengers on the Titanic, broken down by gender and class. This small-multiples plot is too minimalistic. The individual factes are not framed, so it's difficult to see which part of the figure belongs to which facet. Further, the individual bars are not anchored to a clear baseline, and they seem to float.

```{r titanic-survival-by-gender-class-bad, fig.width = 5, fig.asp = 3/4, fig.cap = '(ref:titanic-survival-by-gender-class-bad)'}
titanic %>% mutate(surv = ifelse(survived == 0, "died", "survived")) %>%
  ggplot(aes(sex, fill = sex)) + geom_bar() +
    facet_grid(class ~ surv, scales = "free_x") +
    scale_x_discrete(name = NULL) + 
    scale_y_continuous(limits = c(0, 195), expand = c(0, 0)) +
    scale_fill_manual(values = c("#D55E00D0", "#0072B2D0"), guide = "none") +
    theme_dviz_open(rel_small = 1) +
    theme(axis.line.x = element_blank(),
          axis.ticks.x = element_blank(),
          #axis.text.x = element_text(margin = margin(7, 0, 0, 0)),
          strip.text = element_text(margin = margin(3.5, 3.5, 3.5, 3.5)),
          strip.background  = element_blank(),
          plot.margin = margin(3.5, 14, 3.5, 0)) -> p

stamp_bad(p)
```


(ref:titanic-survival-by-gender-class) Survival of passengers on the Titanic, broken down by gender and class. This is an improved version of Figure \@ref(fig:titanic-survival-by-gender-class-bad). The gray background in each facet clearly delineates the six groupings (survived or died in first, second, or third class) that make up this plot. Thin horizontal lines in the background provide a reference for the bar heights and facility comparison of bar heights among facets.

```{r titanic-survival-by-gender-class, fig.width = 5, fig.asp = 3/4, fig.cap = '(ref:titanic-survival-by-gender-class)'}
colors <- desaturate(lighten(c("#56B4E9", "#56B4E9"), c(.8, -.4)), .8)

titanic %>% mutate(surv = ifelse(survived == 0, "died", "survived")) %>%
  ggplot(aes(sex, fill = sex)) + geom_bar() +
    facet_grid(class ~ surv, scales = "free_x") +
    scale_x_discrete(name = NULL) + 
    scale_y_continuous(limits = c(0, 195), expand = c(0, 0)) +
    scale_fill_manual(values = c("#D55E00D0", "#0072B2D0"), guide = "none") +
    theme_dviz_open(rel_small = 1) +
    theme(axis.line = element_blank(),
          #axis.line.y = element_line(color = colors[2]),
          axis.ticks = element_blank(),
          axis.ticks.length = grid::unit(0, "pt"),
          #axis.ticks.y = element_line(color = colors[2]),
          axis.text.x = element_text(margin = margin(7, 0, 3.5, 0)),
          axis.text.y = element_text(color = colors[2]),
          strip.text = element_text(margin = margin(3.5, 3.5, 3.5, 3.5)),
          strip.background  = element_blank(),
          panel.background = element_rect(fill = colors[1]),
          panel.grid.major.y = element_line(color = "white"),
          plot.margin = margin(3.5, 14, 3.5, 0)) -> p

p
```

Since removing too much non-data ink can be just as bad as adding too much of it, I want to propose an alternative to Tufte's maxim. We need to optimize, rather than maximize, the data--ink ratio. Make sure you don't overload your plot with non-data ink, such that the data remains in the foreground, but don't take it to the point where the data loses context or floats off into space.

```{block type='rmdtip', echo=TRUE}
Optimize the data--ink ratio.
```

## Background grids and axis lines

We can't talk about data--ink ratios without touching on background grids, axis lines, axis ticks, and labels. By Tufte's definition, these elements are not data and thus decrease the data--ink ratio. However, they do carry critical information about the data, and hence are not superfluous. In fact, one of the first rules of data visualization is to label your axes. And if labels are required, then making them sufficiently dark and prominent that they are actually visible and decipherable is also a requirement.

Whether a background grid is required is less clear-cut, and reasonable people can disagree about what the best options are for a given visualization. The R software ggplot2 has popularized a style using a fairly prominent background grid of white lines on a gray background (Figure \@ref(fig:price-plot-ggplot-default)). With apologies to the ggplot2 author Hadley Wickham, who developed this style and for whom I have the utmost respect, I don't find this style particularly attractive. In my experience the gray background frequently detracts from the actual data, and a grid with major and minor lines can be too dense. I also find the gray squares in the legend confusing.

(ref:price-plot-ggplot-default) Stock price over time for four major tech companies. The stock price for each company has been normalized to equal 100 in June 2012. This figure mimicks the ggplot2 default look, with white major and minor grid lines on a gray background. In this particular example, I think the grid lines overpower the data lines, and the result is a figure that is not well balanced and that doesn't place sufficient emphasis on the data.

```{r price-plot-ggplot-default, fig.cap='(ref:price-plot-ggplot-default)'}
price_plot <- ggplot(tech_stocks, aes(x=date, y=price_indexed, color=ticker)) +
  geom_line() +
  scale_color_manual(values=c("#000000", "#E69F00", "#56B4E9", "#009E73"),
                     name="",
                     breaks=c("FB", "GOOG", "MSFT", "AAPL"),
                     labels=c("Facebook", "Alphabet", "Microsoft", "Apple")) +
  scale_x_date(name="year",
               limits=c(ymd("2012-06-01"), ymd("2017-05-31")),
               expand=c(0,0)) + 
  scale_y_continuous(name="stock price, indexed",
                     limits = c(0, 560),
                     expand=c(0,0)) +
  theme_dviz_grid() +
  theme(panel.background = element_rect(fill = "grey90"),
        panel.grid.major = element_line(color = "white"),
        panel.grid.minor = element_line(color = "white", size = rel(0.5)),
        legend.key = element_rect(fill = "grey90", color = "white"),
        axis.ticks = element_line(color = "black"),
        plot.margin = margin(7, 7, 3, 0))

stamp_ugly(price_plot)
```

Arguments in favor of the gray background include that it (i) helps the plot to be perceived as a single visual entity and (ii) prevents the plot to appear as a white box in surrounding dark text [@Wickham2016]. I completely agree with the first point, and it was the reason I used gray backgrounds in Figure \@ref(fig:titanic-survival-by-gender-class). For the second point, I'd like to caution that the perceived darkness of text will depend on the font size, fontface, and line spacing, and the perceived darkness of a figure will depend on the absolute amount and color of ink used, including all data ink. A scientific paper typeset in dense, 10-point Times New Roman will look much darker than a coffee-table book typeset in 14-point Palatino with one-and-a-half line spacing. Likewise, a scatter plot of five points in yellow will look much lighter than a scatter plot of 10,000 points in black. If you want to use a gray figure background, consider the color intensity of your figure foreground, as well as the expected layout and typography of the text around your figures, and adjust the choice of your background gray accordingly. Otherwise, it could happen that your figures end up standing out as dark boxes among the surrounding lighter text. Also, keep in mind that the colors you use to plot your data need to work with the gray background. We tend to perceive colors differently against different backgrounds, and a gray background requires darker and more saturated foreground colors than a white background.

We can go all the way in the opposite direction and remove both the background and the grid lines (Figure \@ref(fig:price-plot-no-grid)). In this case, we need visible axis lines to frame the plot and keep it as a single visual unit. For this particular figure, I think this choice is a worse option, and I have labeled it as "bad". In the absence of any background grid whatsoever, the curves seem to float in space, and it's difficult to see where they go.

(ref:price-plot-no-grid) Indexed stock price over time for four major tech companies.

```{r price-plot-no-grid, fig.cap='(ref:price-plot-no-grid)'}
stamp_bad(price_plot + 
            theme_dviz_open() +
            theme(plot.margin = margin(7, 7, 3, 0)))
```

**edit from here.**

Since all prices are indexed to 100 in June 2012, at a minimum this value should be marked in the plot. Thus, one option would be to add a thin horizontal line at *y* = 100 (Figure \@ref(fig:price-plot-refline)). Alternatively, we can use a minimal "grid" of horizontal lines. For a plot where we are primarily interested in the change in *y* values, vertical grid lines are not needed. Moreover, grid lines positioned at only the major axis ticks will often be sufficient. And, the axis line can be omitted or made very thin, since the horzontal lines clearly mark the extent of the plot (Figure \@ref(fig:price-plot-hgrid)).

(ref:price-plot-refline) Indexed stock price over time for four major tech companies.

```{r price-plot-refline, fig.cap='(ref:price-plot-refline)'}
price_plot2 <- ggplot(tech_stocks, aes(x=date, y=price_indexed, color=ticker)) +
  geom_hline(yintercept = 100, size = 0.5, color="grey70") +
  geom_line() +
  scale_color_manual(values=c("#000000", "#E69F00", "#56B4E9", "#009E73"),
                     name="",
                     breaks=c("FB", "GOOG", "MSFT", "AAPL"),
                     labels=c("Facebook", "Alphabet", "Microsoft", "Apple")) +
  scale_x_date(name="year",
               limits=c(ymd("2012-06-01"), ymd("2017-05-31")),
               expand=c(0,0)) + 
  scale_y_continuous(name="stock price, indexed",
                     limits = c(0, 560),
                     expand=c(0,0)) +
  theme_dviz_open() +
  theme(plot.margin = margin(7, 7, 3, 0))

price_plot2
```


(ref:price-plot-hgrid) Indexed stock price over time for four major tech companies.

```{r price-plot-hgrid, fig.cap='(ref:price-plot-hgrid)'}
price_plot + theme_dviz_hgrid() +
  theme(plot.margin = margin(7, 7, 3, 0))
```

For such a minimal grid, we generally draw the lines orthogonally to direction along which the numbers of interest vary. Therefore, if instead of plotting the stock price over time we plot the five-year increase, as horizontal bars, then we will want to use vertical lines instead (Figure \@ref(fig:price-increase)).

(ref:price-increase) Percent increase in stock price from June 2012 to June 2017, for four major tech companies.

```{r price-increase, fig.cap='(ref:price-increase)'}
perc_increase <- filter(tech_stocks, date == ymd("2017-06-01")) %>%
  mutate(perc=100*(price-index_price)/index_price,
         label=paste(as.character(round(perc)), "%", sep="")) %>%
  arrange(perc)

perc_increase$ticker <- factor(perc_increase$ticker, levels=perc_increase$ticker)

perc_plot <- ggplot(perc_increase, aes(x=ticker, y=perc)) +
  geom_col(fill="#56B4E9") +
  geom_text(aes(label=label), color="white", hjust=1.1, size=5,
            family = dviz_font_family) +
  scale_y_continuous(#name="percent increase\n(June 2012 to June 2017)",
                     name="percent increase",
                     limits=c(0, 499),
                     expand=c(0, 0)) +
  scale_x_discrete(name = NULL,
                     breaks = c("FB", "GOOG", "MSFT", "AAPL"),
                     labels = c("Facebook ", "Alphabet ", "Microsoft ", "Apple ")) +
  coord_flip(clip = "off") +
  theme_dviz_vgrid(rel_small = 1) +
  theme(axis.line.y = element_blank(),
        axis.ticks = element_blank(),
        axis.ticks.length = grid::unit(0, "pt"))

perc_plot
```


```{block type='rmdtip', echo=TRUE}
Grid lines that run perpendicular to the key variable of interest tend to be the most useful.
```

(ref:price-increase-bad) **Figure caption here.** The white grid lines on top of the bars make it look like the bars fall apart.

```{r price-increase-bad, fig.cap='(ref:price-increase-bad)'}
perc_plot2 <- ggplot(perc_increase, aes(x = ticker, y = perc)) +
  geom_col(fill = "#56B4E9", width = 0.8) +
  geom_hline(yintercept = c(100, 200, 300, 400), color = "white", size = 1) +
  geom_text(
    aes(label = label), color = "black", hjust = -0.1, size = 5,
    family = dviz_font_family
  ) +
  scale_y_continuous(
    name = NULL,
    #name = "percent increase",
    limits = c(0, 496),
    expand = c(0, 0)
  ) +
  scale_x_discrete(
    name = NULL,
    breaks = c("FB", "GOOG", "MSFT", "AAPL"),
    labels = c("Facebook", "Alphabet", "Microsoft", "Apple")
  ) +
  coord_flip() +
  theme_dviz_vgrid(rel_small = 1) +
  theme(
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_text(color = "black"),
    panel.grid.major = element_blank(),
    plot.margin = margin(3, 7, 3, 0)
  )

stamp_ugly(perc_plot2)
```

```{r price-increase-old, eval = FALSE}
axis_col <- "grey50"

perc_plot2 <- ggplot(perc_increase, aes(x = ticker, y = perc)) +
  geom_col(fill = "#56B4E9") +
  geom_hline(yintercept = c(100, 200, 300, 400), color = "white", size = 0.75) +
  geom_text(aes(label = label), color = "black", hjust = -0.1, size = 5,
            family = dviz_font_family) +
  scale_y_continuous(name = "percent increase",
                     limits = c(0, 496),
                     expand = c(0, 0)) +
  scale_x_discrete(name = NULL,
                     breaks = c("FB", "GOOG", "MSFT", "AAPL"),
                     labels = c("Facebook", "Alphabet", "Microsoft", "Apple")) +
  coord_flip() +
  theme_dviz_vgrid(rel_small = 1) +
  theme(
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_line(color = axis_col),
    axis.line.x = element_line(color = axis_col),
    axis.line.y = element_line(color = axis_col),
    axis.text.x = element_text(color = axis_col),
    axis.text.y = element_text(color = "black"),
    axis.title = element_text(color = axis_col),
    panel.grid.major = element_blank(),
    plot.margin = margin(3, 7, 3, 0)
  )

stamp_ugly(perc_plot2)
```

Background grids along both axis directions can make sense, however, for scatter plots where there is no primary axis of interest. This presentation frequently looks best without axis lines. Figure \@ref(fig:Aus-athletes-grid) provides an example.

**Mention additive vs. subtractive mental model.**


For figures where the relevant comparison is the *x* = *y* line, I prefer to draw a diagonal line rather than a grid. **update following text:** For example, consider Figure \@ref(fig:echave-et-al), which compares two sets of correlations for 209 protein structures. By drawing the diagonal line, we can see immediately which correlations are systematically stronger. The same observation is much harder to make when the figure has a background grid instead (Figure \@ref(fig:echave-et-al-grid)). Thus, even though Figure \@ref(fig:echave-et-al-grid) looks pleasing, I label it as bad. (*In particular, mention that the deviation of 10A from expeced is not visible in the second figure.*)

(ref:gene-expression) Gene expression levels in a mutant bacteriophage T7 relative to wild-type. Gene expression levels are measured by mRNA abundances, in transcripts per million (TPM). Each dot corresponds to one gene. In the mutant bacteriophage T7, the promoter in front of gene *9* was deleted, and this resulted in reduced mRNA abundances of gene *9* as well as the neighboring genes *8* and *10A* (highlighted). Data source: @Paffetal2018

```{r gene-expression, fig.width = 5, fig.asp = 1, fig.cap = '(ref:gene-expression)'}
df_segment <- data.frame(
  x = c(3.2e-4, 2e-4), 
  xend = c(3.2e-1, 2e-4),
  y = c(2e-4, 3.2e-4),
  yend = c(2e-4, 3.2e-1)
)

gene_expression %>% 
  filter(strain == "phi9v2" & background == "wt") %>%
  ggplot(aes(x=tpm_wt, y=tpm_mutant)) +
  geom_abline(slope = 1, color = "grey") +
  geom_point(pch = 21, fill = "#0072B2D0", color = "white", size = 3) +
  geom_text_repel(
    aes(label = label),
    nudge_y = -0.005,
    fontface = "italic",
    family = "Myriad Pro",
    box.padding = 0.35,
    size = 12/.pt,
    segment.color = NA) +
  geom_segment(
    data = df_segment,
    aes(x = x, xend = xend, y = y, yend = yend),
    size = 0.5, inherit.aes = FALSE) +
  scale_x_log10(
    limits = c(2e-4, 3.3e-1),
    name = "wild-type mRNA abundance (TPM)",
    breaks = c(3.2e-4, 1e-3, 3.2e-3, 1e-2, 3.2e-2, 1e-1, 3.2e-1),
    labels = expression("", 10^-3, "", 10^-2, "", 10^-1, "")
  ) + 
  scale_y_log10(
    limits = c(2e-4, 3.3e-1),
    name = "mutant mRNA abundance (TPM)",
    breaks = c(3.2e-4, 1e-3, 3.2e-3, 1e-2, 3.2e-2, 1e-1, 3.2e-1),
    labels = expression("", 10^-3, "", 10^-2, "", 10^-1, "")
  ) + 
  coord_fixed(expand = FALSE, clip = "off") +
  theme_dviz_open() +
  theme(
    axis.line = element_blank(),
    plot.margin = margin(14, 3.5, 3.5, 0)
  ) -> gene_expression_plot

gene_expression_plot
```


(ref:gene-expression-bad) Gene expression levels in a mutant bacteriophage T7 relative to wild-type. By plotting this dataset against a background grid, **explain what happens**

```{r gene-expression-bad, fig.width = 5, fig.asp = 1, fig.cap = '(ref:gene-expression-bad)'}
gene_expression %>% 
  filter(strain == "phi9v2" & background == "wt") %>%
  ggplot(aes(x=tpm_wt, y=tpm_mutant)) +
  geom_point(pch = 21, fill = "#0072B2D0", color = "white", size = 3) +
  geom_text_repel(
    aes(label = label),
    nudge_y = -0.005,
    fontface = "italic",
    family = "Myriad Pro",
    box.padding = 0.35,
    size = 12/.pt,
    segment.color = NA) +
  scale_x_log10(
    limits = c(2e-4, 3.3e-1),
    name = "wild-type mRNA abundance (TPM)",
    breaks = c(3.2e-4, 1e-3, 3.2e-3, 1e-2, 3.2e-2, 1e-1, 3.2e-1),
    labels = expression("", 10^-3, "", 10^-2, "", 10^-1, "")
  ) + 
  scale_y_log10(
    limits = c(2e-4, 3.3e-1),
    name = "mutant mRNA abundance (TPM)",
    breaks = c(3.2e-4, 1e-3, 3.2e-3, 1e-2, 3.2e-2, 1e-1, 3.2e-1),
    labels = expression("", 10^-3, "", 10^-2, "", 10^-1, "")
  ) + 
  coord_fixed(expand = FALSE, clip = "off") +
  theme_dviz_open() +
  theme(
    panel.grid.major = element_line(colour = "grey80", size = 0.5, linetype = 1),
    plot.margin = margin(14, 3.5, 3.5, 0)
  ) -> gene_expression_plot_bad


stamp_bad(gene_expression_plot_bad)
```


In summary, there is no simple choice of background grid that always works. I encourage you to think carefully about which specific grid or guidelines are most informative for the plot you are making, and to only show those. In general, less is more. Too many or overly thick and dark grid lines can distract your reader's attention away from the data you want to show.

