```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")

library(readr)
library(lubridate)
library(ggrepel)
```

# Maximizing the data signal in visualizations {#maximize-data-signal}


With the rising popularity of the R package ggplot2, which uses a gray background grid as default, graphs with this style have become widespread. With apologies to ggplot2 author Hadley Wickham, for whom I have the utmost respect, I don't find this style particularly attractive. In general, I find that the gray background detracts from the actual data. As an example, consider Figure \@ref(fig:price-plot-ggplot-default), which shows the stock price of four major tech companies, indexed to their value in June 2012. The grid is too busy, and the gray background in the legend is distracting.

(ref:price-plot-ggplot-default) Stock price over time for four major tech companies. The stock price for each company has been normalized to equal 100 in June 2012. 

```{r price-plot-ggplot-default, fig.cap='(ref:price-plot-ggplot-default)'}
price_plot <- ggplot(tech_stocks, aes(x=date, y=price_indexed, color=ticker)) +
  geom_line() +
  scale_color_manual(values=c("#000000", "#E69F00", "#56B4E9", "#009E73"),
                     name="",
                     breaks=c("FB", "GOOG", "MSFT", "AAPL"),
                     labels=c("Facebook", "Alphabet", "Microsoft", "Apple")) +
  scale_x_date(name="year",
               limits=c(ymd("2012-06-01"), ymd("2017-05-31")),
               expand=c(0,0)) + 
  scale_y_continuous(name="stock price, indexed",
                     limits = c(0, 560),
                     expand=c(0,0)) +
  theme_dviz_grid() +
  theme(panel.background = element_rect(fill = "grey90"),
        panel.grid.major = element_line(color = "white"),
        panel.grid.minor = element_line(color = "white", size = rel(0.5)),
        axis.ticks = element_line(color = "black"),
        plot.margin = margin(7, 7, 3, 0))

stamp_ugly(price_plot)
```


We could try to remove the grid altogether, but I think that is a worse option (Figure \@ref(fig:price-plot-no-grid)). Now the curves seem to just float in space, and it's difficult to see where they go. In addition, since all prices are indexed to 100 in June 2012, at a minimum this value should be marked in the plot. Thus, one option would be to add a thin horizontal line at *y* = 100 (Figure \@ref(fig:price-plot-refline)).

(ref:price-plot-no-grid) Indexed stock price over time for four major tech companies.

```{r price-plot-no-grid, fig.cap='(ref:price-plot-no-grid)'}
stamp_bad(price_plot + 
            theme_dviz_open() +
            theme(plot.margin = margin(7, 7, 3, 0)))
```

(ref:price-plot-refline) Indexed stock price over time for four major tech companies.

```{r price-plot-refline, fig.cap='(ref:price-plot-refline)'}
price_plot2 <- ggplot(tech_stocks, aes(x=date, y=price_indexed, color=ticker)) +
  geom_hline(yintercept = 100, size = 0.5, color="grey70") +
  geom_line() +
  scale_color_manual(values=c("#000000", "#E69F00", "#56B4E9", "#009E73"),
                     name="",
                     breaks=c("FB", "GOOG", "MSFT", "AAPL"),
                     labels=c("Facebook", "Alphabet", "Microsoft", "Apple")) +
  scale_x_date(name="year",
               limits=c(ymd("2012-06-01"), ymd("2017-05-31")),
               expand=c(0,0)) + 
  scale_y_continuous(name="stock price, indexed",
                     limits = c(0, 560),
                     expand=c(0,0)) +
  theme_dviz_open() +
  theme(plot.margin = margin(7, 7, 3, 0))

price_plot2
```

Alternatively, we can use just a minimal grid. In particular, for a plot where we are primarily interested in the change in *y* values, vertical grid lines are not needed. Moreover, grid lines positioned at only the major axis ticks will often be sufficient. And, the axis line can be omitted or made very thin (Figure \@ref(fig:price-plot-hgrid)).

(ref:price-plot-hgrid) Indexed stock price over time for four major tech companies.

```{r price-plot-hgrid, fig.cap='(ref:price-plot-hgrid)'}
price_plot + theme_dviz_hgrid() +
  theme(plot.margin = margin(7, 7, 3, 0))
```

For such a minimal grid, we generally draw the lines orthogonally to direction along which the numbers of interest vary. Therefore, if instead of plotting the stock price over time we plot the five-year increase, as horizontal bars, then we will want to use vertical lines instead (Figure \@ref(fig:price-increase)).

(ref:price-increase) Percent increase in stock price from June 2012 to June 2017, for four major tech companies.

```{r price-increase, fig.cap='(ref:price-increase)'}
perc_increase <- filter(tech_stocks, date == ymd("2017-06-01")) %>%
  mutate(perc=100*(price-index_price)/index_price,
         label=paste(as.character(round(perc)), "%", sep="")) %>%
  arrange(perc)

perc_increase$ticker <- factor(perc_increase$ticker, levels=perc_increase$ticker)

perc_plot <- ggplot(perc_increase, aes(x=ticker, y=perc)) +
  geom_col(fill="#56B4E9") +
  geom_text(aes(label=label), color="white", hjust=1.1, size=5,
            family = dviz_font_family) +
  scale_y_continuous(#name="percent increase\n(June 2012 to June 2017)",
                     name="percent increase",
                     limits=c(0, 499),
                     expand=c(0, 0)) +
  scale_x_discrete(name = NULL,
                     breaks = c("FB", "GOOG", "MSFT", "AAPL"),
                     labels = c("Facebook ", "Alphabet ", "Microsoft ", "Apple ")) +
  coord_flip() +
  theme_dviz_vgrid(rel_small = 1) +
  theme(axis.ticks = element_blank(),
        axis.ticks.length = grid::unit(0, "pt"))

perc_plot
```


```{block type='rmdtip', echo=TRUE}
Grid lines that run perpendicular to the key variable of interest tend to be the most useful.
```

(ref:price-increase-bad) **Figure caption here.** The white grid lines on top of the bars make it look like the bars fall apart.

```{r price-increase-bad, fig.cap='(ref:price-increase-bad)'}
perc_plot2 <- ggplot(perc_increase, aes(x = ticker, y = perc)) +
  geom_col(fill = "#56B4E9") +
  geom_hline(yintercept = c(100, 200, 300, 400), color = "white", size = 0.75) +
  geom_text(aes(label = label), color = "black", hjust = -0.1, size = 5,
            family = dviz_font_family) +
  scale_y_continuous(name = "percent increase",
                     limits = c(0, 496),
                     expand = c(0, 0)) +
  scale_x_discrete(name = NULL,
                     breaks = c("FB", "GOOG", "MSFT", "AAPL"),
                     labels = c("Facebook", "Alphabet", "Microsoft", "Apple")) +
  coord_flip() +
  theme_dviz_vgrid(rel_small = 1) +
  theme(axis.ticks.y = element_blank(),
        axis.ticks.x = element_line(color = "black"),
        axis.line.x = element_line(color = "black"),
        axis.line.y = element_line(color = "black"),
        panel.grid.major = element_blank(),
        plot.margin = margin(3, 7, 3, 0)
        )

stamp_ugly(perc_plot2)
```

Background grids along both axis directions can make sense, however, for scatter plots where there is no primary axis of interest. This presentation frequently looks best without axis lines. Figure \@ref(fig:Aus-athletes-grid) provides an example.

(ref:Aus-athletes-grid) Percent body fat versus height in professional male Australian athletes. (Each point represents one athlete.) 

```{r Aus-athletes-grid, fig.cap='(ref:Aus-athletes-grid)'}
male_Aus <- filter(Aus_athletes, sex=="m") %>%
    filter(sport %in% c("basketball", "field", "swimming", "track (400m)",
                        "track (sprint)", "water polo")) %>%
    mutate(sport = case_when(sport == "track (400m)" ~ "track",
                             sport == "track (sprint)" ~ "track",
                             TRUE ~ sport))

male_Aus$sport <- factor(male_Aus$sport,
                         levels = c("field", "water polo", "basketball", "swimming", "track"))

ggplot(male_Aus, aes(x=height, y=pcBfat, color=sport, fill=sport, shape=sport)) +
    geom_point(size = 2.5) +
    scale_shape_manual(values = 21:25) +
    scale_color_OkabeIto(order=c(2, 1, 3, 4, 5), darken = 0.3) +
    scale_fill_OkabeIto(order=c(2, 1, 3, 4, 5), darken = 0.1, alpha = 0.7) +
    ylab("% body fat") + theme_dviz_grid()
```

For figures where the relevant comparison is the *x* = *y* line, I prefer to draw a diagonal line rather than a grid. **update following text:** For example, consider Figure \@ref(fig:echave-et-al), which compares two sets of correlations for 209 protein structures. By drawing the diagonal line, we can see immediately which correlations are systematically stronger. The same observation is much harder to make when the figure has a background grid instead (Figure \@ref(fig:echave-et-al-grid)). Thus, even though Figure \@ref(fig:echave-et-al-grid) looks pleasing, I label it as bad. (*In particular, mention that the deviation of 10A from expeced is not visible in the second figure.*)

(ref:gene-expression) Gene expression levels in a mutant bacteriophage T7 relative to wild-type. Gene expression levels are measured by mRNA abundances, in transcripts per million (TPM). Each dot corresponds to one gene. In the mutant bacteriophage T7, the promoter in front of gene *9* was deleted, and this resulted in reduced mRNA abundances of gene *9* as well as the neighboring genes *8* and *10A* (highlighted). Data source: **reference needed. Paff et al. 2018.**

```{r gene-expression, fig.width = 5, fig.asp = 1, fig.cap = '(ref:gene-expression)'}
df_segment <- data.frame(
  x = c(3.2e-4, 2e-4), 
  xend = c(3.2e-1, 2e-4),
  y = c(2e-4, 3.2e-4),
  yend = c(2e-4, 3.2e-1)
)

gene_expression %>% 
  filter(strain == "phi9v2" & background == "wt") %>%
  ggplot(aes(x=tpm_wt, y=tpm_mutant)) +
  geom_abline(slope = 1, color = "grey") +
  geom_point(pch = 21, fill = "#0072B2D0", color = "white", size = 3) +
  geom_text_repel(
    aes(label = label),
    nudge_y = -0.005,
    fontface = "italic",
    family = "Myriad Pro",
    box.padding = 0.35,
    size = 12/.pt,
    segment.color = NA) +
  geom_segment(
    data = df_segment,
    aes(x = x, xend = xend, y = y, yend = yend),
    size = 0.5, inherit.aes = FALSE) +
  scale_x_log10(
    limits = c(2e-4, 3.3e-1),
    name = "wild-type mRNA abundance (TPM)",
    breaks = c(3.2e-4, 1e-3, 3.2e-3, 1e-2, 3.2e-2, 1e-1, 3.2e-1),
    labels = expression("", 10^-3, "", 10^-2, "", 10^-1, "")
  ) + 
  scale_y_log10(
    limits = c(2e-4, 3.3e-1),
    name = "mutant mRNA abundance (TPM)",
    breaks = c(3.2e-4, 1e-3, 3.2e-3, 1e-2, 3.2e-2, 1e-1, 3.2e-1),
    labels = expression("", 10^-3, "", 10^-2, "", 10^-1, "")
  ) + 
  coord_fixed(expand = FALSE, clip = "off") +
  theme_dviz_open() +
  theme(
    axis.line = element_blank(),
    plot.margin = margin(14, 3.5, 3.5, 0)
  ) -> gene_expression_plot

gene_expression_plot
```


(ref:gene-expression-bad) Gene expression levels in a mutant bacteriophage T7 relative to wild-type. By plotting this dataset against a background grid, **explain what happens**

```{r gene-expression-bad, fig.width = 5, fig.asp = 1, fig.cap = '(ref:gene-expression-bad)'}
gene_expression %>% 
  filter(strain == "phi9v2" & background == "wt") %>%
  ggplot(aes(x=tpm_wt, y=tpm_mutant)) +
  geom_point(pch = 21, fill = "#0072B2D0", color = "white", size = 3) +
  geom_text_repel(
    aes(label = label),
    nudge_y = -0.005,
    fontface = "italic",
    family = "Myriad Pro",
    box.padding = 0.35,
    size = 12/.pt,
    segment.color = NA) +
  scale_x_log10(
    limits = c(2e-4, 3.3e-1),
    name = "wild-type mRNA abundance (TPM)",
    breaks = c(3.2e-4, 1e-3, 3.2e-3, 1e-2, 3.2e-2, 1e-1, 3.2e-1),
    labels = expression("", 10^-3, "", 10^-2, "", 10^-1, "")
  ) + 
  scale_y_log10(
    limits = c(2e-4, 3.3e-1),
    name = "mutant mRNA abundance (TPM)",
    breaks = c(3.2e-4, 1e-3, 3.2e-3, 1e-2, 3.2e-2, 1e-1, 3.2e-1),
    labels = expression("", 10^-3, "", 10^-2, "", 10^-1, "")
  ) + 
  coord_fixed(expand = FALSE, clip = "off") +
  theme_dviz_open() +
  theme(
    panel.grid.major = element_line(colour = "grey80", size = 0.5, linetype = 1),
    plot.margin = margin(14, 3.5, 3.5, 0)
  ) -> gene_expression_plot_bad


stamp_bad(gene_expression_plot_bad)
```


In summary, there is no simple choice of background grid that always works. I encourage you to think carefully about which specific grid or guidelines are most informative for the plot you are making, and to only show those. In general, less is more. Too many or overly thick and dark grid lines can distract your reader's attention away from the data you want to show.


## Boxes and frames

If we overdo removing boxes, frames, grid lines, etc. then we may end up with a figure where elements seem to float in space and are not connected to each other.

**This chapter probably has to come after multi-part figures if we want to discuss figures such as the next one.**

```{r fig.width = 5, fig.asp = 3/4}
titanic %>% mutate(surv = ifelse(survived == 0, "died", "survived")) %>%
  ggplot(aes(sex, fill = sex)) + geom_bar() +
    facet_grid(class ~ surv, scales = "free_x") +
    scale_x_discrete(name = NULL) + 
    scale_y_continuous(limits = c(0, 195), expand = c(0, 0)) +
    scale_fill_manual(values = c("#D55E00D0", "#0072B2D0"), guide = "none") +
    theme_dviz_open(rel_small = 1) +
    theme(axis.line.x = element_blank(),
          axis.ticks.x = element_blank(),
          #axis.text.x = element_text(margin = margin(7, 0, 0, 0)),
          strip.text = element_text(margin = margin(3.5, 3.5, 3.5, 3.5)),
          strip.background  = element_blank(),
          plot.margin = margin(3.5, 14, 3.5, 0)) -> p

stamp_bad(p)
```

```{r fig.width = 5, fig.asp = 3/4}
colors <- desaturate(lighten(c("#56B4E9", "#56B4E9"), c(.8, -.4)), .8)

titanic %>% mutate(surv = ifelse(survived == 0, "died", "survived")) %>%
  ggplot(aes(sex, fill = sex)) + geom_bar() +
    facet_grid(class ~ surv, scales = "free_x") +
    scale_x_discrete(name = NULL) + 
    scale_y_continuous(limits = c(0, 195), expand = c(0, 0)) +
    scale_fill_manual(values = c("#D55E00D0", "#0072B2D0"), guide = "none") +
    theme_dviz_open(rel_small = 1) +
    theme(axis.line = element_blank(),
          #axis.line.y = element_line(color = colors[2]),
          axis.ticks = element_blank(),
          axis.ticks.length = grid::unit(0, "pt"),
          #axis.ticks.y = element_line(color = colors[2]),
          axis.text.x = element_text(margin = margin(7, 0, 3.5, 0)),
          axis.text.y = element_text(color = colors[2]),
          strip.text = element_text(margin = margin(3.5, 3.5, 3.5, 3.5)),
          strip.background  = element_blank(),
          panel.background = element_rect(fill = colors[1]),
          panel.grid.major.y = element_line(color = "white"),
          plot.margin = margin(3.5, 14, 3.5, 0)) -> p

p
```
