```{r echo = FALSE, message = FALSE, warning = FALSE}
# run setup script
source("_common.R")

library(stringr)
library(ggridges)
```

# Multi-part figures {#multi-part-figures}


## Trellis or faceted plots

The idea of Trellis plots was developed in the 1990 by Becker, Cleveland, and colleagues at Bell Labs [@Becker-Cleveland-Shyu-1996].

(ref:titanic-passenger-breakdown) Breakdown of passengers on the Titanic by gender, survival, and class in which they traveled (1st, 2nd, or 3rd).

```{r titanic-passenger-breakdown, fig.cap = '(ref:titanic-passenger-breakdown)'}
titanic %>% mutate(surv = ifelse(survived == 0, "died", "survived")) %>%
  ggplot(aes(sex, fill = sex)) + geom_bar() +
    facet_grid(class ~ surv, scales = "free_x") +
    scale_x_discrete(name = NULL) + 
    scale_y_continuous(limits = c(0, 195), expand = c(0, 0)) +
    scale_fill_manual(values = c("#D55E00D0", "#0072B2D0"), guide = "none") +
    theme_dviz_hgrid(rel_small = 1) +
    theme(axis.line = element_blank(),
          axis.ticks.length = grid::unit(0, "pt"),
          axis.ticks = element_blank(),
          axis.text.x = element_text(margin = margin(7, 0, 0, 0)),
          strip.text = element_text(margin = margin(3.5, 3.5, 3.5, 3.5)),
          strip.background  = element_rect(fill = "grey80", colour = "grey80",
                                            linetype = 1, size = 0.25),
          panel.border = element_rect(colour = "grey80", fill = NA, linetype = 1,
                                      size = 1.))
```

(ref:movie-rankings) Average movie rankings versus number of votes, for movies from 1906 to 2005. Dots represent individual movies, and lines represent the linear regression of the average ranking of each movie versus the logarithm of the number of votes the movie has received. In most years, movies with a higher number of votes have, on average, a higher average ranking. However, this trend has weakend towards the end of the 20th century, and a negative relationship can be seen in the early 2000s. Data Source: Internet Movie Database (IMDB, http://imdb.com/)

```{r movie-rankings, fig.width = 8.5, fig.asp = 1, fig.cap = '(ref:movie-rankings)'}
library(ggplot2movies)

ggplot(filter(movies, year > 1905), aes(y = rating, x = votes)) + 
  geom_point(color = "#0072B250", size = 0.1) + 
  geom_smooth(method = 'lm', se = FALSE, size = 1.25, color = '#D55E00',
              fullrange = TRUE) + 
  scale_x_log10(labels = label_log10, name = "number of votes") + 
  scale_y_continuous(limits = c(0, 10), expand = c(0, 0),
                     breaks = c(0, 5, 10), name = "average rating") + 
  facet_wrap(~year, ncol = 10) +
  theme_dviz_grid(10, rel_small = 1, line_size = 0.25) +
  theme(strip.text = element_text(margin = margin(3.5, 3.5, 3.5, 3.5)),
        panel.border = element_rect(colour = "grey80", fill = NA, linetype = 1,
                                    size = 1.))
```


(ref:BA-degrees-variable-y-lims) Trends in Bachelor's Degrees conferred by U.S. institutions of higher learning. Shown are all degree areas that represent, on average, more than 4% of all degrees. This figure is labeled as bad because all panels use different *y*-axis ranges, and this choice obscures the relative sizes of the different degree areas. Data Source: National Center for Education Statistics

```{r BA-degrees-variable-y-lims, fig.width = 8.5, fig.asp = 0.8, fig.cap = '(ref:BA-degrees-variable-y-lims)'}
BA_degrees %>% 
  mutate(field = ifelse(field == "Communication, journalism, and related programs",
                        "Communication, journalism, and related", field)) -> BA_df

BA_df %>% group_by(field) %>%
  summarize(mean_perc = mean(perc)) %>%
  arrange(desc(mean_perc)) -> BA_top

top_fields <- filter(BA_top, mean_perc>0.04)$field

BA_top_degrees <- filter(BA_df, field %in% top_fields) %>%
  mutate(field = factor(field, levels = top_fields)) %>%
  arrange(field)

p <- ggplot(BA_top_degrees, aes(year, perc)) + 
  geom_line(color = "#0072B2") + 
  facet_wrap(~field, labeller = label_wrap_gen(width = 25), ncol = 3,
             scales = "free") +
  ylab("percent") +
  theme_dviz_hgrid() +
  theme(strip.text = element_text(margin = margin(7, 7, 3, 7)),
        panel.spacing.x = grid::unit(14, "pt"),
        plot.margin = margin(3.5, 14, 3.5, 0)) 

stamp_bad(p)
```

(ref:BA-degrees-fixed-y-lims) Trends in Bachelor's Degrees conferred by U.S. institutions of higher learning. Shown are all degree areas that represent, on average, more than 4% of all degrees. Data Source: National Center for Education Statistics

```{r BA-degrees-fixed-y-lims, fig.width = 8.5, fig.asp = 0.8, fig.cap = '(ref:BA-degrees-fixed-y-lims)'}
ggplot(BA_top_degrees, aes(year, perc)) + 
  geom_line(color = "#0072B2") + 
  facet_wrap(~field, labeller = label_wrap_gen(width = 25), ncol = 3,
             scales = "free") +
  scale_y_continuous(limits = c(0, 0.241), expand = c(0, 0),
                     name = "percent") +
  theme_dviz_hgrid() +
  theme(strip.text = element_text(margin = margin(7, 7, 3, 7)),
        panel.spacing.x = grid::unit(14, "pt"),
        plot.margin = margin(3.5, 14, 3.5, 0)) 

```


(ref:BA-degrees-compound) Trends in Bachelor's Degrees conferred by U.S. institutions of higher learning. (a) From 1970 to 2015, the total number of degrees nearly doubled. (b) Among the most popular degree areas, social sciences, history, and education experienced a major decline, while business and health professions grew. Data Source: National Center for Education Statistics

```{r BA-degrees-compound, fig.asp = 0.4, fig.width = 8.5, fig.cap = '(ref:BA-degrees-compound)'}
BA_degrees %>% 
  mutate(field = ifelse(field == "Communication, journalism, and related programs",
                        "Communication, journalism, and related", field)) -> BA_df

BA_df %>% group_by(year) %>%
  summarize(total = sum(count)) -> BA_totals

textcol <- "gray30"

p1 <- ggplot(BA_totals, aes(year, total/1e6)) + 
  geom_density_line(stat = "identity", color = "#0072B2",
                    fill = desaturate(lighten("#0072B280", .3), .4)) + 
  scale_y_continuous(limits = c(0, 2.05), expand = c(0, 0),
                     name = "degrees awarded (millions)") +
  scale_x_continuous(limits = c(1970, 2016), expand = c(0, 0), name = NULL) +
  theme_dviz_hgrid() +
  theme(axis.title = element_text(color = textcol),
        axis.text = element_text(color = textcol),
        plot.margin = margin(3, 7, 3, 0))

BA_df %>% group_by(field) %>%
  summarize(mean_perc = mean(perc)) %>%
  arrange(desc(mean_perc)) -> BA_top

top_fields <- filter(BA_top, mean_perc>0.055)$field

BA_top_pairs <- filter(BA_df, field %in% top_fields,
                       year %in% c(1971, 2015)) %>%
  mutate(field_wrapped = str_wrap(field, 25))

p2 <- ggplot(BA_top_pairs, aes(x = year, y = perc)) +
  geom_line(aes(group = field), color = "gray60") +
  geom_point(fill = "#0072B2", color = "white", shape = 21, size = 3, stroke = 1.5) +
  scale_x_continuous(limits = c(1971, 2015), breaks = c(1971, 2015),
                     labels = c("1970-71", "2014-15"),
                     expand = expand_scale(mult = c(0.1, 0.04)),
                     name = NULL,
                     position = "top") +
  scale_y_continuous(limits = c(0.02, 0.22), expand = c(0, 0), name = "percent of degrees",
                     sec.axis = dup_axis(breaks = filter(BA_top_pairs, year == 2015)$perc,
                                         labels = filter(BA_top_pairs, year == 2015)$field_wrapped,
                                         name = NULL)) +
  theme_dviz_open() +
  theme(axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_text(color = textcol),
        axis.text.y = element_text(color = textcol),
        axis.line.y.left = element_line(color = textcol),
        axis.text.y.right = element_text(hjust = 0, vjust = .5,
                                         margin = margin(0, 0, 0, 0),
                                         #family = "Myriad Pro Condensed",
                                         color = "black",
                                         lineheight = 0.8
                                         ),
        axis.line.y.right = element_blank(),
        axis.ticks.y.right = element_blank(),
        plot.margin = margin(3, 0, 3, 0))

plot_grid(p1, p2, labels = "auto", rel_widths = c(1.2, 1), align = 'h')
```

(ref:Aus-athletes-composite) Physiology and body-composition of male and female athletes. (a) The data set encompasses 73 female and 85 male professional athletes. (b) Male athletes tend to have higher red blood cell counts (RBC count, reported in units of $10^{12}$ per liter) than female athletes, but there are no such differences for white blood cell counts (WBC count, reported in units of $10^{9}$ per liter). (c) Male athletes tend to have a lower body fat percentage than female athletes performing in the same sport. Data source: @Telford-Cunningham-1991

```{r Aus-athletes-composite, fig.asp = 0.75, fig.cap = '(ref:Aus-athletes-composite)'}
male_sport <- unique(filter(Aus_athletes, sex=="m")$sport)
female_sport <- unique(filter(Aus_athletes, sex=="f")$sport)
both_sport <- male_sport[male_sport %in% female_sport]
athletes_df <- filter(Aus_athletes, sport %in% both_sport) %>%
  mutate(sport = case_when(sport == "track (400m)" ~ "track",
                           sport == "track (sprint)" ~ "track",
                           TRUE ~ sport),
         sex = factor(sex, levels = c("f", "m")))

p1 <- ggplot(athletes_df, aes(x = sex, fill = sex)) + 
  geom_bar() +
  scale_y_continuous(limits = c(0, 95), expand = c(0, 0), name = "number") +
  scale_x_discrete(name = NULL, labels = c("female", "male")) +
  scale_fill_manual(values = c("#D55E00D0", "#0072B2D0"), guide = "none") +
  theme_dviz_hgrid(12, rel_small = 1) + 
  theme(#axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.length = grid::unit(0, "pt"),
        plot.margin = margin(3, 6, 0, 0))

p2 <- ggplot(athletes_df, aes(x = rcc, y = wcc, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2.5) +
  scale_x_continuous(limits = c(3.8, 6.75), name = NULL) +
  scale_y_continuous(limits = c(2.2, 11.), expand = c(0, 0), name = "WBC count") +
  scale_fill_manual(values = c("#D55E00D0", "#0072B2D0"), guide = "none") +
  theme_dviz_hgrid(12, rel_small = 1) +
  theme(plot.margin = margin(3, 0, 0, 0))

p_row <- plot_grid(p1, p2, labels = "auto", align = 'h', rel_widths = c(0.7, 1)) +
  draw_text("RBC count", x = 1, y = 0, size = 12, hjust = 1, vjust = -0.02,
            family = dviz_font_family) + 
  theme(plot.margin = margin(0, 0, 12, 0))

GeomBP <- GeomBoxplot
GeomBP$draw_key <- draw_key_polygon

p3 <- ggplot(athletes_df, aes(x = sport, y = pcBfat, color = sex, fill = sex)) + 
  stat_boxplot(width = 0.5, geom = GeomBP) +
  scale_color_manual(values = c("#D55E00", "#0072B2"), name = NULL,
                     labels = c("female   ", "male")) +
  scale_fill_manual(values = c("#D55E0040", "#0072B240"), guide = "none") +
  scale_x_discrete(name = NULL) +
  scale_y_continuous(name = "% body fat") +
  guides(color = guide_legend(override.aes = list(fill = c("#D55E00D0", "#0072B2D0"),
                                                  color = "white", size = 2),
                              direction = "horizontal")) +
  theme_dviz_hgrid(12, rel_small = 1) +
  theme(axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.length = grid::unit(0, "pt"),
        legend.position = c(1., 0.9),
        legend.justification = "right")

plot_grid(p_row, p3, ncol = 1, labels = c("", "c")) +
            theme(plot.margin = margin(6, 12, 0, 0))
```
