```{r echo = FALSE, message = FALSE, warning = FALSE}
# run setup script
source("_common.R")

library(ggforce)
library(treemapify)
```

# Visualizing nested proportions {#nested-proportions}

(ref:bridges-pie-wrong)  Breakdown of bridges in Pittsburgh by construction material (steel, wood, iron) and by date of construction (crafts, before 1870, and modern, after 1940), shown as a pie chart. This figure is invalid, because the percentages add up to more than 100%. There is overlap between construction material and date of construction. For example, all modern bridges are made of steel, and the majority of crafts bridges are made of wood.


```{r bridges-pie-wrong, fig.width = 4.5, fig.asp = 0.8, fig.cap = '(ref:bridges-pie-wrong)'}

# crafts: before 1870
# emerging: 1870 -- 1889
# mature: 1890 -- 1939
# modern: after 1940

select(bridges, MATERIAL, ERECTED) %>%
  table() %>% 
  reshape2::melt() %>%
  rename(material = MATERIAL, erected = ERECTED, count = value) %>%
  mutate(
    material = case_when(
      material == "IRON" ~ "iron",
      material == "STEEL" ~ "steel",
      material == "WOOD" ~ "wood"
    ),
    erected = case_when(
      erected == "CRAFTS" ~ "crafts",
      erected == "EMERGING" ~ "emerging",
      erected == "MATURE" ~ "mature",
      erected == "MODERN" ~ "modern"
   )
  ) %>%
  group_by(erected) %>%
  mutate(group_count = sum(count)) -> bridges_tidy


n_total <- sum(bridges_tidy$count)

bridges_erected <- filter(bridges_tidy, erected %in% c("crafts", "modern")) %>%
  group_by(erected) %>%
  summarize(
    count = sum(count),
    percent = round(100*count/n_total, 1)
  ) %>%
  rename(type = erected)

bridges_material <- group_by(bridges_tidy, material) %>%
  summarize(
    count = sum(count),
    percent = round(100*count/n_total, 1)
  ) %>%
  rename(type = material)

bridges_material_erected <- rbind(bridges_material, bridges_erected) %>%
  mutate(
    type = factor(type, levels = c("steel", "wood", "iron", "modern", "crafts"))
  ) %>%
  arrange(type)

bridges_pie <- bridges_material_erected %>%
  mutate(
    count_total = sum(count),
    end_angle = 2*pi*cumsum(count)/count_total,   # ending angle for each pie slice
    start_angle = lag(end_angle, default = 0),   # starting angle for each pie slice
    mid_angle = 0.5*(start_angle + end_angle),   # middle of each pie slice, for the text label
    hjust = ifelse(mid_angle>pi, 1, 0),
    vjust = ifelse(mid_angle<pi/2 | mid_angle>3*pi/2, 0, 1)
  )

rpie = 1
rlabel = 1.05 * rpie

p_bridges_pie <- ggplot(bridges_pie) + 
  geom_arc_bar(
    aes(
      x0 = 0, y0 = 0, r0 = 0, r = rpie,
      start = start_angle, end = end_angle, fill = type
    ),
    color = "white", size = 0.5
  ) +
  geom_text(
    aes(
      x = rlabel*sin(mid_angle),
      y = rlabel*cos(mid_angle),
      label = type,
      hjust = hjust, vjust = vjust
    ),
    family = dviz_font_family,
    size = 14/.pt
  ) +
  geom_text(
    aes(
      x = 0.6*sin(mid_angle),
      y = 0.6*cos(mid_angle),
      label = paste0(percent, "%")
    ),
    family = dviz_font_family,
    size = 12/.pt,
    color = c("white", "white", "white", "black", "black")
  ) +
  coord_fixed(clip = "off") +
  scale_x_continuous(
    limits = c(-1.5, 1.5), expand = c(0, 0), name = "", breaks = NULL, labels = NULL
  ) +
  scale_y_continuous(
    limits = c(-1.15, 1.15), expand = c(0, 0), name = "", breaks = NULL, labels = NULL
  ) +
  scale_fill_manual(
    values = c(iron = "#D55E00D0", wood = "#009E73D0", steel = "#0072B2D0",
               crafts = "#F0E442D0", modern = "#56B4E9D0")
  ) +
  theme_dviz_map() +
  theme(legend.position = "none")

stamp_wrong(p_bridges_pie)
```


(ref:bridges-bars-bad) Breakdown of bridges in Pittsburgh by construction material (steel, wood, iron) and by date of construction (crafts, before 1870, and modern, after 1940), shown as a bar plot. Unlike Figure \@ref(fig:bridges-pie-wrong), this visualization is not technically wrong, since it doesn't imply that the bar heights need to add up to 100%. However, it also does not clearly indicate the overlap among different groups, and therefore I have labeled it "bad".


```{r bridges-bars-bad, fig.cap = '(ref:bridges-bars-bad)'}
p_bridges_bars <- ggplot(bridges_material_erected) +
  aes(type, percent, fill = type) +
  geom_col() +
  scale_y_continuous(
    limits = c(0, 75),
    expand = c(0, 0),
    labels = function(x) paste0(x, "%"),
    name = "proportion of bridges"
  ) +
  scale_x_discrete(name = NULL) +
  scale_fill_manual(
    values = c(iron = "#D55E00D0", wood = "#009E73D0", steel = "#0072B2D0",
               crafts = "#F0E442D0", modern = "#56B4E9D0")
  ) +
  coord_cartesian(clip = "off") +
  theme_dviz_hgrid() +
  theme(
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none",
    plot.margin = margin(3.5, 7, 3.5, 0)
  )

stamp_bad(p_bridges_bars)
```

(ref:bridges-mosaic) **Mosaic plot** Note that unlike in a stacked bar plot, in a mosaic plot both the height and the width of individual shaded areas vary. The 

```{r bridges-mosaic, fig.cap = '(ref:bridges-mosaic)'}
select(bridges, MATERIAL, ERECTED) %>%
  table() %>% 
  reshape2::melt() %>%
  rename(material = MATERIAL, erected = ERECTED, count = value) %>%
  mutate(
    material = case_when(
      material == "IRON" ~ "iron",
      material == "STEEL" ~ "steel",
      material == "WOOD" ~ "wood"
    ),
    erected = case_when(
      erected == "CRAFTS" ~ "crafts",
      erected == "EMERGING" ~ "emerging",
      erected == "MATURE" ~ "mature",
      erected == "MODERN" ~ "modern"
   )
  ) %>%
  group_by(erected) %>%
  mutate(group_count = sum(count)) -> bridges_tidy

labels_df <- group_by(bridges_tidy, erected) %>%
  filter(count != 0) %>%
  arrange(desc(material)) %>%
  mutate(
    y = (cumsum(count) - 0.5*count)/group_count,
    y = ifelse(
      erected == "mature" & material == "wood", NA, y
    )
  )

ggplot(bridges_tidy) +
  aes(x = erected, y = count, width = group_count, fill = material) +
  geom_bar(stat = "identity", position = "fill", colour = "white", size = 1) +
  geom_text(
    data = labels_df,
    aes(y = y, label = count, color = material),
    na.rm = TRUE,
    size = 12/.pt,
    family = dviz_font_family
  ) +
  facet_grid(~erected, scales = "free_x", space = "free_x") +
  scale_y_continuous(
    name = NULL,
    #breaks = NULL,
    expand = c(0, 0),
    breaks = filter(labels_df, erected == "crafts")$y,
    labels = filter(labels_df, erected == "crafts")$material,
    sec.axis = dup_axis(
      breaks = filter(labels_df, erected == "modern")$y,
      labels = filter(labels_df, erected == "modern")$material
    )
  ) +
  scale_x_discrete(
    name = NULL
  ) +
  scale_fill_manual(
    values = c("#D55E00D0", "#0072B2D0", "#009E73D0"), 
    guide = "none"
  ) +
  scale_color_manual(
    values = c(iron = "white", wood = "white", steel = "white"),
    guide = "none"
  ) +
  coord_cartesian(clip = "off") +
  theme_dviz_grid(rel_small = 1) +
  theme(
    line = element_blank(),
    strip.text = element_blank(),
    axis.ticks.length = unit(0, "pt"),
    panel.spacing.x = unit(0, "pt")
  )
```

(ref:bridges-treemap) **Treemap**


```{r bridges-treemap, fig.asp = 3/4, fig.cap = '(ref:bridges-treemap)'}
filcols <- c("#D55E00D0", "#0072B2D0", "#009E73D0")
filcols <- c(vapply(filcols, function(x) c(lighten(x, .9), lighten(x, .6), lighten(x, .3), x), character(4)))

ggplot(bridges_tidy, aes(area = count, subgroup = material, fill = interaction(erected, material))) +
  geom_treemap(color = "white", size = 0.5*.pt, alpha = NA) + 
  geom_treemap_subgroup_text(
    family = dviz_font_family,
    colour = "grey50",
    place = "centre", alpha = 0.7,
    grow = TRUE
  ) +
  geom_treemap_subgroup_border(color = "white") +
  geom_treemap_text(
    aes(label = erected, color = interaction(erected, material)),
    family = dviz_font_family,
    place = "centre",
    grow = FALSE
  ) +
  scale_fill_manual(values = filcols) +
  scale_color_manual(values = c(
    crafts.iron = "black", crafts.steel = "black", crafts.wood = "black",
    emerging.iron = "black", emerging.steel = "black", emerging.wood = "black",
    mature.iron = "black", mature.steel = "black", mature.wood = "black",
    modern.iron = "white", modern.steel = "white", modern.wood = "white")
  ) +
  coord_cartesian(clip = "off") +
  guides(colour = "none", fill = "none")

```


```{r fig.width = 8.5}
population_df <- left_join(US_census, US_regions) %>%
    group_by(region, division, state) %>%
    summarize(
      pop2000 = sum(pop2000, na.rm = TRUE),
      pop2010 = sum(pop2010, na.rm = TRUE),
      area = sum(area)
    ) %>%
    ungroup() %>%
    mutate(
      state = factor(state, levels = state),
      region = factor(region, levels = c("West", "South", "Midwest", "Northeast"))
    )

## manually add colors
# hues
hues <- c(300, 50, 250, 100) # purple, brown, blue, green
hues <- c(50, 100, 250, 300) # brown, green, blue, purple

# minimum and maximum population density
minpop <- min(population_df$pop2010)
maxpop <- max(population_df$pop2010)

# turn pop density into color
population_df_color <- population_df %>%
  mutate(index = as.numeric(factor(region))) %>%
  group_by(index) %>%
  mutate(
    value = (pop2010-minpop)/(maxpop-minpop),
    fill = scales::gradient_n_pal(
      colorspace::sequential_hcl(
        6,
        h = hues[index[1]],
        c = c(45, 20),
        l = c(30, 80),
        power = .5
      )
    )(1-value)
  )

ggplot(population_df_color, aes(area = area, subgroup = region, fill = fill)) +
  geom_treemap(color = "white", size = 0.5*.pt, alpha = NA) + 
  geom_treemap_subgroup_text(
    family = dviz_font_family,
    colour = "white",
    place = "centre", alpha = 0.7,
    grow = TRUE
  ) +
  geom_treemap_subgroup_border(color = "white") +
  geom_treemap_text(
    aes(label = state),
    color = "black",
    family = dviz_font_family,
    place = "centre",
    grow = FALSE
  ) +
  scale_fill_identity() +
  coord_cartesian(clip = "off") +
  guides(colour = "none", fill = "none")
```

(ref:bridges-nested-pie) *Figure caption...*

```{r bridges-nested-pie, fig.width = 5, fig.asp = 0.8, fig.cap = '(ref:bridges-nested-pie)'}
bridges_arranged <-
  ungroup(bridges_tidy) %>%
  mutate(material = factor(material, levels = c("wood", "iron", "steel"))) %>%
  arrange(material)

bridges_pie_outer <- bridges_arranged %>%
  mutate(
    count_total = sum(count),
    end_angle = 2*pi*cumsum(count)/count_total,   # ending angle for each pie slice
    start_angle = lag(end_angle, default = 0),   # starting angle for each pie slice
    mid_angle = 0.5*(start_angle + end_angle),   # middle of each pie slice, for the text label
    hjust = ifelse(mid_angle>pi, 1, 0),
    vjust = ifelse(mid_angle<pi/2 | mid_angle>3*pi/2, 0, 1),
    type = erected
  ) %>%
  filter(erected %in% c("crafts", "modern"), count != 0)

bridges_pie_inner <- bridges_arranged %>%
  group_by(material) %>%
  summarize(count = sum(count)) %>%
  mutate(
    count_total = sum(count),
    end_angle = 2*pi*cumsum(count)/count_total,   # ending angle for each pie slice
    start_angle = lag(end_angle, default = 0),   # starting angle for each pie slice
    mid_angle = 0.5*(start_angle + end_angle),   # middle of each pie slice, for the text label
    hjust = ifelse(mid_angle>pi, 1, 0),
    vjust = ifelse(mid_angle<pi/2 | mid_angle>3*pi/2, 0, 1),
    type = material
  ) 

rpie <- 1
rpie1 <- 0.6
rpie2 <- 1
rlabel <- 1.05 * rpie

bridges_nested_pie <- ggplot() + 
  geom_arc_bar(data = bridges_pie_outer,
    aes(
      x0 = 0, y0 = 0, r0 = rpie1, r = rpie2,
      start = start_angle, end = end_angle, fill = type
    ),
    color = "white", size = 0.5
  ) +
  geom_arc_bar(data = bridges_pie_inner,
    aes(
      x0 = 0, y0 = 0, r0 = 0, r = rpie1,
      start = start_angle, end = end_angle, fill = type
    ),
    color = "white", size = 0.5
  ) +
  geom_text(data = bridges_pie_outer,
    aes(
      x = rlabel*sin(mid_angle),
      y = rlabel*cos(mid_angle),
      label = type,
      hjust = hjust, vjust = vjust
    ),
    family = dviz_font_family,
    size = 14/.pt
  ) +
  geom_text(data = bridges_pie_outer,
    aes(
      x = 0.8*sin(mid_angle),
      y = 0.8*cos(mid_angle),
      label = count
    ),
    family = dviz_font_family,
    size = 12/.pt
  ) +
  geom_text(data = bridges_pie_inner,
    aes(
      x = 0.3*sin(mid_angle),
      y = 0.3*cos(mid_angle),
      label = count
    ),
    family = dviz_font_family,
    size = 12/.pt
  ) +
  coord_fixed(clip = "off") +
  scale_x_continuous(
    limits = c(-1.5, 1.5), expand = c(0, 0), name = "", breaks = NULL, labels = NULL
  ) +
  scale_y_continuous(
    limits = c(-1.15, 1.15), expand = c(0, 0), name = "", breaks = NULL, labels = NULL
  ) +
#  scale_fill_manual(
#    values = c(`1st` = "#009E73", `2nd` = "#E69F00", `3rd` = "#F0E442",
#               men = "#0072B2", women = "#D55E00")
#  ) +
  theme_dviz_map() +
  theme(legend.position = "none")

bridges_nested_pie
```

*Rules:*

- Color by a variable that lies on the end 



```{r}
select(bridges, MATERIAL, ERECTED, RIVER, LENGTH) %>%
  table() %>% 
  reshape2::melt() %>%
  rename(material = MATERIAL, erected = ERECTED, count = value) %>%
  mutate(
    material = case_when(
      material == "IRON" ~ "iron",
      material == "STEEL" ~ "steel",
      material == "WOOD" ~ "wood"
    ),
    erected = case_when(
      erected == "CRAFTS" ~ "crafts",
      erected == "EMERGING" ~ "emerging",
      erected == "MATURE" ~ "mature",
      erected == "MODERN" ~ "modern"
   )
  ) -> data

data <- gather_set_data(data, 1:4)
data$x <- factor(data$x, levels = c("material", "erected", "LENGTH", "RIVER"))

ggplot(data, aes(x, id = id, split = y, value = count)) +
  geom_parallel_sets(aes(fill = material), alpha = 0.5, axis.width = 0.13) +
  geom_parallel_sets_axes(axis.width = 0.13, fill = "grey70", color = "grey70") +
  geom_parallel_sets_labels(
    color = 'black',
    family = dviz_font_family,
    size = 10/.pt,
    angle = 90
  ) +
  scale_x_discrete(
    name = NULL,
    expand = c(0, 0.3)
  ) +
  scale_y_continuous(breaks = NULL, expand = c(0, 0))+
  scale_fill_manual(
    values = c(iron = "#D55E00D0", wood = "#009E73D0", steel = "#0072B2D0"),
    guide = "none"
  ) +
  theme_dviz_open() +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_blank()
  )

```

```{r}
ggplot(data, aes(x, id = id, split = y, value = count)) +
  geom_parallel_sets(aes(fill = RIVER), alpha = 0.5, axis.width = 0.13) +
  geom_parallel_sets_axes(axis.width = 0.13, fill = "grey70", color = "grey70") +
  geom_parallel_sets_labels(
    color = 'black',
    family = dviz_font_family,
    size = 10/.pt,
    angle = 90
  ) +
  scale_x_discrete(
    name = NULL,
    expand = c(0, 0.3)
  ) +
  scale_y_continuous(breaks = NULL, expand = c(0, 0))+
#  scale_fill_manual(
#    values = c(iron = "#D55E00D0", wood = "#009E73D0", steel = "#0072B2D0"),
#    guide = "none"
#  ) +
  theme_dviz_open() +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_blank()
  )
```

By changing the order of the groups and the variable used for coloring, we can highlight different aspects of the data. While Figure \@ref(fig:titanic-parallel-sets1) tells a story about who survived and who did not, Figure \@ref(fig:titanic-parallel-sets2) tells a story about crew and passengers traveling in different classes.

