```{r echo = FALSE, message = FALSE, warning = FALSE}
# run setup script
source("_common.R")

library(ggforce)
library(treemapify)
```

# Visualizing nested proportions {#nested-proportions}

*We need to note somewhere that the dataset used in this chapter is different from the other Titanic dataset.*

(ref:titanic-wrong-pie) Breakdown of passenger on the Titanic by class (1st, 2nd, or 3rd) and by gender (women or men, excluding children under the age of 16), shown as a pie chart. This figure is invalid, because the percentages add up to more than 100%. There is overlap between passenger classes and passenger gender, as every female and male passenger travelled in either 1st, 2nd, or 3rd class.

```{r titanic-wrong-pie, fig.width = 4.5, fig.asp = 0.8, fig.cap = '(ref:titanic-wrong-pie)'}
titanic_tidy <- reshape2::melt(Titanic)
titanic_tidy_no_crew <- filter(titanic_tidy, Class != "Crew")

n_passengers <- summarize(
  titanic_tidy_no_crew,
  count = sum(value)
)$count

titanic_sex <- filter(titanic_tidy_no_crew, Age != "Child") %>%
  group_by(Sex) %>%
  summarize(
    count = sum(value),
    percent = round(100*count/n_passengers, 1)
  ) %>%
  rename(type = Sex) %>%
  mutate(
    type = case_when(
      type == "Female" ~ "women",
      type == "Male" ~ "men"
    )
  )

titanic_class <- group_by(titanic_tidy_no_crew, Class) %>%
  summarize(
    count = sum(value),
    percent = round(100*count/n_passengers, 1)
  ) %>%
  rename(type = Class) %>%
  mutate(
    type = paste0(type, " class")
  )

titanic_sex_class <- rbind(titanic_sex, titanic_class)

titanic_pie <- titanic_sex_class %>%
  mutate(
    count_total = sum(count),
    end_angle = 2*pi*cumsum(count)/count_total,   # ending angle for each pie slice
    start_angle = lag(end_angle, default = 0),   # starting angle for each pie slice
    mid_angle = 0.5*(start_angle + end_angle),   # middle of each pie slice, for the text label
    hjust = ifelse(mid_angle>pi, 1, 0),
    vjust = ifelse(mid_angle<pi/2 | mid_angle>3*pi/2, 0, 1)
  )

rpie = 1
rlabel = 1.05 * rpie

p_titanic_pie <- ggplot(titanic_pie) + 
  geom_arc_bar(
    aes(
      x0 = 0, y0 = 0, r0 = 0, r = rpie,
      start = start_angle, end = end_angle, fill = type
    ),
    color = "white", size = 0.5
  ) +
  geom_text(
    aes(
      x = rlabel*sin(mid_angle),
      y = rlabel*cos(mid_angle),
      label = type,
      hjust = hjust, vjust = vjust
    ),
    family = dviz_font_family,
    size = 14/.pt
  ) +
  geom_text(
    aes(
      x = 0.6*sin(mid_angle),
      y = 0.6*cos(mid_angle),
      label = paste0(percent, "%")
    ),
    family = dviz_font_family,
    size = 12/.pt,
    color = c("white", "black", "black", "black", "black")
  ) +
  coord_fixed(clip = "off") +
  scale_x_continuous(
    limits = c(-1.5, 1.5), expand = c(0, 0), name = "", breaks = NULL, labels = NULL
  ) +
  scale_y_continuous(
    limits = c(-1.15, 1.15), expand = c(0, 0), name = "", breaks = NULL, labels = NULL
  ) +
  scale_fill_manual(
    values = c(`1st class` = "#009E73", `2nd class` = "#E69F00", `3rd class` = "#F0E442",
               men = "#0072B2", women = "#D55E00")
  ) +
  theme_dviz_map() +
  theme(legend.position = "none")

stamp_wrong(p_titanic_pie)
```


(ref:titanic-bad-bars) Breakdown of passenger on the Titanic by class (1st, 2nd, or 3rd) and by gender (women or men, excluding children under the age of 16), shown as a bar plot. Unlike Figure \@ref(fig:titanic-wrong-pie), this visualization is not technically wrong, since it doesn't imply that the bar heights need to add up to 100%. However, it also does not clearly indicate the overlap among different groups, and therefore I have labeled it "bad".


```{r titanic-bad-bars, fig.cap = '(ref:titanic-bad-bars2)'}
p_titanic_bars <- ggplot(titanic_sex_class) +
  aes(type, percent, fill = type) +
  geom_col() +
  scale_y_continuous(
    limits = c(0, 65),
    expand = c(0, 0),
    labels = function(x) paste0(x, "%"),
    name = "proportion of passengers"
  ) +
  scale_x_discrete(name = NULL) +
  scale_fill_manual(
    values = c(`1st class` = "#009E73", `2nd class` = "#E69F00", `3rd class` = "#F0E442",
               men = "#0072B2", women = "#D55E00")
  ) +
  coord_cartesian(clip = "off") +
  theme_dviz_hgrid() +
  theme(
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none",
    plot.margin = margin(3.5, 7, 3.5, 0)
  )

stamp_bad(p_titanic_bars)
```

(ref:titanic-nested-pie) *Figure caption...*

```{r titanic-nested-pie, fig.width = 5, fig.asp = 0.8, fig.cap = '(ref:titanic-nested-pie)'}
titanic_nested <- rename(titanic_tidy_no_crew, class = Class) %>%
  mutate(
    gender = factor(case_when(
      Sex == "Male" & Age == "Adult" ~ "men",
      Sex == "Female" & Age == "Adult" ~ "women",
      TRUE ~ "children"
    ), levels = c("men", "women", "children"))
  ) %>%
  group_by(class, gender) %>%
  summarize(
    gender_count = sum(value),
    gender_percent = round(100*gender_count/n_passengers, 1)
  ) %>%
  ungroup() %>%
  left_join(
    rename(titanic_tidy_no_crew, class = Class) %>%
      group_by(class) %>%
      summarize(
        count = sum(value),
        percent = round(100*count/n_passengers, 1)
      )
  )


titanic_pie_outer <- titanic_nested %>%
  mutate(
    count_total = sum(gender_count),
    end_angle = 2*pi*cumsum(gender_count)/count_total,   # ending angle for each pie slice
    start_angle = lag(end_angle, default = 0),   # starting angle for each pie slice
    mid_angle = 0.5*(start_angle + end_angle),   # middle of each pie slice, for the text label
    hjust = ifelse(mid_angle>pi, 1, 0),
    vjust = ifelse(mid_angle<pi/2 | mid_angle>3*pi/2, 0, 1),
    type = gender
  ) %>%
  filter(gender != "children")

titanic_pie_inner <- titanic_nested %>%
  filter(gender == "men") %>%
  mutate(
    count_total = sum(count),
    end_angle = 2*pi*cumsum(count)/count_total,   # ending angle for each pie slice
    start_angle = lag(end_angle, default = 0),   # starting angle for each pie slice
    mid_angle = 0.5*(start_angle + end_angle),   # middle of each pie slice, for the text label
    hjust = ifelse(mid_angle>pi, 1, 0),
    vjust = ifelse(mid_angle<pi/2 | mid_angle>3*pi/2, 0, 1),
    type = class
  ) %>%
  select(-gender, -gender_count, -gender_percent)

rpie1 = 0.6
rpie2 = 1
rlabel = 1.05 * rpie

p_titanic_pie <- ggplot() + 
  geom_arc_bar(data = titanic_pie_outer,
    aes(
      x0 = 0, y0 = 0, r0 = rpie1, r = rpie2,
      start = start_angle, end = end_angle, fill = type
    ),
    color = "white", size = 0.5
  ) +
  geom_arc_bar(data = titanic_pie_inner,
    aes(
      x0 = 0, y0 = 0, r0 = 0, r = rpie1,
      start = start_angle, end = end_angle, fill = type
    ),
    color = "white", size = 0.5
  ) +
  geom_text(data = titanic_pie_outer,
    aes(
      x = rlabel*sin(mid_angle),
      y = rlabel*cos(mid_angle),
      label = gender,
      hjust = hjust, vjust = vjust
    ),
    family = dviz_font_family,
    size = 14/.pt
  ) +
  geom_text(data = titanic_pie_outer,
    aes(
      x = 0.8*sin(mid_angle),
      y = 0.8*cos(mid_angle),
      label = gender_count
    ),
    family = dviz_font_family,
    size = 12/.pt
  ) +
  geom_text(data = titanic_pie_inner,
    aes(
      x = 0.3*sin(mid_angle),
      y = 0.3*cos(mid_angle),
      label = count
    ),
    family = dviz_font_family,
    size = 12/.pt
  ) +
  coord_fixed(clip = "off") +
  scale_x_continuous(
    limits = c(-1.5, 1.5), expand = c(0, 0), name = "", breaks = NULL, labels = NULL
  ) +
  scale_y_continuous(
    limits = c(-1.15, 1.15), expand = c(0, 0), name = "", breaks = NULL, labels = NULL
  ) +
  scale_fill_manual(
    values = c(`1st` = "#009E73", `2nd` = "#E69F00", `3rd` = "#F0E442",
               men = "#0072B2", women = "#D55E00")
  ) +
  theme_dviz_map() +
  theme(legend.position = "none")

p_titanic_pie
```

```{r tianic_treemap, fig.asp = 3/4}
filcols <- c("#66c2a5", "#fdae61", "#abdda4")
filcols <- c(vapply(filcols, function(x) c(lighten(x, .2), x, lighten(x, -.1)), character(3)))

ggplot(titanic_nested, aes(area = gender_count, subgroup = class, fill = interaction(class, gender))) +
  geom_treemap(color = "white", size = 0.5*.pt) + 
  geom_treemap_subgroup_text(
    family = dviz_font_family,
    colour = "grey30", place = "centre", alpha = 0.7,
    grow = TRUE
  ) +
  geom_treemap_subgroup_border(color = "white") +
  geom_treemap_text(
    aes(label = gender),
    family = dviz_font_family,
    colour = "black", place = "centre",
    grow = FALSE
  ) +
  scale_fill_manual(values = filcols) +
  coord_cartesian(clip = "off") +
  guides(colour = "none", fill = "none")


```

```{r}
titanic_nested$gender <- factor(titanic_nested$gender, levels = c("children", "women", "men"))

labels_df <- group_by(titanic_nested, class) %>%
  mutate(
    y = (cumsum(gender_count) - 0.5*gender_count)/count,
    y = ifelse(
      class == "1st" & gender == "children",
      (cumsum(gender_count) - 0*gender_count)/count,
      y
    )
  )

ggplot(titanic_nested) +
  aes(x = class, y = gender_count, width = sqrt(count), fill = gender) +
  geom_bar(stat = "identity", position = "fill", colour = "white", size = 1) +
  geom_text(
    data = labels_df,
    aes(y = y, label = gender_count)
  ) +
  facet_grid(~class, scales = "free_x", space = "free_x") +
  scale_y_continuous(
    name = NULL,
    breaks = NULL,
    expand = expand_scale(mult = c(0, 0.02)),
    sec.axis = dup_axis(
      breaks = filter(labels_df, class == "3rd")$y,
      labels = filter(labels_df, class == "3rd")$gender
    )
  ) +
  scale_x_discrete(
    breaks = c("1st", "2nd", "3rd"),
    labels = c("1st class", "2nd class", "3rd class"),
    name = NULL
  ) +
  scale_fill_manual(
    values = c("#abdda4", "#fdae61", "#66c2a5"), 
    guide = "none"
  ) +
  theme_dviz_grid(rel_small = 1) +
  theme(
    line = element_blank(),
    strip.text = element_blank(),
    panel.spacing.x = unit(0, "pt"),
    axis.ticks.length = unit(0, "pt")
  )
```

(ref:titanic-parallel-sets1) *Figure caption.*

```{r titanic-parallel-sets1, fig.cap = '(ref:titanic-parallel-sets1)'}
data <- reshape2::melt(Titanic)
data <- gather_set_data(data, 1:4)
data$x <- factor(data$x, levels = c("Sex", "Age", "Class", "Survived"))

ggplot(data, aes(x, id = id, split = y, value = value)) +
  geom_parallel_sets(aes(fill = Survived), alpha = 0.5, axis.width = 0.13) +
  geom_parallel_sets_axes(axis.width = 0.13, fill = "grey70", color = "grey70") +
  geom_parallel_sets_labels(
    color = 'black',
    family = dviz_font_family,
    size = 10/.pt,
    angle = 90
  ) +
  scale_x_discrete(
    name = NULL,
    expand = c(0, 0.3)
  ) +
  scale_y_continuous(breaks = NULL, expand = c(0, 0))+
  scale_fill_manual(
    values = c("#56B4E9", "#E69F00"),
    guide = "none"
  ) +
  theme_dviz_open() +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_blank()
  )
```

By changing the order of the groups and the variable used for coloring, we can highlight different aspects of the data. While Figure \@ref(fig:titanic-parallel-sets1) tells a story about who survived and who did not, Figure \@ref(fig:titanic-parallel-sets2) tells a story about crew and passengers traveling in different classes.


(ref:titanic-parallel-sets2) *Figure caption.*

```{r titanic-parallel-sets2, fig.cap = '(ref:titanic-parallel-sets2)'}
data$x <- factor(data$x, levels = c("Class", "Age", "Sex", "Survived"))

ggplot(data, aes(x, id = id, split = y, value = value)) +
  geom_parallel_sets(aes(fill = Class), alpha = 0.5, axis.width = 0.13) +
  geom_parallel_sets_axes(axis.width = 0.13, fill = "grey70", color = "grey70") +
  geom_parallel_sets_labels(
    color = 'black',
    family = dviz_font_family,
    size = 10/.pt,
    angle = 90
  ) +
  scale_x_discrete(
    name = NULL,
    expand = c(0, 0.3)
  ) +
  scale_y_continuous(breaks = NULL, expand = c(0, 0))+
  scale_fill_manual(
    values = c(`1st` = "#009E73", `2nd` = "#E69F00", `3rd` = "#56B4E9", Crew = "#F0E442"),
    guide = "none"
  ) +
  theme_dviz_open() +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_blank()
  )
```

Add mosaic plot? https://stackoverflow.com/a/45688044/4975218
