```{r echo = FALSE, message = FALSE, warning = FALSE}
# run setup script
source("_common.R")
library(tidyverse)
library(lubridate)
library(patchwork)
library(ggforce)
library(ggrepel)
library(ggridges)
library(ggthemes)
library(ggmap)
library(gridExtra)
library(treemapify)
library(hexbin)
library(magick)
library(plot3D)
library(geofacet)
library(ggspatial)
library(rgeos)
library(maptools)
library(sf)
library(statebins)
library(maps)
library(tidybayes)
library(broom)
library(mgcv)
library(emmeans)
library(ungeviz)
library(gganimate)
library(RColorBrewer)
library(gapminder)
library(nycflights13)
```

# Technical notes {-}

The entire book was written in R Markdown, using the **bookdown**, **rmarkdown**, and **knitr** packages. All figures were made with **ggplot2**, with the help of add-on packages **cowplot**, **egg**, **ggforce**, **ggrepel**, **ggridges**, **sf**, and **treemapify**. Color manipulations were done with the **colorspace** and **colorblindr** packages. For many of these packages, the current development version is required to compile all parts of the book.

The source code for the book is available here: https://github.com/clauswilke/dataviz. The book also requires a supporting R package, **dviz.supp**, whose code is available here: https://github.com/clauswilke/dviz.supp.

The book was last compiled using the following environment:
```{r}
sessionInfo()
```
