```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")
library(forcats)
```


# Visualizing amounts

```{r}
# source: Box Office Mojo
# URL: http://www.boxofficemojo.com/weekend/chart/?view=&yr=2017&wknd=51&p=.htm
# downloaded: 2018-02-11

boxoffice <- data.frame(rank = 1:5,
                        title = c("Star Wars: The Last Jedi", "Jumanji: Welcome to the Jungle", "Pitch Perfect 3", "The Greatest Showman", "Ferdinand"),
                        title_short = c("Star Wars", "Jumanji", "Pitch Perfect 3", "Greatest Showman", "Ferdinand"),
                        amount = c(71565498, 36169328, 19928525, 8805843, 7316746),
                        amount_text = c("$71,565,498", "$36,169,328", "$19,928,525", "$8,805,843", "$7,316,746"))

boxoffice_display <- boxoffice %>%
  mutate(A = " ", B = " ", C = " ") %>%
  select(A, rank, title, amount_text, B, C) %>%
  rename(` ` = A,
         Rank = rank,
         Title = title,
         `Weekend gross` = amount_text,
         ` ` = B,
         ` ` = C)

knitr::kable(
  boxoffice_display,
  caption = 'Highest grossing movies for the weekend of December 22-24, 2017. Data source: Box Office Mojo. Used with permission', booktabs = TRUE,
  row.names = FALSE,
  align = c('c', 'c', 'l', 'r', 'c', 'c')#,
  #format = "html",
  #table.attr = "style = \"width: 75%\""
)
```

```{r boxoffice-vertical, fig.width = 8, fig.asp = .5}
boxoffice %>%
  ggplot(aes(x = fct_reorder(title_short, rank), y = amount)) +
    geom_col(fill = "#56B4E9", width = 0.6, alpha = 0.9) +
    scale_y_continuous(expand = c(0, 0),
                       breaks = c(0, 2e7, 4e7, 6e7),
                       labels = c("0", "20", "40", "60"),
                       name = "weekend gross ($1M)") +
    scale_x_discrete(name = NULL,
                     expand = c(0, 0.4)) +
    theme_minimal_hgrid() +
    theme(axis.ticks.length = grid::unit(0, "pt"),
          axis.ticks = element_blank(),
          axis.title = element_text(size = 12))
```

```{r boxoffice-rot-axis-tick-labels, fig.asp = 0.85}
boxoffice %>%
  ggplot(aes(x = fct_reorder(title_short, rank), y = amount)) +
    geom_col(fill = "#56B4E9", alpha = 0.9) +
    scale_y_continuous(expand = c(0, 0),
                       breaks = c(0, 2e7, 4e7, 6e7),
                       labels = c("0", "20", "40", "60"),
                       name = "weekend gross ($1M)") +
    scale_x_discrete(name = NULL) +
    theme_minimal_hgrid() +
    theme(axis.ticks.length = grid::unit(0, "pt"),
          axis.ticks = element_blank(),
          axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
          axis.title = element_text(size = 12),
          plot.margin = margin(7, 14, 7, 7)) -> p_box_axrot

stamp_ugly(p_box_axrot)
```

```{r boxoffice-horizontal}
ggplot(boxoffice, aes(x = fct_reorder(title_short, desc(rank)), y = amount)) +
  geom_col(fill = "#56B4E9", alpha = 0.9) +
  scale_y_continuous(limits = c(0, 7.5e7),
                     expand = c(0, 0),
                     breaks = c(0, 2e7, 4e7, 6e7),
                     labels = c("0", "20", "40", "60"),
                     name = "weekend gross ($1M)") +
  scale_x_discrete(name = NULL,
                   expand = c(0, 0.5)) +
  coord_flip() +
  theme_minimal_vgrid() +
  theme(axis.ticks.length = grid::unit(0, "pt"),
        axis.ticks = element_blank(),
        axis.title = element_text(size = 12))
```

```{r boxoffice-horizontal-alphabetic}
p <- ggplot(boxoffice, aes(x = fct_rev(title_short), y = amount)) +
  geom_col(fill = "#56B4E9", alpha = 0.9) +
  scale_y_continuous(limits = c(0, 7.5e7),
                     expand = c(0, 0),
                     breaks = c(0, 2e7, 4e7, 6e7),
                     labels = c("0", "20", "40", "60"),
                     name = "weekend gross ($1M)") +
  scale_x_discrete(name = NULL,
                   expand = c(0, 0.5)) +
  coord_flip() +
  theme_minimal_vgrid() +
  theme(axis.ticks.length = grid::unit(0, "pt"),
        axis.ticks = element_blank(),
        axis.title = element_text(size = 12))

stamp_bad(p)
```


```{r income-by-age}
income_by_age %>% filter(race == "all") %>%
  ggplot(aes(x = age, y = median_income)) +
    geom_col(fill = "#56B4E9", alpha = 0.9) +
    scale_y_continuous(expand = c(0, 0),
                       name = "median income (USD)",
                       breaks = c(0, 20000, 40000, 60000),
                       labels = c("$0", "$20,000", "$40,000", "$60,000")) +
    xlab("age (years)") +
    theme_minimal_hgrid() +
    theme(axis.ticks.length = grid::unit(0, "pt"),
          axis.ticks = element_blank(),
          axis.line = element_blank(),
          plot.margin = margin(7, 14, 7, 7))
```

```{r income-by-age-sorted}
income_by_age %>% filter(race == "all") %>%
  ggplot(aes(x = fct_reorder(age, desc(median_income)), y = median_income)) +
    geom_col(fill = "#56B4E9", alpha = 0.9) +
    scale_y_continuous(expand = c(0, 0),
                       name = "median income (USD)",
                       breaks = c(0, 20000, 40000, 60000),
                       labels = c("$0", "$20,000", "$40,000", "$60,000")) +
    xlab("age (years)") +
    theme_minimal_hgrid() +
    theme(axis.ticks.length = grid::unit(0, "pt"),
          axis.ticks = element_blank(),
          axis.line = element_blank(),
          plot.margin = margin(7, 14, 7, 7)) -> p_income_sorted

stamp_bad(p_income_sorted)
```


```{r income-by-age-race-dodged, fig.width = 8, fig.asp = 0.5}
income_by_age %>% filter(race %in% c("white", "asian", "black", "hispanic")) %>%
   mutate(race = fct_relevel(race, c("asian", "white", "hispanic", "black")),
          age = fct_recode(age, "≥ 75" = "> 74")) -> income_df

# Take the darkest four colors from 5-class ColorBrewer palette "PuBu"
colors_four = RColorBrewer::brewer.pal(5, "PuBu")[5:2]

ggplot(income_df, aes(x = age, y = median_income, fill = race)) +
  geom_col(position = "dodge", alpha = 0.9) +
  scale_y_continuous(expand = c(0, 0),
                     name = "median income (USD)",
                     breaks = c(0, 20000, 40000, 60000, 80000, 100000),
                     labels = c("$0", "$20,000", "$40,000", "$60,000", "$80,000", "$100,000")) +
  scale_fill_manual(values = colors_four, name = NULL) +
  xlab("age (years)") +
  theme_minimal_hgrid() +
  theme(axis.ticks.length = grid::unit(0, "pt"),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        plot.margin = margin(7, 14, 7, 7)) -> p_income_race_dodged

stamp_ugly(p_income_race_dodged)
```

```{r income-by-race-age-dodged, fig.width = 8, fig.asp = 0.4}
# Take the darkest seven colors from 8-class ColorBrewer palette "PuBu"
colors_seven = RColorBrewer::brewer.pal(8, "PuBu")[2:8]

ggplot(income_df, aes(x = race, y = median_income, fill = age)) +
  geom_col(position = "dodge", alpha = 0.9) +
  scale_y_continuous(expand = c(0, 0),
                     name = "median income (USD)",
                     breaks = c(0, 20000, 40000, 60000, 80000, 100000),
                     labels = c("$0", "$20,000", "$40,000", "$60,000", "$80,000", "$100,000")) +
  scale_fill_manual(values = colors_seven, name = "age (yrs)") +
  xlab(label = NULL) +
  theme_minimal_hgrid() +
  theme(axis.ticks.length = grid::unit(0, "pt"),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        legend.title.align = 0.5,
        plot.margin = margin(7, 14, 7, 7)) -> p_income_age_dodged

stamp_phantom(p_income_age_dodged)
```

```{r income-by-age-race-faceted, fig.width = 8.5}
income_df %>%
   mutate(age = fct_recode(age, "15–24" = "15 to 24", "25–34" = "25 to 34", "35–44" = "35 to 44",
                                "45–54" = "45 to 54", "55–64" = "55 to 64", "65–74" = "65 to 74")) -> income_age_abbrev_df

ggplot(income_age_abbrev_df, aes(x = age, y = median_income)) +
  geom_col(fill = "#56B4E9", alpha = 0.9) +
  scale_y_continuous(expand = c(0, 0),
                     name = "median income (USD)",
                     breaks = c(0, 20000, 40000, 60000, 80000, 100000),
                     labels = c("$0", "$20,000", "$40,000", "$60,000", "$80,000", "$100,000")) +
  xlab(label = "age (years)") +
  facet_wrap(~race, scales = "free_x") +
  theme_minimal_hgrid(12) +
  theme(axis.ticks.length = grid::unit(0, "pt"),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        strip.text = element_text(size = 12),
        plot.margin = margin(6, 12, 6, 6)) -> p_income_age_dodged

stamp_phantom(p_income_age_dodged)
```


(ref:titanic-passengers-by-class-sex) Numbers of female and male passengers on the Titanic traveling in 1st, 2nd, and 3rd class.

```{r titanic-passengers-by-class-sex, fig.width = 5.5, fig.cap = '(ref:titanic-passengers-by-class-sex)'}
titanic_groups <- titanic_all %>% filter(class != "*") %>% 
  select(class, sex) %>% 
  group_by(class, sex) %>% 
  tally() %>% arrange(class, desc(sex)) %>%
  mutate(sex = factor(sex, levels = c("female", "male"))) %>%
  group_by(class) %>%
  mutate(nlabel = cumsum(n) - n/2) %>%
  ungroup() %>%
  mutate(class = paste(class, "class"))

ggplot(titanic_groups, aes(x = class, y = n, fill = sex)) +
  geom_col(position = "stack", color = "white", size = 1, width = 1) +
  geom_text(aes(y = nlabel, label = n), color = "white", size = 14/.pt) +
  scale_x_discrete(expand = c(0, 0), name = NULL) +
  scale_y_continuous(expand = c(0, 0), breaks = NULL, name = NULL) +
  scale_fill_manual(values = c("#D55E00", "#0072B2"),
                    breaks = c("female", "male"),
                    labels = c("female passengers   ", "male passengers"),
                    name = NULL) +
  theme_minimal_grid() +
  theme(panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(size = 14),
        #legend.position = c(0.13, 0.6),
        #legend.justification = c(0, 0),
        legend.position = "bottom",
        legend.justification = "center",
        legend.background = element_rect(fill = "white"))
```


I have also added the actual numerical values that each bar represents. Whenever your plot is meant to display only a small number of key values, it makes sense to add the actual numbers to the plot. This substantially increases the amount of information conveyed by your plot without adding much visual noise.

**Mention that class cannot be reordered, because it is an ordered factor and it has its own intrinsic order.**


(ref:Americas-life-expect) Life expectancies of countries in the Americas, for the year 2007. Data source: Gapminder project

```{r Americas-life-expect, fig.width = 7., fig.asp = .8, fig.cap = '(ref:Americas-life-expect)'}
library(gapminder)
gap_asia_2007 <- gapminder %>% filter(year == 2007, continent == "Americas")
ggplot(gap_asia_2007, aes(x = lifeExp, y = fct_reorder(country, lifeExp))) +
  geom_point(color = "#0072B2", size = 2.5) +
  scale_x_continuous(name = "life expectancy (years)",
                     limits = c(59.7, 81.5),
                     expand = c(0, 0)) +
  scale_y_discrete(name = NULL, expand = c(0, 0.5)) +
  theme_minimal_grid() +
  theme(axis.ticks.length = grid::unit(0, "pt"),
        axis.title = element_text(size = 12),
        plot.margin = margin(7, 21, 7, 7))
```

(ref:Americas-life-expect-bad) Life expectancies of countries in the Americas, for the year 2007. **More text here.** Data source: Gapminder project

```{r Americas-life-expect-bad, fig.width = 7., fig.asp = .8, fig.cap = '(ref:Americas-life-expect-bad)'}
p <- ggplot(gap_asia_2007, aes(x = lifeExp, y = country)) +
  geom_point(color = "#0072B2", size = 2.5) +
  scale_x_continuous(name = "life expectancy (years)",
                     limits = c(59.7, 81.5),
                     expand = c(0, 0)) +
  scale_y_discrete(name = NULL, expand = c(0, 0.5)) +
  theme_minimal_grid() +
  theme(axis.ticks.length = grid::unit(0, "pt"),
        axis.title = element_text(size = 12),
        plot.margin = margin(7, 21, 7, 7))

stamp_bad(p)
```
