```{r echo = FALSE, message = FALSE, warning = FALSE}
# run setup script
source("_common.R")

library(dplyr)
library(tidyr)
```

# Visualizing associations among two or more quantitative variables {#visualizing-associations}

Outline:

- Basic x-y scatter plot
- Very dense plots: 2d histograms or contour lines (maybe `faithful`?)
- More than 2 variables: Map to color or size, or make matrix of x-y plots
- Correlation matrix (pottery or bumpus dataset?)
- PCA (pottery dataset?)
- Paired data (diagonal plot, parallel dots plot)

```{r}
ggplot(diamonds, aes(carat, price)) + 
  geom_point(size = 0.1, color = darken("#E9F2D5", .1), alpha = 0.5) + 
  geom_density2d(color = "#2D3184", size = 0.25) +
  scale_x_continuous(limits = c(0, 4.2), expand = c(0, 0.1)) +
  scale_y_log10(limits = c(300, 20000),
                breaks = c(300, 1000, 3000, 10000)) +
  #facet_grid(color~clarity) +
  theme_dviz_grid()
```

```{r}
ggplot(diamonds, aes(carat, price)) + 
  geom_bin2d(bins = 50) +
  scale_x_continuous(limits = c(0, 4.2), expand = c(0, 0.1)) +
  scale_y_log10(limits = c(300, 20000),
                breaks = c(300, 1000, 3000, 10000)) +
  scale_fill_continuous_sequential("Blue-Yellow", rev = TRUE, begin = 0) +
  theme_dviz_grid() +
  theme(legend.position = c(0.9, 0.4),
        legend.justification = c(1, 0.5),
        legend.box.background = element_rect(fill = "white",
                                             color = NA),
        legend.box.margin = margin(3.5, 3.5, 3.5, 3.5))
```



```{r fig.width = 4., fig.asp = 1}
forensic_glass <- MASS::fgl

cm <- cor(select(forensic_glass, -type, -RI, -Si))
df_wide <- as.data.frame(cm)
df_long <- stack(df_wide)
names(df_long) <- c("cor", "var1")
df_long <- cbind(df_long, var2 = rep(rownames(cm), length(rownames(cm))))
clust <- hclust(as.dist(1-cm), method="average") 
levels <- clust$labels[clust$order]
df_long$var1 <- factor(df_long$var1, levels = levels)
df_long$var2 <- factor(df_long$var2, levels = levels)
ggplot(filter(df_long, as.integer(var1) < as.integer(var2)),
       aes(var1, var2, fill=cor)) + 
  geom_tile(color = "white", size = 1) + 
  scale_x_discrete(position = "top", name = NULL, expand = c(0, 0)) +
  scale_y_discrete(name = NULL, expand = c(0, 0)) +
  scale_fill_continuous_carto(
    palette = "Earth", rev = TRUE,
    limits = c(-.5, .5), breaks = c(-.5, 0, .5),
    name = "correlation",
    guide = guide_colorbar(
      direction = "horizontal",
      label.position = "bottom",
      title.position = "top",
      barwidth = grid::unit(140, "pt"),
      barheight = grid::unit(17.5, "pt"))) +
  coord_fixed() +
  theme_dviz_open(rel_small = 1) +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.ticks.length = grid::unit(3, "pt"),
        legend.position = c(.97, .0),
        legend.justification = c(1, 0),
        legend.title.align = 0.5)

```

```{r fig.width = 4., fig.asp = 1}
ggplot(filter(df_long, as.integer(var1) < as.integer(var2)),
       aes(var1, var2, fill=cor, size = abs(cor))) + 
  geom_point(shape = 21, stroke = 0) + 
  scale_x_discrete(position = "top", name = NULL, expand = c(0, 0.5)) +
  scale_y_discrete(name = NULL, expand = c(0, 0.5)) +
  scale_size_area(max_size = 19, limits = c(0, 0.5), guide = "none") +
  scale_fill_continuous_carto(
    palette = "Earth", rev = TRUE,
    limits = c(-.5, .5), breaks = c(-.5, 0, .5),
    name = "correlation",
    guide = guide_colorbar(
      direction = "horizontal",
      label.position = "bottom",
      title.position = "top",
      barwidth = grid::unit(140, "pt"),
      barheight = grid::unit(17.5, "pt"))) +
  coord_fixed() +
  theme_dviz_open(rel_small = 1) +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.ticks.length = grid::unit(3, "pt"),
        legend.position = c(.97, .0),
        legend.justification = c(1, 0),
        legend.title.align = 0.5)

```


```{r}
select(forensic_glass, -type, -RI, -Si) %>%
  scale() %>%
  prcomp() -> pca

pca_data <- data.frame(pca$x, type = forensic_glass$type) %>%
  mutate(type = case_when(
    type == "WinF" ~ "window",
    type == "WinNF" ~ "window",
    type == "Veh" ~ "window",
    type == "Con" ~ "container",
    type == "Tabl" ~ "tableware",
    type == "Head" ~ "headlamp"
  )) %>%
  mutate(type = factor(type, levels = c("headlamp", "tableware", "container", "window")))

colors = darken(c("#D55E00", "#0072B2", "#009E73", "#E69F00"), .3)
fills = c("#D55E0040", "#0072B240", "#009E7340", "#E69F0040")

ggplot(pca_data, aes(x = PC1, y = PC2, color = type, fill = type, shape = type)) + 
  geom_point() +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = fills) +
  scale_shape_manual(values = c(22:24, 21)) +
  scale_x_continuous(name = "PC 1") +
  scale_y_continuous(name = "PC 2") +
  theme_dviz_grid()
```

```{r fig.width = 4.5, fig.asp = 1}
rotation_data <- data.frame(pca$rotation, type = row.names(pca$rotation))

rotation_labels <- data.frame(type = c("Ba", "Al", "Na", "K", "Fe", "Ca", "Mg"),
                              hjust = c(1, 1, 1, 0.5, 0, 0.5, 0),
                              vjust = c(0.5, 0.5, 1, 1, 0.5, 0, 0.5),
                              nudge_x = c(-0.01, -0.01, 0.01, 0, .01, 0, .01),
                              nudge_y = c(0, 0, -.01, -.01, 0, .01, 0))

rotation_data <- left_join(rotation_data, rotation_labels, by = "type")

arrow_style <- arrow(length = unit(0.05, "inches"),
                     type = "closed")
ggplot(rotation_data) + 
  geom_segment(aes(xend = PC1, yend = PC2),
               x = 0, y = 0,
               arrow = arrow_style,
               color = "#0072B2",
               fill = "#0072B2") + 
  geom_text(aes(x = PC1 + nudge_x, y = PC2 + nudge_y, label = type,
                hjust = hjust, vjust = vjust),
            size = 12/.pt, color = darken('#D55E00', .3),
            family = dviz_font_family) + 
  scale_x_continuous(limits = c(-.8, .8), name = "PC 1") + 
  scale_y_continuous(limits = c(-.7, .8), breaks = c(-0.5, 0, .5),
                     name = "PC 2") +
  coord_fixed() +
  theme_dviz_grid()
```
