```{r echo = FALSE, message = FALSE, warning = FALSE}
# run setup script
source("_common.R")

library(dplyr)
library(tidyr)
library(ggrepel)
```

# Visualizing associations among two or more quantitative variables {#visualizing-associations}


(ref:blue-jays-scatter) Head length (measured from the tip of the bill to the back of the head, in mm) versus body mass (in gram), for 123 blue jays. Each dot corresponds to one bird. There is a moderate tendency for heavier birds to have longer heads. Data source: Keith Tarvin, Oberlin College

```{r blue-jays-scatter, fig.width = 5, fig.asp = 3/4, fig.cap='(ref:blue-jays-scatter)'}
ggplot(blue_jays, aes(Mass, Head)) + 
  geom_point(pch = 21, fill = "gray25", color = "white", size = 2.5) +
  scale_x_continuous(name = "body mass (g)") +
  scale_y_continuous(name = "head length (mm)") +
  theme_dviz_grid()
```

(ref:blue-jays-scatter-sex) Head length versus body mass for 123 blue jays. The birds' sex is indicated by color. At the same body mass, male birds tend to have longer heads (and specifically, longer bills) than female birds. Data source: Keith Tarvin, Oberlin College

```{r blue-jays-scatter-sex, fig.width = 5, fig.asp = 3/4, fig.cap='(ref:blue-jays-scatter-sex)'}
ggplot(blue_jays, aes(Mass, Head, fill = KnownSex)) + 
  geom_point(pch = 21, color = "white", size = 2.5) +
  scale_x_continuous(name = "body mass (g)") +
  scale_y_continuous(name = "head length (mm)") +
  scale_fill_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 3)
      )
    ) +
  theme_dviz_grid() +
  theme(
    legend.text = element_text(vjust = 0.6),
    legend.position = c(1, 0.01),
    legend.justification = c(1, 0),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
```


```{r blue-jays-scatter-bubbles, fig.width = 5, fig.asp = 5/4, fig.cap='(ref:blue-jays-scatter-bubbles)'}
blue_jays$sex <- ifelse(blue_jays$KnownSex == "F", "female birds", "male birds")
blue_jays$sex <- factor(blue_jays$sex, levels = c("male birds", "female birds"))
ggplot(blue_jays, aes(Mass, Head, size = Skull, fill = KnownSex)) + 
  geom_point(pch = 21, color = "white") +
  facet_wrap(~sex, ncol = 1, scales = "fixed") +
  scale_x_continuous(name = "body mass (g)") +
  scale_y_continuous(name = "head length (mm)", breaks = c(52, 54, 56, 58, 60)) +
  scale_fill_manual(values = c(F = "#D55E00", M = "#0072B2"), guide = "none") +
  scale_radius(
    name = "skull size (mm)",
    range = c(2, 7),
    limits = c(28, 34),
    breaks = c(28, 30, 32, 34),
    labels = c("28   ", "30   ", "32   ", "34"),
    guide = guide_legend(
      direction = "horizontal",
      title.position = "top",
      title.hjust = 0.5,
      label.position = "right",
      override.aes = list(fill = "gray40")
    )
  ) +
  theme_dviz_grid() +
  theme(
    legend.position = c(1, 0.53),
    legend.justification = c(1, 0),
    legend.spacing.x = unit(2, "pt"),
    legend.spacing.y = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt"),
    strip.text = element_text(size = 14, family = "Myriad Pro Semibold")
  )
```

```{r fig.asp = 3/4}
blue_jays %>% select(BirdID, KnownSex, Head, Mass, Skull) %>%
  gather(var_x, val_x, Head:Skull) %>%
  left_join(select(blue_jays, BirdID, Head, Mass, Skull)) %>%
  gather(var_y, val_y, Head:Skull) -> bj_matrix

labels <- c(
  Head = "head length (mm)",
  Mass = "body mass (g)",
  Skull = "skull size (mm)"
)

ggplot(bj_matrix, aes(val_x, val_y, fill = KnownSex)) + 
  geom_point(pch = 21, color = "white", size = 2, stroke = 0.2) +
  scale_x_continuous(expand = expand_scale(mult = 0.1),
                     breaks = scales::pretty_breaks(4, min.n = 3)) +
  scale_y_continuous(expand = expand_scale(mult = 0.1),
                     breaks = scales::pretty_breaks(4, min.n = 3)) +
  scale_fill_manual(
    values = c(F = "#D55E00D0", M = "#0072B2D0"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 2.5)
      )
    ) +
  labs(x = NULL, y = NULL) +
  facet_grid(
    var_y ~ var_x,
    scales = "free",
    switch = "both",
    labeller = labeller(
      var_x = labels,
      var_y = labels
    )
  ) +
  coord_cartesian(clip = "off") +
  theme_dviz_grid() +
  panel_border(colour = "grey90", size = 0.5) +
  theme(
    strip.placement = "outside",
    strip.text.x = element_text(vjust = 1, margin = margin(0, 0, 0, 0)),
    strip.text.y = element_text(vjust = 0, angle = -90),
    legend.box.spacing = grid::unit(0, "pt"),
    legend.text = element_text(vjust = 0.6),
    legend.position = "top",
    legend.justification = "right",
    legend.spacing.x = unit(2, "pt"),
    legend.key.width = unit(10, "pt"),
    plot.margin = margin(3.5, 1, 3.5, 0)
  )
```
After these two figures, do one with varying point sizes and one with small multiples of various quantities plotted against each other.



```{r eval = FALSE}
# For overlapping points section, show birds as contours (no sex), then with sex (2 contour colours), then diamons as contours (labelled bad), then contours faceted? And then maybe the following plot?

ggplot(diamonds, aes(carat, price)) + 
  geom_point(size = 0.1, color = "#E2E6CA", alpha = 0.5) + 
  geom_density2d(color = "#035F33", size = 0.25) +
  scale_x_continuous(limits = c(0, 4.2), expand = c(0, 0.1)) +
  scale_y_log10(limits = c(300, 20000),
                breaks = c(300, 1000, 3000, 10000)) +
  theme_dviz_grid()

ggplot(diamonds, aes(carat, price)) + 
  geom_bin2d(bins = 50) +
  scale_x_continuous(limits = c(0, 4.2), expand = c(0, 0.1)) +
  scale_y_log10(limits = c(300, 20000),
                breaks = c(300, 1000, 3000, 10000)) +
  scale_fill_continuous_sequential("Green-Yellow", rev = TRUE, begin = 0) +
  theme_dviz_grid() +
  theme(legend.position = c(0.9, 0.4),
        legend.justification = c(1, 0.5),
        legend.box.background = element_rect(fill = "white",
                                             color = NA),
        legend.box.margin = margin(3.5, 3.5, 3.5, 3.5))
```

## Correlations and dimension reduction

```{r correlations, fig.asp = 283/434}
r <- c(.2, .6, .9, -.2, -.6, -.9)
labels_df <- tibble(
  r = factor(r, levels = r),
  label = paste0("r = ", r),
  x = c(-2.3, -2.3, -2.3, 2.3, 2.3, 2.3),
  hjust = c(0, 0, 0, 1, 1, 1),
  y = 2.3
)

ggplot(correlation_examples, aes(x, y)) + 
  geom_point(pch = 21, color = "white", size = 2, fill = "#0072B2") + 
  geom_text(
    data = labels_df,
    aes(label = label, hjust = hjust),
    vjust = 1,
    family = "Myriad Pro",
    size = 12/.pt
  ) +
  facet_wrap(~r) +
  coord_fixed(xlim = c(-2.5, 2.5), ylim = c(-2.5, 2.5), expand = FALSE, clip = "off") +
  theme_dviz_map() +
  theme(
    strip.text = element_blank(),
    panel.spacing = grid::unit(12, "pt"),
    plot.margin = margin(3.5, 1, 3.5, 1)
  ) +
  panel_border(colour = "grey80", size = 0.5)
```

```{r fig.width = 4., fig.asp = 1}
cm <- cor(select(forensic_glass, -type, -RI, -Si))
df_wide <- as.data.frame(cm)
df_long <- stack(df_wide)
names(df_long) <- c("cor", "var1")
df_long <- cbind(df_long, var2 = rep(rownames(cm), length(rownames(cm))))
clust <- hclust(as.dist(1-cm), method="average") 
levels <- clust$labels[clust$order]
df_long$var1 <- factor(df_long$var1, levels = levels)
df_long$var2 <- factor(df_long$var2, levels = levels)
ggplot(filter(df_long, as.integer(var1) < as.integer(var2)),
       aes(var1, var2, fill=cor)) + 
  geom_tile(color = "white", size = 1) + 
  scale_x_discrete(position = "top", name = NULL, expand = c(0, 0)) +
  scale_y_discrete(name = NULL, expand = c(0, 0)) +
  scale_fill_continuous_carto(
    palette = "Earth", rev = TRUE,
    limits = c(-.5, .5), breaks = c(-.5, 0, .5),
    name = "correlation",
    guide = guide_colorbar(
      direction = "horizontal",
      label.position = "bottom",
      title.position = "top",
      barwidth = grid::unit(140, "pt"),
      barheight = grid::unit(17.5, "pt"))) +
  coord_fixed() +
  theme_dviz_open(rel_small = 1) +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.ticks.length = grid::unit(3, "pt"),
        legend.position = c(.97, .0),
        legend.justification = c(1, 0),
        legend.title.align = 0.5)

```

```{r fig.width = 4., fig.asp = 1}
ggplot(filter(df_long, as.integer(var1) < as.integer(var2)),
       aes(var1, var2, fill=cor, size = abs(cor))) + 
  geom_point(shape = 21, stroke = 0) + 
  scale_x_discrete(position = "top", name = NULL, expand = c(0, 0.5)) +
  scale_y_discrete(name = NULL, expand = c(0, 0.5)) +
  scale_size_area(max_size = 19, limits = c(0, 0.5), guide = "none") +
  scale_fill_continuous_carto(
    palette = "Earth", rev = TRUE,
    limits = c(-.5, .5), breaks = c(-.5, 0, .5),
    name = "correlation",
    guide = guide_colorbar(
      direction = "horizontal",
      label.position = "bottom",
      title.position = "top",
      barwidth = grid::unit(140, "pt"),
      barheight = grid::unit(17.5, "pt"))) +
  coord_fixed() +
  theme_dviz_open(rel_small = 1) +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.ticks.length = grid::unit(3, "pt"),
        legend.position = c(.97, .0),
        legend.justification = c(1, 0),
        legend.title.align = 0.5)

```


```{r}
select(forensic_glass, -type, -RI, -Si) %>%
  scale() %>%
  prcomp() -> pca

pca_data <- data.frame(pca$x, type = forensic_glass$type) %>%
  mutate(type = case_when(
    type == "WinF" ~ "window",
    type == "WinNF" ~ "window",
    type == "Veh" ~ "window",
    type == "Con" ~ "container",
    type == "Tabl" ~ "tableware",
    type == "Head" ~ "headlamp"
  )) %>%
  mutate(type = factor(type, levels = c("headlamp", "tableware", "container", "window")))

colors = darken(c("#D55E00", "#0072B2", "#009E73", "#E69F00"), .3)
fills = c("#D55E0040", "#0072B240", "#009E7340", "#E69F0040")

ggplot(pca_data, aes(x = PC1, y = PC2, color = type, fill = type, shape = type)) + 
  geom_point() +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = fills) +
  scale_shape_manual(values = c(22:24, 21)) +
  scale_x_continuous(name = "PC 1") +
  scale_y_continuous(name = "PC 2") +
  theme_dviz_grid()
```

```{r fig.width = 4.5, fig.asp = 1}
rotation_data <- data.frame(pca$rotation, type = row.names(pca$rotation))

rotation_labels <- data.frame(type = c("Ba", "Al", "Na", "K", "Fe", "Ca", "Mg"),
                              hjust = c(1, 1, 1, 0.5, 0, 0.5, 0),
                              vjust = c(0.5, 0.5, 1, 1, 0.5, 0, 0.5),
                              nudge_x = c(-0.01, -0.01, 0.01, 0, .01, 0, .01),
                              nudge_y = c(0, 0, -.01, -.01, 0, .01, 0))

rotation_data <- left_join(rotation_data, rotation_labels, by = "type")

arrow_style <- arrow(length = unit(0.05, "inches"),
                     type = "closed")
ggplot(rotation_data) + 
  geom_segment(aes(xend = PC1, yend = PC2),
               x = 0, y = 0,
               arrow = arrow_style,
               color = "#0072B2",
               fill = "#0072B2") + 
  geom_text(aes(x = PC1 + nudge_x, y = PC2 + nudge_y, label = type,
                hjust = hjust, vjust = vjust),
            size = 12/.pt, color = darken('#D55E00', .3),
            family = dviz_font_family) + 
  scale_x_continuous(limits = c(-.8, .8), name = "PC 1") + 
  scale_y_continuous(limits = c(-.7, .8), breaks = c(-0.5, 0, .5),
                     name = "PC 2") +
  coord_fixed() +
  theme_dviz_grid()
```

## Paired data

Mention: x vs y plot is good for large numbers of observations and to show overall patterns, parallel plot (name?) is good when we have small number of observations and are primarily interested in the specific identity.

Maybe do 2 x vs y plots, because they are so important; do one with direct labeling highlight a specific case.



For remainder of the chapter, use CO2 emission data.
```{r fig.width = 5, fig.asp = 0.9}
CO2_emissions %>% filter(year %in% c(1970, 2010)) %>%
  spread(year, emissions) %>% na.omit() -> emissions_1970_2010

df_segment <- data.frame(
  x = c(0.01, 0.008), 
  xend = c(100, 0.008),
  y = c(0.008, 0.01),
  yend = c(0.008, 100)
)


ggplot(emissions_1970_2010, aes(`1970`, `2010`)) + 
  geom_abline(slope = 1, color = "grey") +
  geom_point(pch = 21, fill = "#0072B2D0", color = "white", size = 2, stroke = 0.3) + 
  geom_segment(
    data = df_segment,
    aes(x = x, xend = xend, y = y, yend = yend),
    size = 0.5,
    inherit.aes = FALSE
  ) +
  scale_x_log10(
    limits = c(0.008, 125),
    breaks = c(0.01, 0.1, 1, 10, 100),
    labels = c("0.01", "0.1", "1.0", "10", "100"),
    name = "1970 CO2 emissions (tons / person)"
  ) + 
  scale_y_log10(
    limits = c(0.008, 125),
    breaks = c(0.01, 0.1, 1, 10, 100),
    labels = c("0.01", "0.1", "1.0", "10", "100"),
    name = "2010 CO2 emissions (tons / person)"
  ) +
  coord_fixed(expand = FALSE, clip = "off") +
  theme_dviz_open() +
  theme(axis.line = element_blank())
```

parallel dots plot:

```{r fig.width = 4.5, fig.asp = 1}
CO2_emissions %>% filter(year %in% c(2000, 2005, 2010)) %>%
  spread(year, emissions) %>% 
  na.omit() %>%
  mutate(diff = `2010` - `2000`) %>%
  arrange(desc(abs(diff))) -> emissions_diff

emissions_diff[1:10,] %>% select(-diff) %>% 
  gather(year, emissions, `2000`:`2010`) %>%
  mutate(year = as.numeric(year)) -> emissions_top_ten

textcol <- "gray30"

labels_df <- filter(emissions_top_ten, year == 2010) %>%
  left_join(
    tibble(
      country = c("Trinidad and Tobago", "Qatar", "United Arab Emirates", "Oman", "Bahrain", "Singapore", "Netherlands Antilles", "Kazakhstan", "Equatorial Guinea", "Kuwait"),
      nudge_y = c(.1, .1, .1, .1, -.3, .1, .7, .1, .1, .1)
    )
  )

ggplot(filter(emissions_top_ten, year != 2005), aes(x = year, y = emissions)) +
  geom_line(aes(group = country), color = "gray60") +
  geom_point(color = "white", size = 4) +
  geom_point(color = "#0072B2", size = 2) +
  geom_text(
    data = labels_df,
    aes(x = year + 0.45, y = emissions + nudge_y, label = country),
    family = "Myriad Pro",
    size = 10/.pt,
    hjust = 0) +
  scale_x_continuous(
    limits = c(2000, 2020), breaks = c(2000, 2010),
    labels = c("2000", "2010"),
    expand = expand_scale(add = c(1, 0)),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(
    limits = c(0, 63.2),
    expand = c(0, 0),
    name = "CO2 emissions (tons / person)"
  ) +
  theme_dviz_open() +
  theme(
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(color = textcol),
    axis.text.y = element_text(color = textcol),
    axis.line.y.left = element_line(color = textcol),
    axis.text.y.right = element_text(
      hjust = 0, vjust = .5,
      margin = margin(0, 0, 0, 0),
      color = "black",
      lineheight = 0.8
    ),
    axis.line.y.right = element_blank(),
    axis.ticks.y.right = element_blank(),
    plot.margin = margin(3.5, 0, 7, 0)
  )

```


```{r fig.width = 4.5, fig.asp = 1}
ggplot(emissions_top_ten, aes(x = year, y = emissions)) +
  geom_line(aes(group = country), color = "gray60") +
  geom_point(color = "white", size = 4) +
  geom_point(color = "#0072B2", size = 2) +
  geom_text(
    data = labels_df,
    aes(x = year + 0.45, y = emissions + nudge_y, label = country),
    family = "Myriad Pro",
    size = 10/.pt,
    hjust = 0) +
  scale_x_continuous(
    limits = c(2000, 2020), breaks = c(2000, 2005, 2010),
    labels = c("2000", "2005", "2010"),
    expand = expand_scale(add = c(1, 0)),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(
    limits = c(0, 63.2),
    expand = c(0, 0),
    name = "CO2 emissions (tons / person)"
  ) +
  coord_cartesian(clip = "off") +
  theme_dviz_open() +
  theme(
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(color = textcol),
    axis.text.y = element_text(color = textcol),
    axis.line.y.left = element_line(color = textcol),
    axis.text.y.right = element_text(
      hjust = 0, vjust = .5,
      margin = margin(0, 0, 0, 0),
      color = "black",
      lineheight = 0.8
    ),
    axis.line.y.right = element_blank(),
    axis.ticks.y.right = element_blank(),
    plot.margin = margin(3.5, 0, 7, 0)
  )

```
