```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")

library(lubridate)
```

# Visualizing trends {#visualizing-trends}



## Smoothing

(ref:dow-jones) Daily closing values of the Dow Jones Industrial Average for the year 2009. Data source: Yahoo! Finance

```{r dow-jones, fig.asp = 0.5, fig.cap = '(ref:dow-jones)'}
dow_jones_industrial %>% filter(date >= ymd("2008/12/31") & date <= ymd("2010/01/10")) %>%
  ggplot(aes(date, close)) + 
  geom_line(color = "grey20", size = .5) + 
  scale_x_date(limits = c(ymd("2008-12-31"), ymd("2010-01-10")), expand = c(0, 0)) +
  xlab(NULL) + ylab("Dow Jones Industrial Average") +
  theme_dviz_grid(12) +
  theme(
    plot.margin = margin(3, 12, 3, 0)
  )
```

(ref:dow-jones-moving-ave) Daily closing values of the Dow Jones Industrial Average for the year 2009. Data source: Yahoo! Finance

```{r dow-jones-moving-ave, fig.asp = 1, fig.cap = '(ref:dow-jones-moving-ave)'}
p1 <- dow_jones_industrial %>% filter(date >= ymd("2008/12/31") & date <= ymd("2010/01/10")) %>%
  mutate(
    close_20d_ave_c = moving_ave(date, close, 20),
    close_50d_ave_c = moving_ave(date, close, 50),
    close_100d_ave_c = moving_ave(date, close, 100)
  ) %>%
  ggplot(aes(date, close)) + 
  geom_line(color = "grey20", size = .25) +
  geom_line(aes(date, close_20d_ave_c, color = "20d"), size = .75, na.rm = TRUE) +
  geom_line(aes(date, close_50d_ave_c, color = "50d"), size = .75, na.rm = TRUE) +
  geom_line(aes(date, close_100d_ave_c, color = "100d"), size = .75, na.rm = TRUE) +
  scale_color_manual(
    values = c(
      `20d` = "#009e73",
      `50d` = "#d55e00",
      `100d` = "#0072b2"
    ),
    breaks = c("20d", "50d", "100d"),
    labels = c("20 day average", "50 day average", "100 day average"),
    name = NULL
  ) + 
  scale_x_date(
    limits = c(ymd("2008-12-31"), ymd("2010-01-10")), expand = c(0, 0),
    labels = NULL
  ) +
  xlab(NULL) + ylab("Dow Jones industrial average") +
  theme_dviz_grid(12) +
  theme(
    plot.margin = margin(3, 12, 3, 0),
    legend.position = c(1, 0),
    legend.justification = c(1, 0),
    legend.box.background = element_rect(fill = "white", color = NA),
    legend.box.margin = margin(6, 12, 0, 12),
    axis.ticks.x = element_blank()
  )

p2 <- dow_jones_industrial %>% filter(date >= ymd("2008/12/31") & date <= ymd("2010/01/10")) %>%
  mutate(
    close_20d_ave = moving_ave(date, close, 20, center = FALSE),
    close_50d_ave = moving_ave(date, close, 50, center = FALSE),
    close_100d_ave = moving_ave(date, close, 100, center = FALSE)
  ) %>%
  ggplot(aes(date, close)) + 
  geom_line(color = "grey20", size = .25) +
  geom_line(aes(date, close_20d_ave, color = "20d"), size = .75, na.rm = TRUE) +
  geom_line(aes(date, close_50d_ave, color = "50d"), size = .75, na.rm = TRUE) +
  geom_line(aes(date, close_100d_ave, color = "100d"), size = .75, na.rm = TRUE) +
  scale_color_manual(
    values = c(
      `20d` = "#009e73",
      `50d` = "#d55e00",
      `100d` = "#0072b2"
    ),
    breaks = c("20d", "50d", "100d"),
    labels = c("20 day average", "50 day average", "100 day average"),
    name = NULL,
    guide = "none"
  ) + 
  scale_x_date(limits = c(ymd("2008-12-31"), ymd("2010-01-10")), expand = c(0, 0)) +
  xlab(NULL) + ylab("Dow Jones Industrial Average") +
  theme_dviz_grid(12) +
  theme(
    plot.margin = margin(3, 12, 3, 0)
  )

plot_grid(p1, p2, ncol = 1, align = 'h', labels = "auto")
```

(ref:dow-jones-loess) Daily closing values of the Dow Jones Industrial Average for the year 2009. Data source: Yahoo! Finance

```{r dow-jones-loess, fig.asp = 0.5, fig.cap = '(ref:dow-jones-loess)'}
# LOESS (locally estimated scatterplot smoothing) 

dow_jones_industrial %>% filter(date >= ymd("2008/12/31") & date <= ymd("2010/01/10")) %>%
  mutate(
    close_100d_ave = moving_ave(date, close, 100)
  ) %>%
  ggplot(aes(date, close)) + 
  geom_line(color = "grey20", size = .25) +
  geom_line(aes(date, close_100d_ave, color = "100d"), size = .75, na.rm = TRUE) +
  geom_smooth(aes(color = "smooth"), size = .75, na.rm = TRUE, se = FALSE) +
  scale_color_manual(
    values = c(
      `100d` = "#d55e00",
      smooth = "#0072b2"
    ),
    breaks = c("smooth", "100d"),
    labels = c("LOESS smoother", "100 day average"),
    name = NULL
  ) + 
  scale_x_date(limits = c(ymd("2008-12-31"), ymd("2010-01-10")), expand = c(0, 0)) +
  xlab(NULL) + ylab("Dow Jones Industrial Average") +
  theme_dviz_grid(12) +
  theme(
    legend.position = c(1, 0.48),
    legend.justification = c(1, 0.5),
    legend.box.background = element_rect(fill = "white", color = NA),
    legend.box.margin = margin(0, 12, 6, 12),
    plot.margin = margin(3, 12, 3, 0)
  )
```



(ref:tank-capacity-smoothers) **Figure title.** (a) LOESS smoother. (b) Cubic regression splines with 5 knots. (c) Thin-plate regression spline with 3 knots. (d)  Gaussian process spline with 6 knots.

```{r tank-capacity-smoothers, fig.width=8.5, fig.asp = 0.75, fig.cap='(ref:tank-capacity-smoothers)'}
cars93 <- MASS::Cars93

cars_base <- ggplot(cars93, aes(x = Price, y = Fuel.tank.capacity)) + geom_point(color = "grey60") + 
  scale_x_continuous(name = "price (USD)",
                     breaks = c(20, 40, 60),
                     labels = c("$20,000", "$40,000", "$60,000")) +
  scale_y_continuous(name = "fuel tank capacity\n(US gallons)") +
  theme_minimal_grid(12)  
  
p1 <- cars_base + geom_smooth(se = FALSE, method = "loess", formula = y ~ x, color = "#0072B2")
p2 <- cars_base + geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 5, bs = 'cr'), color = "#0072B2")
p3 <- cars_base + geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 3), color = "#0072B2")
p4 <- cars_base + geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 6, bs = 'gp'), color = "#0072B2")

plot_grid(p1, p2, p3, p4, align = 'hv',
          labels = 'auto', label_fontface = "plain")

# See also: https://altaf-ali.github.io/ggplot_tutorial/smoother.html
```

## Explicit models

(ref:tank-capacity-model) **Figure title.** The solid line corresponds to the formula $y = 68.20 - 46.37 x/(x - 1,560)$.

```{r tank-capacity-model, fig.asp = 0.75, fig.cap = '(ref:tank-capacity-model)'}
fit.out <- nls(Fuel.tank.capacity ~ a*Price/(Price + b) + c, data = cars93, start = c(a = -45, b = -1, c = 70))
fit.df <- data.frame(Price = 7:62,
                     Fuel.tank.capacity = predict(fit.out, data.frame(Price = 7:62)))

ggplot(cars93, aes(x = Price, y = Fuel.tank.capacity)) + geom_point(color = "grey60") + 
  geom_line(data = fit.df, size = 1, color = "#0072B2") +
  scale_x_continuous(name = "price (USD)",
                     breaks = c(20, 40, 60),
                     labels = c("$20,000", "$40,000", "$60,000")) +
  scale_y_continuous(name = "fuel tank capacity\n(US gallons)") +
  theme_minimal_grid()
```


(ref:blue-jays-scatter-line) Head length versus body mass for 123 blue jays. The birds' sex is indicated by color. At the same body mass, male birds tend to have longer heads (and specifically, longer bills) than female birds. Data source: Keith Tarvin, Oberlin College

```{r blue-jays-scatter-line, fig.width = 5, fig.asp = 3/4, fig.cap='(ref:blue-jays-scatter-line)'}
ggplot(blue_jays, aes(Mass, Head, color = KnownSex, fill = KnownSex)) + 
  geom_point(pch = 21, color = "white", size = 2.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(name = "body mass (g)") +
  scale_y_continuous(name = "head length (mm)") +
  scale_fill_manual(
    values = c(F = "#D55E00B0", M = "#0072B2B0"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(direction = "horizontal")) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(direction = "horizontal")) +
  theme_dviz_grid() +
  theme(
    legend.position = c(1, 0.01),
    legend.justification = c(1, 0),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(20, "pt")
  )
```

(ref:biorxiv-logscale) Monthly submissions to the preprint server bioRxiv. **comment on the log scale.** Data source: Jordan Anaya, http://www.prepubmed.org/

```{r biorxiv-logscale, fig.cap = '(ref:biorxiv-logscale)'}
preprint_growth %>% filter(archive == "bioRxiv") %>%
  filter(count > 0) -> biorxiv_growth

ggplot(biorxiv_growth, aes(date, count)) + 
  geom_line(color = "#0072B2", size = .75) +
  scale_y_log10(limits = c(30, 1600),
                breaks = c(10*(3:9), 100*(1:9), 1000*(1:2)),
                labels = c("", "", "50", "", "", "", "",
                           "100", "", "", "", "500", "", "", "", "",
                           "1000", ""), expand = c(0, 0),
                name = "preprints / month") + 
  scale_x_date(name = NULL) +
  theme_dviz_open() +
  theme(plot.margin = margin(7, 7, 3, 0))
```


## Detrending and time-series decomposition


```{r}
hpi_trends <- house_prices %>%
  filter(year(date) >= 1980) %>%
  filter(state %in% c("California", "Nevada", "West Virginia", "Texas")) %>%
  mutate(
    date_numeric = as.numeric(date),
    hpi = house_price_index,
    log_hpi = log(hpi)
  ) %>%
  group_by(state) %>%
  mutate(
    hpi_trend = {
      coefs <- coef(lm(log_hpi ~ date_numeric))
      exp(coefs[1] + coefs[2]*date_numeric)
    },
    hpi_detrended = hpi/hpi_trend
  )

ggplot(hpi_trends, aes(date, hpi)) +
  geom_line(aes(y = hpi_trend)) +
  geom_line() +
  scale_y_log10() +
  facet_wrap(~state)

ggplot(hpi_trends, aes(date, hpi_detrended)) +
  geom_hline(yintercept = 1) +
  geom_line() +
  facet_wrap(~state) +
  scale_y_log10(
    breaks = c(0.752, 1, 1.33, 1.77)
  )

```

```{r}
# use complete years only
CO2_complete <- filter(CO2, year >= 1959, year < 2018)
# convert to time series object
CO2_ts <- ts(data = CO2_complete$co2_interp, start = 1959, end = c(2017, 12), frequency = 12)
# detrend via STL method
CO2_stl <- stl(CO2_ts, s.window = 7)
CO2_detrended <- mutate(
  CO2_complete,
  seasonal = t(CO2_stl$time.series)[1, ],
  trend = t(CO2_stl$time.series)[2, ],
  remainder = t(CO2_stl$time.series)[3, ]
)

# plot
p1 <- ggplot(CO2_detrended, aes(date_dec, co2_interp)) + 
  geom_line()
p2 <- ggplot(CO2_detrended, aes(date_dec, seasonal)) + 
  geom_line()
p3 <- ggplot(CO2_detrended, aes(date_dec, trend)) + 
  geom_line()
p4 <- ggplot(CO2_detrended, aes(date_dec, remainder)) + 
  geom_line()

plot_grid(p1, p2, p3, p4, ncol = 1, align = 'v')
```

```{r}
# links with useful articles about detrending:
# https://anomaly.io/seasonal-trend-decomposition-in-r/
# http://r-statistics.co/Time-Series-Analysis-With-R.html
# https://econometricswithr.wordpress.com/2015/10/27/extracting-business-cycles-from-raw-data-in-r/
```
