```{r echo = FALSE, message = FALSE, warning = FALSE}
# run setup script
source("_common.R")

library(dplyr)
library(tidyr)
```

# Visualizing associations among two or more quantitative variables {#visualizing-associations}

Outline:

- Basic x-y scatter plot
- Very dense plots: 2d histograms or contour lines (maybe `faithful`?)
- More than 2 variables: Map to color or size, or make matrix of x-y plots
- Correlation matrix (pottery or bumpus dataset?)
- PCA (pottery dataset?)
- Paired data (diagonal plot, parallel dots plot)

```{r}
ggplot(diamonds, aes(carat, price)) + 
  geom_point(size = 0.1) + 
  geom_density2d() +
  scale_x_continuous(limits = c(0, 5.1), expand = c(0, 0)) +
  scale_y_log10(limits = c(300, 20000),
                breaks = c(300, 1000, 3000, 10000)) +
  #facet_grid(color~clarity) +
  theme_dviz_grid()
```

```{r}
ggplot(diamonds, aes(carat, price)) + 
  geom_bin2d(bins = 100) +
  scale_x_continuous(limits = c(0, 5.1), expand = c(0, 0)) +
  scale_y_log10(limits = c(300, 20000),
                breaks = c(300, 1000, 3000, 10000)) +
  scale_fill_viridis_c(option = "B", begin = 0.1) +
  theme_dviz_grid()
```



```{r fig.width = 4.5, fig.asp = 1}
df <- MASS::fgl

cm <- cor(select(df, -type, -RI))
df_wide <- as.data.frame(cm)
df_long <- stack(df_wide)
names(df_long) <- c("cor", "var1")
df_long <- cbind(df_long, var2 = rep(rownames(cm), length(rownames(cm))))
clust <- hclust(as.dist(1-cm), method="average") 
levels <- clust$labels[clust$order]
df_long$var1 <- factor(df_long$var1, levels = levels)
df_long$var2 <- factor(df_long$var2, levels = levels)
ggplot(filter(df_long, as.integer(var1) < as.integer(var2)),
       aes(var1, var2, fill=cor)) + 
  geom_tile(color = "white", size = 1) + 
  scale_x_discrete(position = "top", name = NULL, expand = c(0, 0)) +
  scale_y_discrete(name = NULL, expand = c(0, 0)) +
  scale_fill_continuous_carto(
    palette = "Earth", rev = TRUE,
    limits = c(-.5, .5), breaks = c(-.5, 0, .5),
    name = "correlation",
    guide = guide_colorbar(
      direction = "horizontal",
      label.position = "bottom",
      title.position = "top",
      barwidth = grid::unit(140, "pt"),
      barheight = grid::unit(17.5, "pt"))) +
  coord_fixed() +
  theme_dviz_open(rel_small = 1) +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.ticks.length = grid::unit(3, "pt"),
        legend.position = c(.97, .0),
        legend.justification = c(1, 0),
        legend.title.align = 0.5)

```

```{r fig.width = 4.5, fig.asp = 1}
ggplot(filter(df_long, as.integer(var1) < as.integer(var2)),
       aes(var1, var2, fill=cor, size = abs(cor))) + 
  geom_point(shape = 21, stroke = 0) + 
  scale_x_discrete(position = "top", name = NULL, expand = c(0, 0.5)) +
  scale_y_discrete(name = NULL, expand = c(0, 0.5)) +
  scale_size_area(max_size = 19, limits = c(0, 0.5), guide = "none") +
  scale_fill_continuous_carto(
    palette = "Earth", rev = TRUE,
    limits = c(-.5, .5), breaks = c(-.5, 0, .5),
    name = "correlation",
    guide = guide_colorbar(
      direction = "horizontal",
      label.position = "bottom",
      title.position = "top",
      barwidth = grid::unit(140, "pt"),
      barheight = grid::unit(17.5, "pt"))) +
  coord_fixed() +
  theme_dviz_open(rel_small = 1) +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.ticks.length = grid::unit(3, "pt"),
        legend.position = c(.97, .0),
        legend.justification = c(1, 0),
        legend.title.align = 0.5)

```
