```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")

library(lubridate)
library(tidyr)
```

# Visualizing trends {#visualizing-trends}



## Smoothing

*A few sentences stock market intro.*

The market lost over 2400 points (~27%) in the first three months of the year, and then it slowly recovered for the remainder of the year. How can we visualize these longer-term trends while de-emphasizing the less important short-term fluctuations?

(ref:dow-jones) Daily closing values of the Dow Jones Industrial Average for the year 2009. Data source: Yahoo! Finance

```{r dow-jones, fig.asp = 0.5, fig.cap = '(ref:dow-jones)'}
dow_jones_industrial %>% filter(date >= ymd("2008/12/31") & date <= ymd("2010/01/10")) %>%
  ggplot(aes(date, close)) + 
  geom_line(color = "grey20", size = .5) + 
  scale_x_date(limits = c(ymd("2008-12-31"), ymd("2010-01-10")), expand = c(0, 0)) +
  xlab(NULL) + ylab("Dow Jones Industrial Average") +
  theme_dviz_grid(12) +
  theme(
    plot.margin = margin(3, 12, 3, 0)
  )
```

In the financial world, it is commonplace to visualize trends with moving averages. To calculate a moving average for a stock price, we take a time window, say the first 20 days in the time series, calculate the average price over these 20 days, then move the time window by a day so it now spans the 2nd to 21st day, calculate the average over these 20 days, move the time window again, and so on. The result is a new time series consisting of a sequence of averaged prices.

*financial vs. statistical moving average, end of interval vs. center.*


(ref:dow-jones-moving-ave) Daily closing values of the Dow Jones Industrial Average for the year 2009, shown together with their 20-day, 50-day, and 100-day moving averages. (a) The moving averages are plotted at the end of the moving time windows. (b) The moving averages are plotted in the center of the moving time windows. Data source: Yahoo! Finance

```{r dow-jones-moving-ave, fig.asp = 1, fig.cap = '(ref:dow-jones-moving-ave)'}
p1 <- dow_jones_industrial %>% filter(date >= ymd("2008/12/31") & date <= ymd("2010/01/10")) %>%
  mutate(
    close_20d_ave = moving_ave(date, close, 20, center = FALSE),
    close_50d_ave = moving_ave(date, close, 50, center = FALSE),
    close_100d_ave = moving_ave(date, close, 100, center = FALSE)
  ) %>%
  ggplot(aes(date, close)) + 
  geom_line(color = "grey20", size = .25) +
  geom_line(aes(date, close_20d_ave, color = "20d"), size = .75, na.rm = TRUE) +
  geom_line(aes(date, close_50d_ave, color = "50d"), size = .75, na.rm = TRUE) +
  geom_line(aes(date, close_100d_ave, color = "100d"), size = .75, na.rm = TRUE) +
  scale_color_manual(
    values = c(
      `20d` = "#009e73",
      `50d` = "#d55e00",
      `100d` = "#0072b2"
    ),
    breaks = c("20d", "50d", "100d"),
    labels = c("20 day average", "50 day average", "100 day average"),
    name = NULL
  ) + 
  scale_x_date(
    limits = c(ymd("2008-12-31"), ymd("2010-01-10")), expand = c(0, 0),
    labels = NULL
  ) +
  xlab(NULL) + ylab("Dow Jones industrial average") +
  theme_dviz_grid(12) +
  theme(
    plot.margin = margin(3, 12, 3, 0),
    legend.position = c(1, 0),
    legend.justification = c(1, 0),
    legend.box.background = element_rect(fill = "white", color = NA),
    legend.box.margin = margin(6, 12, 0, 12),
    axis.ticks.x = element_blank()
  )

p2 <- dow_jones_industrial %>% filter(date >= ymd("2008/12/31") & date <= ymd("2010/01/10")) %>%
  mutate(
    close_20d_ave = moving_ave(date, close, 20, center = TRUE),
    close_50d_ave = moving_ave(date, close, 50, center = TRUE),
    close_100d_ave = moving_ave(date, close, 100, center = TRUE)
  ) %>%
  ggplot(aes(date, close)) + 
  geom_line(color = "grey20", size = .25) +
  geom_line(aes(date, close_20d_ave, color = "20d"), size = .75, na.rm = TRUE) +
  geom_line(aes(date, close_50d_ave, color = "50d"), size = .75, na.rm = TRUE) +
  geom_line(aes(date, close_100d_ave, color = "100d"), size = .75, na.rm = TRUE) +
  scale_color_manual(
    values = c(
      `20d` = "#009e73",
      `50d` = "#d55e00",
      `100d` = "#0072b2"
    ),
    breaks = c("20d", "50d", "100d"),
    labels = c("20 day average", "50 day average", "100 day average"),
    name = NULL,
    guide = "none"
  ) + 
  scale_x_date(limits = c(ymd("2008-12-31"), ymd("2010-01-10")), expand = c(0, 0)) +
  xlab(NULL) + ylab("Dow Jones Industrial Average") +
  theme_dviz_grid(12) +
  theme(
    plot.margin = margin(3, 12, 3, 0)
  )

plot_grid(p1, p2, ncol = 1, align = 'h', labels = "auto")
```

Because the drop to below 7000 points happened rapidly and the market quickly recovered to levels above 7500 points, the smoothing function removes this severe dip and replaces it with a gentle bend that doesn't dip much below 8000 points.

(ref:dow-jones-loess) Daily closing values of the Dow Jones Industrial Average for the year 2009. Data source: Yahoo! Finance

```{r dow-jones-loess, fig.asp = 0.5, fig.cap = '(ref:dow-jones-loess)'}
# LOESS (locally estimated scatterplot smoothing) 

dow_jones_industrial %>% filter(date >= ymd("2008/12/31") & date <= ymd("2010/01/10")) %>%
  mutate(
    close_100d_ave = moving_ave(date, close, 100)
  ) %>%
  ggplot(aes(date, close)) + 
  geom_line(color = "grey20", size = .25) +
  geom_line(aes(date, close_100d_ave, color = "100d"), size = .75, na.rm = TRUE) +
  geom_smooth(aes(color = "smooth"), size = .75, na.rm = TRUE, se = FALSE) +
  scale_color_manual(
    values = c(
      `100d` = "#d55e00",
      smooth = "#0072b2"
    ),
    breaks = c("smooth", "100d"),
    labels = c("LOESS smoother", "100 day average"),
    name = NULL
  ) + 
  scale_x_date(limits = c(ymd("2008-12-31"), ymd("2010-01-10")), expand = c(0, 0)) +
  xlab(NULL) + ylab("Dow Jones Industrial Average") +
  theme_dviz_grid(12) +
  theme(
    legend.position = c(1, 0.48),
    legend.justification = c(1, 0.5),
    legend.box.background = element_rect(fill = "white", color = NA),
    legend.box.margin = margin(0, 12, 6, 12),
    plot.margin = margin(3, 12, 3, 0)
  )
```

There are many different smoothing functions ... This is important, because most statistical packages will provide a default smoother, and you may never realize to what extent the results you see depend on the specific default choices made by your statistical software.

(ref:tank-capacity-smoothers) **Figure title.** (a) LOESS smoother. (b) Cubic regression splines with 5 knots. (c) Thin-plate regression spline with 3 knots. (d)  Gaussian process spline with 6 knots.

```{r tank-capacity-smoothers, fig.width=8.5, fig.asp = 0.75, fig.cap='(ref:tank-capacity-smoothers)'}
cars93 <- MASS::Cars93

cars_base <- ggplot(cars93, aes(x = Price, y = Fuel.tank.capacity)) + geom_point(color = "grey60") + 
  scale_x_continuous(name = "price (USD)",
                     breaks = c(20, 40, 60),
                     labels = c("$20,000", "$40,000", "$60,000")) +
  scale_y_continuous(name = "fuel tank capacity\n(US gallons)") +
  theme_minimal_grid(12)  
  
p1 <- cars_base + geom_smooth(se = FALSE, method = "loess", formula = y ~ x, color = "#0072B2")
p2 <- cars_base + geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 5, bs = 'cr'), color = "#0072B2")
p3 <- cars_base + geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 3), color = "#0072B2")
p4 <- cars_base + geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 6, bs = 'gp'), color = "#0072B2")

plot_grid(p1, p2, p3, p4, align = 'hv',
          labels = 'auto', label_fontface = "plain")

# See also: https://altaf-ali.github.io/ggplot_tutorial/smoother.html
```

## Explicit models

(ref:tank-capacity-model) **Figure title.** The solid line corresponds to the formula $y = 68.20 - 46.37 x/(x - 1,560)$.

```{r tank-capacity-model, fig.asp = 0.75, fig.cap = '(ref:tank-capacity-model)'}
fit.out <- nls(Fuel.tank.capacity ~ a*Price/(Price + b) + c, data = cars93, start = c(a = -45, b = -1, c = 70))
fit.df <- data.frame(Price = 7:62,
                     Fuel.tank.capacity = predict(fit.out, data.frame(Price = 7:62)))

ggplot(cars93, aes(x = Price, y = Fuel.tank.capacity)) + geom_point(color = "grey60") + 
  geom_line(data = fit.df, size = 1, color = "#0072B2") +
  scale_x_continuous(name = "price (USD)",
                     breaks = c(20, 40, 60),
                     labels = c("$20,000", "$40,000", "$60,000")) +
  scale_y_continuous(name = "fuel tank capacity\n(US gallons)") +
  theme_minimal_grid()
```


(ref:blue-jays-scatter-line) Head length versus body mass for 123 blue jays. The birds' sex is indicated by color. At the same body mass, male birds tend to have longer heads (and specifically, longer bills) than female birds. Data source: Keith Tarvin, Oberlin College

```{r blue-jays-scatter-line, fig.width = 5, fig.asp = 3/4, fig.cap='(ref:blue-jays-scatter-line)'}
ggplot(blue_jays, aes(Mass, Head, color = KnownSex, fill = KnownSex)) + 
  geom_point(pch = 21, color = "white", size = 2.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(name = "body mass (g)") +
  scale_y_continuous(name = "head length (mm)") +
  scale_fill_manual(
    values = c(F = "#D55E00B0", M = "#0072B2B0"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(direction = "horizontal")) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(direction = "horizontal")) +
  theme_dviz_grid() +
  theme(
    legend.position = c(1, 0.01),
    legend.justification = c(1, 0),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(20, "pt")
  )
```

(ref:biorxiv-logscale) Monthly submissions to the preprint server bioRxiv. **comment on the log scale.** Data source: Jordan Anaya, http://www.prepubmed.org/

```{r biorxiv-logscale, fig.cap = '(ref:biorxiv-logscale)'}
preprint_growth %>% filter(archive == "bioRxiv") %>%
  filter(count > 0) -> biorxiv_growth

ggplot(biorxiv_growth, aes(date, count)) + 
  geom_line(color = "#0072B2", size = .75) +
  scale_y_log10(limits = c(30, 1600),
                breaks = c(10*(3:9), 100*(1:9), 1000*(1:2)),
                labels = c("", "", "50", "", "", "", "",
                           "100", "", "", "", "500", "", "", "", "",
                           "1000", ""), expand = c(0, 0),
                name = "preprints / month") + 
  scale_x_date(name = NULL) +
  theme_dviz_open() +
  theme(plot.margin = margin(7, 7, 3, 0))
```


## Detrending and time-series decomposition

(ref:hpi-trends) Freddie Mac House Price Index from 1980 through 2017, for four selected states (California, Nevada, Texas, and West Virginia). The House Price Index is a unitless number that tracks relative house prices in the chosen geographic region over time. The index is scaled arbitrarily such that it equals 100 in December of the year 2000. The blue lines show the monthly fluctuations in the index and the straight gray lines show the long-term price trends in the respective states. Data source: Freddie Mac House Prices Index

```{r hpi-trends, fig.cap = '(ref:hpi-trends)'}
hpi_trends <- house_prices %>%
  filter(year(date) >= 1980) %>%
  filter(state %in% c("California", "Nevada", "West Virginia", "Texas")) %>%
  mutate(
    date_numeric = as.numeric(date),
    hpi = house_price_index,
    log_hpi = log(hpi)
  ) %>%
  group_by(state) %>%
  mutate(
    hpi_trend = {
      coefs <- coef(lm(log_hpi ~ date_numeric))
      exp(coefs[1] + coefs[2]*date_numeric)
    },
    hpi_detrended = hpi/hpi_trend
  )

ggplot(hpi_trends, aes(date, hpi)) +
  geom_line(aes(y = hpi_trend), color = "grey50", size = 0.4) +
  geom_line(color = "#0072B2") +
  scale_x_date(name = NULL) +
  scale_y_log10(name = "House Price Index (Dec. 2000 = 100)") +
  facet_wrap(~state, scales = "free_x") +
  theme_dviz_hgrid(12, rel_small = 1) +
  theme(
    strip.background = element_rect(fill = "grey90"),
    axis.line.x = element_line(color = "grey50"),
    axis.ticks.x = element_line(color = "grey50"),
    axis.ticks.y = element_blank(),
    axis.text.y = element_text(margin = margin(0, 0, 0, 0))
  )
    
```

```{r}
ggplot(hpi_trends, aes(date, hpi_detrended)) +
  geom_hline(yintercept = 1, color = "grey50", size = 0.4) +
  geom_line(color = "#0072B2") +
  scale_x_date(name = NULL) +
  scale_y_log10(
    name = "House Price Index (detrended)",
    breaks = c(0.752, 1, 1.33, 1.77),
    labels = c("0.75", "1.00", "1.33", "1.77")
  ) +
  facet_wrap(~state, scales = "free_x") +
  theme_dviz_hgrid(12, rel_small = 1) +
  theme(
    strip.background = element_rect(fill = "grey90"),
    axis.line.x = element_line(color = "grey50"),
    axis.ticks.x = element_line(color = "grey50"),
    axis.ticks.y = element_blank(),
    axis.text.y = element_text(margin = margin(0, 0, 0, 0))
  )
 
 
```

```{r fig.asp = 0.9}
# use complete years only
CO2_complete <- filter(CO2, year >= 1959, year < 2018)
# convert to time series object
CO2_ts <- ts(data = CO2_complete$co2_interp, start = 1959, end = c(2017, 12), frequency = 12)
# detrend via STL method
CO2_stl <- stl(CO2_ts, s.window = 7)
CO2_detrended <- mutate(
  CO2_complete,
  `monthly average` = co2_interp,
  `seasonal fluctuations` = t(CO2_stl$time.series)[1, ],
  `longterm trend` = t(CO2_stl$time.series)[2, ],
  remainder = t(CO2_stl$time.series)[3, ]
)

facet_labels <- c("monthly average", "longterm trend", "seasonal fluctuations", "remainder")

CO2_detrended %>%
  select(date_dec, `monthly average`, `seasonal fluctuations`, `longterm trend`, remainder) %>%
  gather(variable, value, -date_dec) %>%
  mutate(variable = factor(variable, levels = facet_labels)) %>%
  filter(date_dec >= 1989) %>%
  ggplot(aes(date_dec, value)) +
  geom_line(color = "#0072B2") +
  geom_point(
    data = data.frame(
      variable = factor(
        rep(facet_labels, each = 2),
        levels = facet_labels
      ),
      x = 1990,
      #y = c(295, 424, 295, 424, -4.1, 4.3, -.81, .85)
      y = c(324, 419, 324, 419, -4.1, 4.3, -.81, .85)
    ),
    aes(x, y),
    color = NA,
    na.rm = TRUE
  ) +
  scale_y_continuous(
    name = parse(text = "`CO`[2]*` concentration (ppm)`"),
    expand = c(0, 0)
  ) +
  scale_x_continuous(
    limits = c(1989, 2018.2),
    name = NULL,
    breaks = c(1990, 1995, 2000, 2005, 2010, 2015),
    labels = c("1990", "", "2000", "", "2010", ""),
    expand = c(0, 0)
  ) +
  facet_wrap(facets = vars(variable), ncol = 1, scales = "free") +
  theme_dviz_grid(12) +
  theme(
    plot.margin = margin(3, 0, 3, 1),
    strip.text = element_text(size = 11)
  )
```

```{r}
# links with useful articles about detrending:
# https://anomaly.io/seasonal-trend-decomposition-in-r/
# http://r-statistics.co/Time-Series-Analysis-With-R.html
# https://econometricswithr.wordpress.com/2015/10/27/extracting-business-cycles-from-raw-data-in-r/
```
