```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")
```

# Visualizing trends {#visualizing-trends}

Topics to cover:

- linear fits
- non-linear fits
- exponential growth
- averaging/detrending?

(ref:tank-capacity-smoothers) **Figure title.** (a) LOESS smoother. (b) Cubic regression splines with 5 knots. (c) Thin-plate regression spline with 3 knots. (d)  Gaussian process spline with 6 knots.

```{r tank-capacity-smoothers, fig.height = 8.5 * 0.75, fig.cap='(ref:tank-capacity-smoothers)'}
cars93 <- MASS::Cars93

cars_base <- ggplot(cars93, aes(x = Price, y = Fuel.tank.capacity)) + geom_point(color = "grey60") + 
  scale_x_continuous(name = "price (USD)",
                     breaks = c(20, 40, 60),
                     labels = c("$20,000", "$40,000", "$60,000")) +
  scale_y_continuous(name = "fuel tank capacity\n(US gallons)") +
  theme_minimal_grid(12)  
  
p1 <- cars_base + geom_smooth(se = FALSE, method = "loess", formula = y ~ x, color = "#0072B2")
p2 <- cars_base + geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 5, bs = 'cr'), color = "#0072B2")
p3 <- cars_base + geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 3), color = "#0072B2")
p4 <- cars_base + geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 6, bs = 'gp'), color = "#0072B2")

plot_grid(p1, p2, p3, p4, align = 'hv',
          labels = 'auto', label_fontface = "plain")

# See also: https://altaf-ali.github.io/ggplot_tutorial/smoother.html
```


(ref:tank-capacity-model) **Figure title.** The solid line corresponds to the formula $y = 68.20 - 46.37 x/(x - 1,560)$.

```{r tank-capacity-model, fig.height = 6 * 0.75, fig.cap = '(ref:tank-capacity-model)'}
fit.out <- nls(Fuel.tank.capacity ~ a*Price/(Price + b) + c, data = cars93, start = c(a = -45, b = -1, c = 70))
fit.df <- data.frame(Price = 7:62,
                     Fuel.tank.capacity = predict(fit.out, data.frame(Price = 7:62)))

p <- ggplot(cars93, aes(x = Price, y = Fuel.tank.capacity)) + geom_point(color = "grey60") + 
  geom_line(data = fit.df, size = 1, color = "#0072B2") +
  scale_x_continuous(name = "price (USD)",
                     breaks = c(20, 40, 60),
                     labels = c("$20,000", "$40,000", "$60,000")) +
  scale_y_continuous(name = "fuel tank capacity\n(US gallons)") +
  theme_minimal_grid()

pad_figure(p, fig.width = 6)
```
