```{r echo = FALSE, message = FALSE, warning = FALSE}
# run setup script
source("_common.R")
```

# Visualizing associations among two or more quantitative variables {#visualizing-associations}

Outline:

- Basic x-y scatter plot
- Very dense plots: 2d histograms or contour lines (maybe `faithful`?)
- More than 2 variables: Map to color or size, or make matrix of x-y plots
- Correlation matrix (pottery or bumpus dataset?)
- PCA (pottery dataset?)
- Paired data (diagonal plot, parallel dots plot)

```{r}
ggplot(diamonds, aes(carat, price)) + 
  geom_point(size = 0.1) + 
  geom_density2d() +
  scale_x_continuous(limits = c(0, 5.1), expand = c(0, 0)) +
  scale_y_log10(limits = c(300, 20000),
                breaks = c(300, 1000, 3000, 10000)) +
  theme_dviz_grid()
```

```{r}
ggplot(diamonds, aes(carat, price)) + 
  geom_bin2d(bins = 100) +
  scale_x_continuous(limits = c(0, 5.1), expand = c(0, 0)) +
  scale_y_log10(limits = c(300, 20000),
                breaks = c(300, 1000, 3000, 10000)) +
  scale_fill_viridis_c(option = "B", begin = 0.2) +
  theme_dviz_grid()
```
