```{r echo = FALSE, message = FALSE, warning = FALSE}
# run setup script
source("_common.R")

library(forcats)
library(patchwork)
library(lubridate)
library(mgcv)
library(mvtnorm)
library(rlang)
library(tidyr)
library(purrr)
library(broom)
library(emmeans)
library(ungeviz)
library(ggridges)
```

# Visualizing uncertainty {#visualizing-uncertainty}

*Chapter under construction. Tentative outline:*

- *Error bars and confidence bands*
- *Distributions*
- *Hypothetical outcomes plots*
- *Value suppression*


## Error bars and confidence bands

(ref:cocoa-data-vs-CI) Relationship between original data and mean, standard deviation, standard error, and confidence intervals. The original data points (small green dots) represent expert ratings of 125 chocolate bars from manufacturers in Canada, rated on a scale from 1 (unpleasant) to 5 (elite). The large orange dots represent the mean of the ratings. Error bars indicate, from top to bottom, the standard deviation, the standard error (standard deviation of the mean), and 80%, 95%, and 99% confidence intervals of the mean. Data source: Brady Brelinski, Manhattan Chocolate Society

```{r cocoa-data-vs-CI, fig.asp = 0.48, warning = FALSE, message = FALSE, fig.cap = '(ref:cocoa-data-vs-CI)'}

# color for individual small data points
point_color <- darken("#009E73", .3)

cacao %>% 
  filter(location == "Canada") -> cacao_single

fit <- lm(rating ~ 1, data = cacao_single)

CI_df <- data.frame(type = c(0.8, 0.95, 0.99)) %>%
  mutate(df = map(type, ~tidy(emmeans(fit, ~ 1, options = list(level = .x))))) %>%
  unnest() %>%
  select(type, estimate, std.error, conf.low, conf.high) %>%
  mutate(type = paste0(signif(100*type, 2), "% confidence interval"))

CI_df <- rbind(
  CI_df,
  data.frame(
    type = "standard error",
    estimate = CI_df$estimate[1],
    std.error = CI_df$std.error[1],
    conf.low = CI_df$estimate[1] - CI_df$std.error[1],
    conf.high = CI_df$estimate[1] + CI_df$std.error[1]
  ),
  data.frame(
    type = "standard deviation",
    estimate = mean(cacao_single$rating),
    std.error = CI_df$std.error[1],
    conf.low = mean(cacao_single$rating) - sd(cacao_single$rating),
    conf.high = mean(cacao_single$rating) + sd(cacao_single$rating)
  ),
  data.frame(
    type = "original data", estimate = mean(cacao_single$rating), std.error = NA,
    conf.low = NA, conf.high = max(cacao_single$rating)
  )
) %>%
  mutate(
    type = fct_relevel(factor(type), "original data", "standard deviation", "standard error"),
    label = case_when(
      type == "original data" ~ NA_character_,
      TRUE ~ as.character(type) #paste0("mean +/- ", type)
    )
  )

label_x <- filter(CI_df, type == "standard deviation")$conf.high + 0.04

ggplot(CI_df, aes(estimate, type)) + 
  geom_point(
    data = cacao_single, aes(rating, "original data"), 
    position = position_jitter(height = 0.6, width = 0.02, seed = 7843),
    color = point_color,
    size = 0.3
  ) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, na.rm = TRUE) +
  geom_point(size = 2, color = "#D55E00") +
  geom_label(
    aes(label_x, label = label), hjust = 0, nudge_y = 0.01, na.rm = TRUE,
    family = dviz_font_family,
    size = 12/.pt,
    label.size = 0
  ) +
  geom_label(
    data = filter(CI_df, type == "original data"),
    aes(conf.high + 0.06, label = type), hjust = 0, nudge_y = 0.01,
    family = dviz_font_family,
    size = 12/.pt,
    label.size = 0
  ) +
  geom_text(
    data = filter(CI_df, type == "original data"),
    aes(estimate, label = "mean"), hjust = 0.2, vjust = 0, nudge_y = 0.2,
    family = dviz_font_family,
    size = 12/.pt
  ) +
  scale_x_continuous(
    limits = c(1.95, 4.1),
    expand = c(0, 0),
    name = "chocolate flavor rating"
  ) +
  scale_y_discrete(
    name = NULL,
    limits = rev(levels(CI_df$type)),
    expand = expand_scale(add = c(0.6, 0.8)),
    breaks = NULL
  ) +
  coord_cartesian(clip = "off") +
  theme_dviz_vgrid(12, rel_small = 1) +
  theme(
    plot.margin = margin(3, 73, 3, 0),
    axis.line.x = element_line(),
    axis.ticks.x = element_line(color = "black"),
    axis.title.x = element_text(hjust = 1)
  )

```

(ref:cocoa-CI-vs-n) Confidence intervals widen with smaller sample size. Chocolate bars from Canada and Switzerland have comparable mean ratings and comparable standard deviations (indicated with simple black error bars). However, over three times as many Canadian bars were rated than Swiss bars, and therefore the confidence intervals around the means (indicated with stacked, colored confidence bands) are substantially wider for the mean of the Swiss ratings than for the mean of the Canadian ratings. Data source: Brady Brelinski, Manhattan Chocolate Society

```{r cocoa-CI-vs-n, fig.asp = 0.46, warning = FALSE, message = FALSE, fig.cap = '(ref:cocoa-CI-vs-n)'}
cacao_CA <- filter(cacao, location == "Canada")
cacao_CH <- filter(cacao, location == "Switzerland")

fit_CA <- lm(rating ~ 1, data = cacao_CA)
fit_CH <- lm(rating ~ 1, data = cacao_CH)

CI_CA <- data.frame(level = c(0.99, 0.95, 0.8)) %>%
  mutate(df = map(level, ~tidy(emmeans(fit_CA, ~ 1, options = list(level = .x))))) %>%
  unnest() %>%
  select(level, estimate, conf.low, conf.high) %>%
  mutate(
    level = paste0(signif(100*level, 2), "%"),
    type = "CI",
    location = "Canada"
  )

CI_CH <- data.frame(level = c(0.99, 0.95, 0.8)) %>%
  mutate(df = map(level, ~tidy(emmeans(fit_CH, ~ 1, options = list(level = .x))))) %>%
  unnest() %>%
  select(level, estimate, conf.low, conf.high) %>%
  mutate(
    level = paste0(signif(100*level, 2), "%"),
    type = "CI",
    location = "Switzerland"
  )

CI_df <- rbind(CI_CA, CI_CH)

sd_df <- rbind(
  data.frame(
    level = NA,
    estimate = mean(cacao_CA$rating),
    conf.low = mean(cacao_CA$rating) - sd(cacao_CA$rating),
    conf.high = mean(cacao_CA$rating) + sd(cacao_CA$rating),
    type = "original data",
    location = "Canada"
  ),
  data.frame(
    level = NA,
    estimate = mean(cacao_CH$rating),
    conf.low = mean(cacao_CH$rating) - sd(cacao_CH$rating),
    conf.high = mean(cacao_CH$rating) + sd(cacao_CH$rating),
    type = "original data",
    location = "Switzerland"
  )
)

#label_x <- filter(CI_df, type == "standard deviation")$conf.high + 0.04

ggplot(rbind(CI_df, sd_df), aes(estimate, interaction(location, type))) + 
  geom_point(
    data = cacao_CA, # draw two separate layers to get jittering right relative to previous figure
    aes(rating, interaction(location, "original data")),
    position = position_jitter(height = 0.6, width = 0.02, seed = 7843),
    color = point_color,
    size = 0.3
  ) +
  geom_point(
    data = cacao_CH,
    aes(rating, interaction(location, "original data")),
    position = position_jitter(height = 0.6, width = 0.02, seed = 7844),
    color = point_color,
    size = 0.3
  ) +
  geom_errorbarh(
    data = sd_df,
    aes(y = interaction(location, "original data"), xmin = conf.low, xmax = conf.high),
    height = 0.2
  ) +
#  geom_tile(
#    data = CI_df,
#    aes(width = conf.high - conf.low, fill = level), height = 0.2
#  ) +
  geom_errorbarh(
    data = CI_df,
    aes(y = interaction(location, "CI"), xmin = conf.low, xmax = conf.high, color = level, size = level),
    height = 0
  ) +
  geom_errorbarh(
    data = CI_df,
    aes(y = interaction(location, "CI"), xmin = conf.low, xmax = conf.high, color = level),
    height = 0.2
  ) +
  geom_point(size = 2, color = "#D55E00") +
  geom_label(
    data = data.frame(
      estimate = 4.06,
      location = c("Canada", "Switzerland"),
      type = "original data",
      label = c(
        paste0("Canada,\nn = ", nrow(cacao_CA)),
        paste0("Switzerland,\nn = ", nrow(cacao_CH))
      )
    ),
    aes(label = label), hjust = 0, vjust = 0.5, nudge_y = 0.01,
    family = dviz_font_family,
    size = 12/.pt,
    label.size = 0
  ) +
  scale_x_continuous(
    limits = c(1.95, 4.1),
    expand = c(0, 0),
    name = "chocolate flavor rating"
  ) +
  scale_y_discrete(
    name = NULL,
    limits = rev(c("Canada.original data", "Canada.CI", "dummy", "Switzerland.original data", "Switzerland.CI")),
    expand = expand_scale(add = c(1, 0.8)),
    breaks = NULL
  ) +
  scale_fill_manual(
    aesthetics = c("color", "fill"),
    name = "confidence level",
    values = c(
     `80%` = desaturate(darken("#0072B2", .2), .3),
     `95%` = desaturate(lighten("#0072B2", .2), .3),
     `99%` = desaturate(lighten("#0072B2", .4), .3)
    ),
    guide = guide_legend(
      direction = "horizontal",
      title.position = "top",
      label.position = "bottom"
    )
  ) +
  scale_size_manual(
    name = "confidence level",
    values = c(
     `80%` = 1.5,
     `95%` = 1,
     `99%` = 0.5
    ),
    guide = guide_legend(
      direction = "horizontal",
      title.position = "top",
      label.position = "bottom"
    )
  ) +
  coord_cartesian(clip = "off") +
  theme_dviz_vgrid(12, rel_small = 1) +
  theme(
    plot.margin = margin(3, 73, 3, 0),
    axis.line.x = element_line(),
    axis.ticks.x = element_line(color = "black"),
    axis.title.x = element_text(hjust = 1),
    legend.position = c(0, 0.01),
    legend.justification = c(0, 0),
    legend.key.height = grid::unit(6, "pt"),
    legend.key.width = grid::unit(24, "pt"),
    legend.spacing.x = grid::unit(6, "pt"),
    legend.spacing.y = grid::unit(3, "pt"),
    legend.box.background = element_rect(fill = "white", color = NA),
    legend.box.spacing = grid::unit(0, "pt"),
    legend.title.align = 0.5
  )

```

(ref:mean-chocolate-ratings) Mean chocolate flavor ratings and associated confidence intervals for chocolate bars from manufacturers in six different countries. Data source: Brady Brelinski, Manhattan Chocolate Society

```{r mean-chocolate-ratings, fig.asp = 0.5, fig.cap = '(ref:mean-chocolate-ratings)'}
countries <- c("Austria", "Belgium", "Canada", "Peru", "Switzerland")

cacao_small <- cacao %>% 
  filter(location %in% c("U.S.A.", countries)) %>%
  mutate(location = factor(location, levels = c("U.S.A.", countries)))

fit <- lm(rating ~ location, data = cacao_small)

conf_df <- data.frame(level = c(0.99, 0.95, 0.8)) %>%
  mutate(df = map(level, ~tidy(emmeans(fit, ~location, options = list(level = .x))))) %>%
  unnest() %>%
  select(level, location, estimate, std.error, conf.low, conf.high) %>%
  mutate(level = paste0(signif(100*level, 2), "%"))

ggplot(conf_df, aes(estimate, reorder(location, estimate))) + 
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high, color = level, size = level),
    height = 0
  ) +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high, color = level),
    height = 0.2
  ) +
  geom_point(data = filter(conf_df, level == "80%"), size = 2.5, color = "#D55E00") +
  scale_x_continuous(
    limits = c(2.6, 3.6),
#    expand = c(0, 0),
    name = "chocolate flavor rating"
  ) +
 scale_y_discrete(
#    position = "right",
#    limits = rev(c("Canada.original data", "Canada.CI", "Switzerland.original data", "Switzerland.CI")),
#    breaks = NULL,
   name = NULL
  ) +
  scale_color_manual(
    name = "confidence level",
    values = c(
     `80%` = desaturate(darken("#0072B2", .2), .3),
     `95%` = desaturate(lighten("#0072B2", .2), .3),
     `99%` = desaturate(lighten("#0072B2", .4), .3)
    ),
    guide = guide_legend(
      direction = "horizontal",
      title.position = "top",
      label.position = "bottom"
    )
  ) +
  scale_size_manual(
    name = "confidence level",
    values = c(
     `80%` = 1.5,
     `95%` = 1,
     `99%` = 0.5
    ),
    guide = guide_legend(
      direction = "horizontal",
      title.position = "top",
      label.position = "bottom"
    )
  ) +
  coord_cartesian(clip = "off") +
  theme_dviz_hgrid(12, rel_small = 1) +
  theme(
    axis.line.x = element_line(color = "black"),
    axis.ticks.x = element_line(color = "black"),
    axis.title.x = element_text(hjust = 1),
    legend.position = c(1, 0.01),
    legend.justification = c(1, 0),
    legend.key.height = grid::unit(6, "pt"),
    legend.key.width = grid::unit(24, "pt"),
    legend.spacing.x = grid::unit(6, "pt"),
    legend.spacing.y = grid::unit(3, "pt"),
    legend.box.background = element_rect(fill = "white", color = NA),
    legend.box.spacing = grid::unit(0, "pt"),
    legend.title.align = 0.5
  )

```

(ref:chocolate-ratings-contrasts) Mean chocolate flavor ratings for manufacturers from five different countries, relative to the mean rating of U.S. chocolate bars. Canadian chocolate bars are significantly higher rated that U.S. bars. For the other four countries there is no significant difference in mean rating to the U.S. at the 95% confidence level. Data source: Brady Brelinski, Manhattan Chocolate Society

```{r chocolate-ratings-contrasts, fig.asp = 0.5, fig.cap = '(ref:chocolate-ratings-contrasts)'}
# need reference grid for contrasts
fit_rg <- ref_grid(fit)

contrasts_dunnettx <- data.frame(level = c(0.99, 0.95, 0.8)) %>%
  mutate(
    df = map(
      level,
      ~data.frame(confint(contrast(fit_rg, method = "trt.vs.ctrl1"), level = .x))
    )
  ) %>%
  unnest() %>%
  select(level, contrast, estimate, std.error = SE, conf.low = lower.CL, conf.high = upper.CL) %>%
  mutate(
    level = paste0(signif(100*level, 2), "%"),
    contrast = stringr::str_extract(as.character(contrast), "[a-zA-Z]+")
  )

ggplot(contrasts_dunnettx, aes(x = estimate, y = reorder(contrast, estimate))) +
  geom_vline(xintercept = 0, linetype = 2, color = "gray50") +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high, color = level, size = level),
    height = 0
  ) +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high, color = level),
    height = 0.2
  ) +
  geom_point(data = filter(contrasts_dunnettx, level == "80%"), size = 2.5, color = "#D55E00") +
  scale_x_continuous(
    name = "difference in mean rating",
    sec.axis = dup_axis(
      name = NULL,
      breaks = 0,
      labels = "U.S.A. mean rating"
    )
  ) +
  scale_y_discrete(
    name = NULL
  ) +
  scale_color_manual(
    name = "confidence level",
    values = c(
     `80%` = desaturate(darken("#0072B2", .2), .3),
     `95%` = desaturate(lighten("#0072B2", .2), .3),
     `99%` = desaturate(lighten("#0072B2", .4), .3)
    ),
    guide = guide_legend(
      direction = "horizontal",
      title.position = "top",
      label.position = "bottom"
    )
  ) +
  scale_size_manual(
    name = "confidence level",
    values = c(
     `80%` = 1.5,
     `95%` = 1,
     `99%` = 0.5
    ),
    guide = guide_legend(
      direction = "horizontal",
      title.position = "top",
      label.position = "bottom"
    )
  ) +
  coord_cartesian(clip = "off") +
  theme_dviz_hgrid(12, rel_small = 1) +
  theme(
    axis.line.x = element_line(color = "black"),
    axis.line.x.top = element_blank(),
    axis.ticks.x = element_line(color = "black"),
    axis.ticks.x.top = element_line(color = "gray50"),
    #axis.title.x = element_text(hjust = 1),
    legend.position = c(1, 0.01),
    legend.justification = c(1, 0),
    legend.key.height = grid::unit(6, "pt"),
    legend.key.width = grid::unit(24, "pt"),
    legend.spacing.x = grid::unit(6, "pt"),
    legend.spacing.y = grid::unit(3, "pt"),
    legend.box.background = element_rect(fill = "white", color = NA),
    legend.box.spacing = grid::unit(0, "pt"),
    legend.title.align = 0.5
  )

```

```{r confidence-visualizations, fig.width = 8.5, fig.asp = 0.5, fig.cap = '(ref:confidence-visualizations)'}
cacao_smaller <- filter(cacao_small, location != "Switzerland")

fit <- lm(rating ~ location, data = cacao_smaller)
fit_rg <- ref_grid(fit)

contrasts_dunnettx <- data.frame(level = c(0.99, 0.95, 0.8)) %>%
  mutate(
    df = map(
      level,
      ~data.frame(confint(contrast(fit_rg, method = "trt.vs.ctrl1"), level = .x))
    )
  ) %>%
  unnest() %>%
  select(level, contrast, estimate, std.error = SE, conf.low = lower.CL, conf.high = upper.CL) %>%
  mutate(
    level = paste0(signif(100*level, 2), "%"),
    contrast = stringr::str_extract(as.character(contrast), "[a-zA-Z]+")
  )

p1 <- ggplot(contrasts_dunnettx, aes(x = estimate, y = reorder(contrast, estimate))) +
  geom_vline(xintercept = 0, linetype = 2, color = "gray50") +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high, color = level, size = level),
    height = 0
  ) +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high, color = level),
    height = 0.2
  ) +
  geom_point(data = filter(contrasts_dunnettx, level == "80%"), size = 2.5, color = "#D55E00") +
  scale_x_continuous(
    name = "difference in mean rating",
    limits = c(-.65, .47),
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL) +
  scale_color_manual(
    name = "confidence level",
    values = c(
     `80%` = desaturate(darken("#0072B2", .2), .3),
     `95%` = desaturate(lighten("#0072B2", .2), .3),
     `99%` = desaturate(lighten("#0072B2", .4), .3)
    ),
    guide = guide_legend(
      direction = "horizontal",
      title.position = "top",
      label.position = "bottom"
    )
  ) +
  scale_size_manual(
    name = "confidence level",
    values = c(
     `80%` = 1.5,
     `95%` = 1,
     `99%` = 0.5
    ),
    guide = guide_legend(
      direction = "horizontal",
      title.position = "top",
      label.position = "bottom"
    )
  ) +
  coord_cartesian(clip = "off") +
  theme_dviz_hgrid(12, rel_small = 1) +
  theme(
    axis.line.x = element_line(color = "black"),
    axis.line.x.top = element_blank(),
    axis.ticks.x = element_line(color = "black"),
    axis.ticks.x.top = element_line(color = "gray50"),
    #axis.title.x = element_text(hjust = 1),
    legend.position = "none",
    #legend.position = c(1, .01),
    legend.justification = c(1, 0),
    legend.key.height = grid::unit(6, "pt"),
    legend.key.width = grid::unit(24, "pt"),
    legend.spacing.x = grid::unit(6, "pt"),
    legend.spacing.y = grid::unit(3, "pt"),
    legend.box.background = element_rect(fill = "white", color = NA),
    legend.box.spacing = grid::unit(0, "pt"),
    legend.title.align = 0.5
  )

p2 <- ggplot(filter(contrasts_dunnettx, level == "95%"),
             aes(x = estimate, y = reorder(contrast, estimate))) +
  geom_vline(xintercept = 0, linetype = 2, color = "gray50") +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    height = 0.2
  ) +
  geom_point(size = 2.5, color = "#D55E00") +
  scale_x_continuous(
    name = "difference in mean rating",
    limits = c(-.65, .47),
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL) +
  coord_cartesian(clip = "off") +
  theme_dviz_hgrid(12, rel_small = 1) +
  theme(
    axis.line.x = element_line(color = "black"),
    axis.line.x.top = element_blank(),
    axis.ticks.x = element_line(color = "black"),
    axis.ticks.x.top = element_line(color = "gray50")
  )

p3 <- ggplot(filter(contrasts_dunnettx, level == "95%"),
             aes(x = estimate, y = reorder(contrast, estimate))) +
  geom_vline(xintercept = 0, linetype = 2, color = "gray50") +
  stat_confidence_density(
    aes(moe = conf.high-estimate, fill = stat(ndensity)),
    height = 0.7,
    confidence = 0.95,
    alpha = NA
  ) +
  geom_segment(
    aes(
      x = estimate, xend = estimate,
      y = as.integer(reorder(contrast, estimate)) - 0.35,
      yend = as.integer(reorder(contrast, estimate)) + 0.35
    ),
    size = 2, color = "#D55E00"
  ) +
  scale_fill_gradient(low = "#81A7D600", high = "#345A7FD0") +
  scale_x_continuous(
    name = "difference in mean rating",
    limits = c(-.65, .47),
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL) +
  coord_cartesian(clip = "off") +
  theme_dviz_hgrid(12, rel_small = 1) +
  theme(
    axis.line.x = element_line(color = "black"),
    axis.line.x.top = element_blank(),
    axis.ticks.x = element_line(color = "black"),
    axis.ticks.x.top = element_line(color = "gray50")
  )


p4 <- ggplot(filter(contrasts_dunnettx, level == "95%"),
             aes(x = estimate, y = reorder(contrast, estimate))) +
  geom_vline(xintercept = 0, linetype = 2, color = "gray50") +
  stat_confidence_density(
    aes(moe = conf.high-estimate, height = stat(density)),
    geom = "ridgeline",
    confidence = 0.95,
    scale = 0.2,
    min_height = 0.001,
    alpha = 0.7,
    fill = "#81A7D6",
    size = 0.4
  ) +
  geom_segment(
    aes(
      x = estimate, xend = estimate,
      y = as.integer(reorder(contrast, estimate)) - 0.2,
      yend = as.integer(reorder(contrast, estimate)) + 0.5
    ),
    size = 2, color = "#D55E00"
  ) +
  scale_fill_gradient(low = "#81A7D600", high = "#345A7FD0") +
  scale_x_continuous(
    name = "difference in mean rating",
    limits = c(-.65, .47),
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL) +
  coord_cartesian(clip = "off") +
  theme_dviz_hgrid(12, rel_small = 1) +
  theme(
    axis.line.x = element_line(color = "black"),
    axis.line.x.top = element_blank(),
    axis.ticks.x = element_line(color = "black"),
    axis.ticks.x.top = element_line(color = "gray50"),
    axis.text.y = element_text(vjust = 0)
  )

plot_grid(
  p1, NULL, p2,
  NULL, NULL, NULL,
  p3, NULL, p4,
  labels = c("a", "", "b", "", "", "", "c", "", "d"),
  rel_widths = c(1, .02, 1),
  rel_heights = c(1, .02, 1)
)
```


(ref:butterfat-bars) *Update: Density estimates of the butterfat percentage in the milk of four cattle breeds.* Data Source: Canadian Record of Performance for Purebred Dairy Cattle

```{r butterfat-bars, fig.width = 5, fig.cap = '(ref:butterfat-bars)'}
cows %>% filter(breed != "Canadian") %>%
  mutate(breed = as.character(breed)) %>%
  group_by(breed) %>%
  summarize(
    mean = mean(butterfat),
    se = sd(butterfat)/sqrt(n())
  ) %>%
  mutate(breed = fct_reorder(factor(breed), desc(mean)))-> cow_means

ggplot(cow_means, aes(x = breed, y = mean, ymin = mean - 1.96*se, ymax = mean + 1.96*se)) +
  geom_col(fill = "#56B4E9", alpha = 0.7, width = 0.7) +
  geom_linerange(size = 0.75) +
  scale_x_discrete(
    labels = c("Jersey", "Guernsey", "Ayrshire", "Holstein-\nFriesian"),
    name = NULL
  ) +
  scale_y_continuous(
    name = "mean butterfat contents", expand = c(0, 0),
    labels = scales::percent_format(accuracy = 1, scale = 1)
  ) +
  coord_cartesian(clip = "off") +
  theme_dviz_hgrid() +
  theme(axis.line.x = element_blank())
```

(ref:median-age-income) Median income versus median age for 67 counties in Pennsylvania. Error bars represent 90% confidence intervals. Data source: 2015 Five-Year American Community Survey

```{r median-age-income, fig.width = 5, fig.asp = 3/4, fig.cap = '(ref:median-age-income)'}
filter(US_income_age, state == "Pennsylvania") %>%
  ggplot(aes(age, income)) + 
  geom_errorbarh(
    aes(xmin = age - age.moe, xmax = age + age.moe),
    height = 0,
    color = "#0072B2"
  ) +
  geom_errorbar(
    aes(ymin = income - income.moe, ymax = income + income.moe),
    width = 0,
    color = "#0072B2"
  ) +
  geom_point(color = "#0072B2", size = 0.7) +
  scale_x_continuous(name = "median age") +
  scale_y_continuous(
    name = "median income",
    labels = scales::dollar
  ) +
  theme_dviz_open()
```

```{block type='rmdtip', echo=TRUE}
Whenever you visualize uncertainty with error bars, you must specify what quantity and/or confidence level the error bars represent.
```


```{r fig.width = 5.5, fig.asp = 3/4}
blue_jays_male <- filter(blue_jays, KnownSex == "M")

fit <- gam(Head ~ Mass, data = blue_jays_male, method = "REML")

newdata <- data.frame(Mass = seq(59, 82, length.out = 100))

sample_df <- sample_outcomes(fit, newdata, 15, unconditional = TRUE)
ci_df <- confidence_band(fit, newdata, unconditional = TRUE)

ggplot(blue_jays_male, aes(Mass, Head)) + 
  geom_ribbon(data = ci_df, aes(ymin = lo, ymax = hi), fill="grey70", color = NA, alpha = 1/2) +
  geom_point(color = "grey60", size = 1.5) +
  #geom_line(data = sample_df, aes(group = .draw), color = "#0072B2", size = 0.3) +
  geom_line(data = ci_df, color = "#0072B2", size = 1) +
  scale_x_continuous(
    limits = c(59, 82),
    expand = c(0, 0),
    name = "body mass (g)") +
  scale_y_continuous(
    limits = c(52, 61),
    expand = c(0, 0),
    name = "head length (mm)"
  ) +
  theme_dviz_open()
```

```{r fig.width = 5.5, fig.asp = 3/4}

ggplot(blue_jays_male, aes(Mass, Head)) + 
  geom_ribbon(data = ci_df, aes(ymin = lo, ymax = hi), fill="grey70", color = NA, alpha = 1/2) +
  geom_point(color = "grey60", size = 1.5) +
  geom_line(data = sample_df, aes(group = .draw), color = "#0072B2", size = 0.3) +
  #geom_line(data = ci_df, color = "#0072B2", size = 1) +
  scale_x_continuous(
    limits = c(59, 82),
    expand = c(0, 0),
    name = "body mass (g)") +
  scale_y_continuous(
    limits = c(52, 61),
    expand = c(0, 0),
    name = "head length (mm)"
  ) +
  theme_dviz_open()

```
**Graded confidence band**

```{r fig.width = 5.5, fig.asp = 3/4}
ci_99_df <- confidence_band(fit, newdata, level = 0.99, unconditional = TRUE)
ci_80_df <- confidence_band(fit, newdata, level = 0.80, unconditional = TRUE)

ggplot(blue_jays_male, aes(Mass, Head)) + 
  geom_ribbon(
    data = ci_99_df,
    aes(ymin = lo, ymax = hi, fill = "99%"),
    color = NA
  ) +
  geom_ribbon(
    data = ci_df,
    aes(ymin = lo, ymax = hi, fill = "95%"),
    color = NA
  ) +
  geom_ribbon(
    data = ci_80_df,
    aes(ymin = lo, ymax = hi, fill = "80%"),
    color = NA
  ) +
  geom_point(color = "grey40", size = 1.5) +
  geom_line(data = ci_df, color = "#D55E00", size = 1.5) +
  scale_fill_manual(
    name = "confidence level",
    values = c(
     `80%` = desaturate(lighten("#0072B2", .4), .3),
     `95%` = desaturate(lighten("#0072B2", .6), .3),
     `99%` = desaturate(lighten("#0072B2", .8), .3)
    ),
    guide = guide_legend(
      direction = "horizontal",
      title.position = "top",
      label.position = "bottom"
    )
  ) +
  scale_x_continuous(
    limits = c(59, 82),
    expand = c(0, 0),
    name = "body mass (g)"
  ) +
  scale_y_continuous(
    limits = c(52, 61),
    expand = c(0, 0),
    name = "head length (mm)"
  ) +
  theme_dviz_open() +
  theme(
    legend.position = c(1, .01),
    legend.justification = c(1, 0),
    legend.key.height = grid::unit(12, "pt"),
    legend.key.width = grid::unit(24, "pt"),
    legend.spacing.x = grid::unit(6, "pt"),
    legend.spacing.y = grid::unit(3, "pt"),
    legend.box.background = element_rect(fill = "white", color = NA),
    legend.box.spacing = grid::unit(0, "pt"),
    legend.title = element_text(size = 12, hjust = 0.5)
  )
```

```{r tank-capacity-uncertain, eval = FALSE, fig.width = 8.5, fig.asp = 3/8, fig.cap='(ref:tank-capacity-uncertain)'}
cars93 <- MASS::Cars93

set.seed(8692282)

fit <- gam(Fuel.tank.capacity ~ s(Price, k = 5, bs = 'cr'), data = cars93)
#fit <- gam(Fuel.tank.capacity ~ s(Price, k = 6, bs = 'gp'), data=cars93)
#fit <- gam(Fuel.tank.capacity ~ s(Price), data = cars93, method = "REML")

newdata <- data.frame(Price = seq(min(cars93$Price), max(cars93$Price), length.out = 100))
sample_df <- sample_outcomes(fit, newdata, 10, unconditional = FALSE)
ci_df <- confidence_band(fit, newdata, unconditional = FALSE)

cars_base <- ggplot(cars93, aes(x = Price, y = Fuel.tank.capacity)) + 
  scale_x_continuous(
    name = "price (USD)",
    breaks = c(20, 40, 60),
    labels = c("$20,000", "$40,000", "$60,000")
  ) +
  scale_y_continuous(name = "fuel-tank capacity\n(US gallons)") +
  theme_minimal_grid(12)

p1 <- cars_base +
  geom_ribbon(data = ci_df, aes(ymin = lo, ymax = hi), fill="grey70", color = NA, alpha = 1/2) +
  geom_point(color = "grey60") +
  geom_line(data = ci_df, color = "#0072B2", size = 1)

p2 <- cars_base +
  geom_ribbon(data = ci_df, aes(ymin = lo, ymax = hi), fill="grey70", color = NA, alpha = 1/2) +
  geom_point(color = "grey60") +
  geom_line(data = sample_df, aes(group = .draw), color = "#0072B2", size = 0.3)

plot_grid(
  p1, p2, align = 'hv',
  labels = 'auto'
)

```

(ref:mpg-uncertain) Fuel efficiency versus displacement, for 32 cars (1973–74 models). Cubic regression splines with 5 knots. **Complete caption.** Data source: Motor Trend, 1974.

```{r mpg-uncertain, fig.width = 8.5, fig.asp = 3/8, fig.cap='(ref:mpg-uncertain)'}
set.seed(8692282)

fit <- gam(mpg ~ s(disp, bs = 'cr', k = 5), data = mtcars, method = "REML")
#fit <- gam(mpg ~ s(disp, bs = 'ts', k = 6), data = mtcars, method = "REML")

newdata <- data.frame(disp = seq(min(mtcars$disp), max(mtcars$disp), length.out = 100))
sample_df <- sample_outcomes(fit, newdata, 10, unconditional = TRUE)
ci_df <- confidence_band(fit, newdata, unconditional = TRUE)

cars_base <- ggplot(mtcars, aes(x = disp, y = mpg)) + 
  scale_x_continuous(
    name = "displacement (cu. in.)",
    expand = c(0.01, 0)
  ) +
  ylab("fuel efficiency (mpg)") +
  theme_dviz_grid(12)

p1 <- cars_base +
  geom_ribbon(data = ci_df, aes(ymin = lo, ymax = hi), fill="grey70", color = NA, alpha = 1/2) +
  geom_point(color = "grey60") +
  geom_line(data = ci_df, color = "#0072B2", size = 1)

p2 <- cars_base +
  geom_ribbon(data = ci_df, aes(ymin = lo, ymax = hi), fill="grey70", color = NA, alpha = 1/2) +
  geom_point(color = "grey60") +
  geom_line(data = sample_df, aes(group = .draw), color = "#0072B2", size = 0.3)

plot_grid(
  p1, NULL, p2, align = 'hv',
  nrow = 1,
  labels = c("a", "", "b"),
  rel_widths = c(1, .02, 1)
)

```


## Methods of uncertainty estimates

Explain confidence via hypothesis testing: 95% confidence means less than 5% prob. of seeing result we saw if null were true.

Table: (\#tab:estimation-frameworks) Comparisons of different approaches to parameter and uncertainty estimation.

------------------------------------------------------------------------------------------
approach       interpretability    computational       complexity of     assumptions on
                of estimates         efficiency        modeling setup       data set
------------ ------------------ ------------------- ------------------- ------------------
Bayesian          high                 low                 high                few

frequentist       moderate              high              moderate            several
parametric  

bootstrap           low                moderate             low                none
------------------------------------------------------------------------------------------


## Hypothetical outcomes plots

Hypothetical outcomes plots: https://medium.com/hci-design-at-uw/hypothetical-outcomes-plots-experiencing-the-uncertain-b9ea60d7c740

## Value-suppressing uncertainty palette

