```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")

library(dplyr)
library(tidyr)
```


# Visualizing distributions: Histograms and density plots {#histograms-density-plots}

We frequently encounter the situation where we would like to understand how a particular variable is distributed in a dataset. To give a concrete example, consider the passengers of the ship Titanic, which sank on April 15, 2012. There were approximately 1317 passengers on the Titanic (not counting crew), and we have reported ages for 756 of them. We might want to know how many passengers of what ages there were on the Titanic, i.e., how many children, young adults, middle-aged people, seniors, and so on. We call the relative proportions of different ages among the passengers the *age distribution* of the passengers.

## Visualizing a single distribution

We can obtain a sense of the age distribution among the passengers by grouping all passengers into bins with comparable ages and then counting the number of passengers in each bin. This procedure results in a table such as Table \@ref(tab:titanic-ages). 
```{r titanic-ages, message=FALSE, warning=FALSE}
library(readr)
df <- read_csv("datasets/Titanic.csv")
titanic <- na.omit(data.frame(age = df$Age, sex = df$Sex, survived = df$Survived))
age_counts <- hist(titanic$age, breaks=(0:15)*5+.01, plot = FALSE)$counts
age_hist <- data.frame(`age range` = c("0--5", "6--10", "11--15", "16--20", "21--25", "26--30", "31--35", "36--40", "41--45", "46--50", "51--55", "56--60", "61--65", "66--70", "71--75"),
                       count = age_counts,
                       check.names = FALSE)

knitr::kable(
  list(
    age_hist[1:6,], age_hist[7:12,], age_hist[13:15,]
  ),
  caption = 'Numbers of passenger with known age on the Titanic.', booktabs = TRUE,
  row.names = FALSE
)

```

We can visualize this table by drawing filled rectangles whose heights correspond to the counts and whose widths correspond to the width of the age bins (Figure \@ref(fig:titanic-ages-hist1)). Such a visualization is called a histogram.

(ref:titanic-ages-hist1) Histogram of the ages of Titanic passengers.

```{r titanic-ages-hist1, fig.cap='(ref:titanic-ages-hist1)'}
age_hist <- cbind(age_hist, age = (1:15)*5-2.5)
h1 <- ggplot(age_hist, aes(x = age, y = count)) + geom_col(width = 4.7, fill = "#56B4E9")  + 
  scale_y_continuous(expand = c(0, 0), breaks = 25*(0:5)) +
  scale_x_continuous(limits = c(0, 75), expand = c(0, 0)) +
  theme_dviz_hgrid()
h1
```

Because histograms are generated by binning the data, their exact visual appearance depends on the choice of the bin width. Most visualization programs that generate histograms will choose a bin width by default, but chances are that bin width is not the most appropriate one for any histogram you may want to make. It is therefore critical to always try different bin widths to verify that the resulting histogram reflects the underlying data accurately. In general, if the bin width is too small, then the histogram becomes overly peaky and visually busy and the main trends in the data may be obscured. On the other hand, if the bin width is too large, then smaller features in the distribution of the data may disappear.

For the age distribution of Titanic passengers, we can see that a bin width of one year is too small and a bin width of fifteen years is too large, whereas bin widths between three to five years work fine (Figure \@ref(fig:titanic-ages-hist-grid)).

(ref:titanic-ages-hist-grid) Histograms depend on the chosen bin width. Here, the same age distribution of Titanic passengers is shown with four different bin widths: (a) one year; (b) three years; (c) five years; (d) fifteen years.

```{r titanic-ages-hist-grid, fig.width=8.5, fig.cap='(ref:titanic-ages-hist-grid)'}
age_hist_1 <- data.frame(age = (1:75)-0.5, 
                         count = hist(titanic$age, breaks=(0:75) + .01, plot = FALSE)$counts)

age_hist_3 <- data.frame(age = (1:25)*3-1.5, 
                         count = hist(titanic$age, breaks=(0:25)*3 + .01, plot = FALSE)$counts)

age_hist_15 <- data.frame(age = (1:5)*15-7.5, 
                         count = hist(titanic$age, breaks=(0:5)*15 + .01, plot = FALSE)$counts)

h2 <- ggplot(age_hist_1, aes(x = age, y = count)) + geom_col(width = .85, fill = "#56B4E9")  + 
  scale_y_continuous(expand = c(0, 0), breaks = 10*(0:5)) +
  scale_x_continuous(limits = c(0, 75), expand = c(0, 0)) +
  theme_dviz_hgrid(12)

h3 <- ggplot(age_hist_3, aes(x = age, y = count)) + geom_col(width = 2.75, fill = "#56B4E9")  + 
  scale_y_continuous(expand = c(0, 0), breaks = 25*(0:5)) +
  scale_x_continuous(limits = c(0, 75), expand = c(0, 0)) +
  theme_dviz_hgrid(12)

h4 <- ggplot(age_hist_15, aes(x = age, y = count)) + geom_col(width = 14.5, fill = "#56B4E9")  + 
  scale_y_continuous(expand = c(0, 0), breaks = 100*(0:4)) +
  scale_x_continuous(limits = c(0, 75), expand = c(0, 0)) +
  theme_dviz_hgrid(12)

plot_grid(h2, h3, h1 + theme_dviz_hgrid(12), h4, align = 'hv',
          labels = 'auto')

```

```{block type='rmdtip', echo=TRUE}
When making a histogram, always explore multiple bin widths.
```

Histograms have been a popular visualization option since at least the 18th Century, and they are easily generated by hand. More recently, as extensive computing power has become available in everyday devices such as laptops and cell phones, we see them increasingly replaced by density plots. **complete** 


(ref:titanic-ages-dens1) **Caption needed.**

```{r titanic-ages-dens1, fig.cap='(ref:titanic-ages-dens1)'}
ggplot(titanic, aes(x=age)) + 
  geom_density(fill = "#56B4E9", bw = 2, kernel = "gaussian") + 
  scale_y_continuous(limits = c(0, 0.04), expand = c(0.01, 0), name = "density") +
  scale_x_continuous(limits = c(0, 75), expand = c(0.01, 0)) +
  theme_dviz_grid()
```

*Discuss scaling, different choices for kernel and bandwith, problems with density estimates going outside the allowed data bounds (e.g., negative).*


(ref:titanic-ages-dens-grid) Kernel density estimates depend on the chosen kernel and bandwidth. Here, the same age distribution of Titanic passengers is shown for four different combinations of these parameters: (a) Gaussian kernel, bandwidth = 0.5; (b) Gaussian kernel, bandwidth = 2; (c) Gaussian kernel, bandwidth = 5; (d) Rectangular kernel, bandwidth = 2.

```{r titanic-ages-dens-grid, fig.cap='(ref:titanic-ages-dens-grid)'}
pdens1 <- ggplot(titanic, aes(x=age)) + 
  geom_density(fill = "#56B4E9", bw = .5, kernel = "gaussian") + 
  scale_y_continuous(limits = c(0, 0.048), expand = c(0.01, 0), name = "density") +
  scale_x_continuous(limits = c(0, 75), expand = c(0.01, 0)) +
  theme_dviz_grid(12)

pdens2 <- ggplot(titanic, aes(x=age)) + 
  geom_density(fill = "#56B4E9", bw = 2, kernel = "gaussian") + 
  scale_y_continuous(limits = c(0, 0.048), expand = c(0.01, 0), name = "density") +
  scale_x_continuous(limits = c(0, 75), expand = c(0.01, 0)) +
  theme_dviz_grid(12)

pdens3 <- ggplot(titanic, aes(x=age)) + 
  geom_density(fill = "#56B4E9", bw = 5, kernel = "gaussian") + 
  scale_y_continuous(limits = c(0, 0.048), expand = c(0.01, 0), name = "density") +
  scale_x_continuous(limits = c(0, 75), expand = c(0.01, 0)) +
  theme_dviz_grid(12)

pdens4 <- ggplot(titanic, aes(x=age)) + 
  geom_density(fill = "#56B4E9", bw = 2, kernel = "rectangular") + 
  scale_y_continuous(limits = c(0, 0.048), expand = c(0.01, 0), name = "density") +
  scale_x_continuous(limits = c(0, 75), expand = c(0.01, 0)) +
  theme_dviz_grid(12)

plot_grid(pdens1, pdens2, pdens3, pdens4, align = 'hv',
          labels = 'auto')
```

(ref:titanic-ages-dens-negative) **Caption needed.**

```{r titanic-ages-dens-negative, fig.cap='(ref:titanic-ages-dens-negative)'}
ggplot(titanic, aes(x=age)) + 
  geom_density(fill = "#56B4E9") + 
  scale_y_continuous(limits = c(0, 0.04), expand = c(0.01, 0), name = "density") +
  scale_x_continuous(limits = c(-10, 80), expand = c(0.01, 0)) +
  theme_dviz_grid()
```


## Visualizing multiple distributions at the same time

(ref:titanic-age-stacked-hist) **Caption needed.**

```{r titanic-age-stacked-hist, fig.cap='(ref:titanic-age-stacked-hist)'}
data.frame(age = (1:25)*3-1.5, 
           male = hist(filter(titanic, sex == "male")$age, breaks=(0:25)*3 + .01, plot = FALSE)$counts,
           female = hist(filter(titanic, sex == "female")$age, breaks=(0:25)*3 + .01, plot = FALSE)$counts) %>%
  gather(gender, count, -age) -> gender_counts

gender_counts$gender <- factor(gender_counts$gender, levels = c("female", "male"))

p_hist_stacked <- ggplot(gender_counts, aes(x=age, y=count, fill = gender)) + 
  geom_col(position = "stack") +
  scale_x_continuous(limits = c(0, 75), expand = c(0, 0)) +
  scale_y_continuous(name = "count", expand = c(0, 0)) +
  scale_fill_manual(values = c("#D55E00", "#0072B2")) +
  theme_dviz_hgrid() +
  theme(legend.position = c(.9, .92),
        legend.justification = c("right", "top"),
        legend.box.background = element_rect(fill = "white", color = "white"))

stamp_bad(p_hist_stacked)
```
There are two key problems with this type of visualization: First, it is never entirely clear whether all bars are meant to start at zero or exactly where the color changes. Are there about 25 females of age 18--20 or are there almost 80? (The former is the case.) Second, the bar heights for the female counts cannot be directly compared, because the bars all start at a different height.

We could try to address these problems by having all bars start at zero and making the bars partially transparent (Figure \@ref(fig:titanic-age-overlapping-hist)).

(ref:titanic-age-overlapping-hist) **Caption needed.**

```{r titanic-age-overlapping-hist, fig.cap='(ref:titanic-age-overlapping-hist)'}
p_hist_overlapped <- ggplot(gender_counts, aes(x=age, y=count, fill = gender)) + 
  geom_col(position = "identity", alpha = 0.7) +
  scale_x_continuous(limits = c(0, 75), expand = c(0, 0)) +
  scale_y_continuous(name = "count", expand = c(0, 0)) +
  scale_fill_manual(values = c("#D55E00", "#0072B2")) +
  theme_dviz_hgrid() +
  theme(legend.position = c(.9, .92),
        legend.justification = c("right", "top"),
        legend.box.background = element_rect(fill = "white", color = "white"))

stamp_bad(p_hist_overlapped)
```

However, this approach generates new problems. Now it appears that there are actually three different groups, not just two, and we're still not entirely sure where each bar starts and ends.


(ref:titanic-age-overlapping-dens) **Caption needed.** *Mention special scaling here.*

```{r titanic-age-overlapping-dens, fig.cap='(ref:titanic-age-overlapping-dens)'}
titanic2 <- titanic
titanic2$sex <- factor(titanic2$sex, levels = c("male", "female"))
p_dens_overlapped <- ggplot(titanic2, aes(x=age, y=..count.., fill = sex)) + 
  geom_density(bw = 2, alpha = 0.7) +
  scale_x_continuous(limits = c(0, 75), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, 20), name = "scaled density", expand = c(0, 0)) +
  scale_fill_manual(values = c("#0072B2", "#D55E00"), name = "gender") +
  theme_dviz_hgrid() +
  theme(legend.position = c(.9, .92),
        legend.justification = c("right", "top"),
        legend.box.background = element_rect(fill = "white", color = "white"))

stamp_good(p_dens_overlapped)
```

*Discuss scaling*


When we want to visualize exactly two distributions, we can make two separate histograms, rotate them by 90 degrees, and have the bars in one histogram point into the opposite direction of the other. This trick is commonly employed when visualizing age distributions, and the resulting plot is usually called an *age pyramid* (Figure \@ref(fig:titanic-age-pyramic)).

(ref:titanic-age-pyramid) The age distributions of male and female Titanic passengers visualized as an age pyramid.

```{r titanic-age-pyramid, fig.cap='(ref:titanic-age-pyramid)'}
ggplot(gender_counts, aes(x=age, y=ifelse(gender == "male",-1, 1)*count, fill = gender)) + 
  geom_col() +
  scale_x_continuous(limits = c(0, 75), expand = c(0, 0)) +
  scale_y_continuous(name = "count", breaks = 20*(-2:1)) +
  scale_fill_manual(values = c("#D55E00", "#0072B2"), guide = "none") +
  draw_text(x = 70, y = -35, "male") +
  draw_text(x = 70, y = 25, "female") +
  coord_flip() +
  theme_dviz_grid() +
  theme(axis.title.x = element_text(hjust = 0.61))


```

Importantly, this trick does not work when there are more than two distributions we want to visualize at the same time. For example, to visualize the length distributions of sepals for three different iris species, density plots are by far the best choice (Figure \@ref(fig:iris-densities)).

(ref:iris-densities) **Caption needed.**

```{r iris-densities, message=FALSE, warning=FALSE, fig.cap='(ref:iris-densities)'}
# compute densities for sepal lengths
iris_dens <- group_by(iris, Species) %>%
  do(ggplot2:::compute_density(.$Sepal.Length, NULL)) %>%
  rename(Sepal.Length = x)

# get the maximum values
iris_max <- filter(iris_dens, density == max(density)) %>%
  ungroup() %>%
  mutate(hjust = c(0, 0.4, 0),
         vjust = c(1, 0, 1),
         nudge_x = c(0.11, 0, 0.2),
         nudge_y = c(-0.02, 0.02, -0.02) 
        )

iris_p <- ggplot(iris_dens, aes(x = Sepal.Length, y = density, fill = Species)) + 
  geom_density(stat = "identity", alpha = 0.7) +
  geom_text(data = iris_max, aes(label = Species, hjust = hjust, vjust = vjust,
                                 x = Sepal.Length + nudge_x, 
                                 y = density + nudge_y), size = 12/.pt) +
  scale_fill_manual(values = c("#56B4E9", "#E69F00", "#009E73"),
                    breaks = c("virginica", "versicolor", "setosa"),
                    guide = "none") +
  scale_x_continuous(expand = c(0, 0), name = "sepal length") +
  scale_y_continuous(limits = c(0, 1.5), expand = c(0, 0)) +
  theme_dviz_hgrid()
  
iris_p
```

```{block type='rmdtip', echo=TRUE}
To visualize several distributions at once, kernel density plots will generally work better than histograms.
```

