```{r echo = FALSE, message = FALSE, warning = FALSE}
# run setup script
source("_common.R")

library(forcats)
library(patchwork)
library(lubridate)
library(mgcv)
library(mvtnorm)
library(rlang)
library(tidyr)
library(purrr)
library(broom)
library(emmeans)
```

# Visualizing uncertainty {#visualizing-uncertainty}

*Chapter under construction. Tentative outline:*

- *Error bars and confidence bands*
- *Distributions*
- *Hypothetical outcomes plots*
- *Value suppression*


## Error bars and confidence bands

(ref:cocoa-data-vs-CI) *Figure caption.*

```{r cocoa-data-vs-CI, fig.asp = 0.4, warning = FALSE, message = FALSE, fig.cap = '(ref:cocoa-data-vs-CI)'}
cacao %>% 
  filter(location == "Canada") -> cacao_single

fit <- lm(rating ~ 1, data = cacao_single)

CI_df <- data.frame(type = c(0.8, 0.95, 0.99)) %>%
  mutate(df = map(type, ~tidy(emmeans(fit, ~ 1, options = list(level = .x))))) %>%
  unnest() %>%
  select(type, estimate, std.error, conf.low, conf.high) %>%
  mutate(type = paste0(signif(100*type, 2), "% confidence interval"))

CI_df <- rbind(
  CI_df,
  data.frame(
    type = "standard error",
    estimate = CI_df$estimate[1],
    std.error = CI_df$std.error[1],
    conf.low = CI_df$estimate[1] - CI_df$std.error[1],
    conf.high = CI_df$estimate[1] + CI_df$std.error[1]
  ),
  data.frame(
    type = "standard deviation",
    estimate = mean(cacao_single$rating),
    std.error = CI_df$std.error[1],
    conf.low = mean(cacao_single$rating) - sd(cacao_single$rating),
    conf.high = mean(cacao_single$rating) + sd(cacao_single$rating)
  ),
  data.frame(
    type = "original data", estimate = mean(cacao_single$rating), std.error = NA,
    conf.low = NA, conf.high = max(cacao_single$rating)
  )
) %>%
  mutate(
    type = fct_relevel(factor(type), "original data", "standard deviation", "standard error"),
    label = case_when(
      type == "original data" ~ NA_character_,
      TRUE ~ as.character(type) #paste0("mean +/- ", type)
    )
  )

label_x <- filter(CI_df, type == "standard deviation")$conf.high + 0.04

ggplot(CI_df, aes(estimate, type)) + 
  geom_point(
    data = cacao_single, aes(rating, "original data"), 
    position = position_jitter(height = 0.6, width = 0.02),
    color = "#0072B2",
    size = 0.2
  ) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, na.rm = TRUE) +
  geom_point(size = 2, color = "#D55E00") +
  geom_label(
    aes(label_x, label = label), hjust = 0, nudge_y = 0.01, na.rm = TRUE,
    family = dviz_font_family,
    size = 12/.pt,
    label.size = 0
  ) +
  geom_label(
    data = filter(CI_df, type == "original data"),
    aes(conf.high + 0.06, label = type), hjust = 0, nudge_y = 0.01,
    family = dviz_font_family,
    size = 12/.pt,
    label.size = 0
  ) +
  geom_text(
    data = filter(CI_df, type == "original data"),
    aes(estimate, label = "mean"), hjust = 0.2, vjust = 0, nudge_y = 0.2,
    family = dviz_font_family,
    size = 12/.pt
  ) +
  scale_x_continuous(
    position = "bottom",
    limits = c(1.95, 4.1),
    expand = c(0, 0),
    name = "chocolate flavor rating"
  ) +
  scale_y_discrete(
    position = "right",
    name = NULL,
    limits = rev(levels(CI_df$type)),
    breaks = NULL
  ) +
  coord_cartesian(clip = "off") +
  theme_dviz_vgrid(12, rel_small = 1) +
  theme(
    plot.margin = margin(3, 73, 3, 0),
    axis.line.x = element_line(),
    axis.ticks.x = element_line(color = "black"),
    axis.title.x = element_text(hjust = 1)
  )

```

```{r cocoa-CI-vs-n, fig.asp = 0.4, warning = FALSE, message = FALSE, fig.cap = '(ref:cocoa-CI-vs-n)'}
cacao_CA <- filter(cacao, location == "Canada")
cacao_CH <- filter(cacao, location == "Switzerland")

fit_CA <- lm(rating ~ 1, data = cacao_CA)
fit_CH <- lm(rating ~ 1, data = cacao_CH)

CI_CA <- data.frame(level = c(0.8, 0.95, 0.99)) %>%
  mutate(df = map(level, ~tidy(emmeans(fit_CA, ~ 1, options = list(level = .x))))) %>%
  unnest() %>%
  select(level, estimate, conf.low, conf.high) %>%
  mutate(
    level = paste0(signif(100*level, 2), "%"),
    type = "CI",
    location = "Canada"
  )

CI_CH <- data.frame(level = c(0.8, 0.95, 0.99)) %>%
  mutate(df = map(level, ~tidy(emmeans(fit_CH, ~ 1, options = list(level = .x))))) %>%
  unnest() %>%
  select(level, estimate, conf.low, conf.high) %>%
  mutate(
    level = paste0(signif(100*level, 2), "%"),
    type = "CI",
    location = "Switzerland"
  )

CI_df <- rbind(
  CI_CA,
  data.frame(
    level = NA,
    estimate = mean(cacao_CA$rating),
    conf.low = mean(cacao_CA$rating) - sd(cacao_CA$rating),
    conf.high = mean(cacao_CA$rating) + sd(cacao_CA$rating),
    type = "original data",
    location = "Canada"
  ),
  CI_CH,
  data.frame(
    level = NA,
    estimate = mean(cacao_CH$rating),
    conf.low = mean(cacao_CH$rating) - sd(cacao_CH$rating),
    conf.high = mean(cacao_CH$rating) + sd(cacao_CH$rating),
    type = "original data",
    location = "Switzerland"
  )
)

#label_x <- filter(CI_df, type == "standard deviation")$conf.high + 0.04

ggplot(CI_df, aes(estimate, interaction(location, type))) + 
  geom_point(
    data = rbind(cacao_CA, cacao_CH), 
    aes(rating, interaction(location, "original data")), 
    position = position_jitter(height = 0.6, width = 0.02),
    color = "#0072B2",
    size = 0.2
  ) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, na.rm = TRUE) +
  geom_point(size = 2, color = "#D55E00") +
#  geom_label(
#    aes(label_x, label = label), hjust = 0, nudge_y = 0.01, na.rm = TRUE,
#    family = dviz_font_family,
#    size = 12/.pt,
#    label.size = 0
#  ) +
#  geom_label(
#    data = filter(CI_df, type == "original data"),
#    aes(conf.high + 0.06, label = type), hjust = 0, nudge_y = 0.01,
#    family = dviz_font_family,
#    size = 12/.pt,
#    label.size = 0
#  ) +
  scale_x_continuous(
    position = "bottom",
    limits = c(1.95, 4.1),
    expand = c(0, 0),
    name = "chocolate flavor rating"
  ) +
  scale_y_discrete(
    position = "right",
    name = NULL,
    limits = rev(c("Canada.original data", "Canada.CI", "Switzerland.original data", "Switzerland.CI")),
    breaks = NULL
  ) +
  coord_cartesian(clip = "off") +
  theme_dviz_vgrid(12, rel_small = 1) +
  theme(
    plot.margin = margin(3, 73, 3, 0),
    axis.line.x = element_line(),
    axis.ticks.x = element_line(color = "black"),
    axis.title.x = element_text(hjust = 1)
  )

```

```{r fig.asp = 1.2, eval = FALSE}
n <- c(10, 30, 100, 300, 1000)

df <- data.frame(n = unlist(lapply(n, function(x) rep(x, x))),
                 x = c(rnorm(sum(n))))

df %>% group_by(n) %>%
  summarize(mean = mean(x),
            se = sd(x)/sqrt(n())) -> df_mean

p1 <- ggplot(df, aes(x = factor(n), y = x)) + 
  geom_point(size = 0.5, position = position_jitter(width = 0.3)) +
  scale_x_discrete(breaks = NULL, name = NULL) +
  scale_y_continuous(breaks = c(-2, 0, 2)) +
  theme_dviz_hgrid() +
  theme(plot.margin = margin(3, 0, 14, 0))
p2 <- ggplot(df, aes(x = factor(n), y = x)) + 
  geom_boxplot(fill = "gray90") +
  scale_x_discrete(breaks = NULL, name = NULL) +
  scale_y_continuous(breaks = c(-2, 0, 2)) +
  theme_dviz_hgrid() +
  theme(plot.margin = margin(3, 0, 14, 0))
p3 <- ggplot(df_mean, aes(x = factor(n), y = mean)) + 
  geom_pointrange(aes(ymin = mean - se, ymax = mean + se)) +
  scale_x_discrete(name = "n") +
  scale_y_continuous(breaks = c(-.2, 0, .2), name = "mean(x)") +
  theme_dviz_hgrid()

p1 + p2 + p3 + plot_layout(ncol = 1)

```


```{r fig.width = 5, eval = FALSE}
cows %>% filter(breed != "Canadian") %>%
  group_by(breed) %>%
  summarize(mean = mean(butterfat),
            se = sd(butterfat)/sqrt(n())) %>%
  mutate(breed = fct_reorder(breed, desc(mean)))-> cow_means

ggplot(cow_means, aes(x = breed, y = mean, ymin = mean - se, ymax = mean + se)) +
  geom_pointrange() +
  scale_x_discrete(labels = c("Jersey", "Guernsey", "Ayrshire", "Holstein-\nFriesian"),
                   name = NULL) +
  scale_y_continuous(name = "mean % butterfat", expand = c(0, 0)) +
  theme_dviz_hgrid()
```

```{r fig.width = 5, eval = FALSE}
ggplot(cow_means, aes(x = breed, y = mean, ymin = mean - se, ymax = mean + se)) +
  geom_col(fill = "gray70") +
  geom_linerange() +
  scale_x_discrete(labels = c("Jersey", "Guernsey", "Ayrshire", "Holstein-\nFriesian"),
                   name = NULL) +
  scale_y_continuous(name = "mean % butterfat", expand = c(0, 0)) +
  theme_dviz_hgrid()
```

```{block type='rmdtip', echo=TRUE}
Whenever you visualize uncertainty with error bars, you must specify what quantity and/or confidence level the error bars represent.
```


*Include a coefficient plot from a regression model, as in Cleveland's book?*

```{r}
# relevant materials: 
# Bowman "Graphs for Uncertainty": http://www.rss.org.uk/Images/PDF/events/2018/Bowman-5-Sept-2018.pdf
# R package denstrip: https://cran.r-project.org/package=denstrip

library(ggplot2)
library(rlang)
library(dplyr)
library(tidyr)

stat_conf_band <- function(mapping = NULL, data = NULL,
                           geom = "tile", position = "identity",
                           ...,
                           confidence = 0.95,
                           xlim = NULL,
                           n = 501,
                           na.rm = FALSE,
                           show.legend = FALSE,
                           inherit.aes = TRUE) {
  layer(
    data = data,
    mapping = mapping,
    stat = StatConfBand,
    geom = geom,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(
      confidence = confidence,
      n = n,
      na.rm = na.rm,
      xlim = xlim,
      ...
    )
  )
}

fit_normal <- function(mean, moe, confidence = 0.95) {
  # convert to two-tailed value
  confidence <- 1-(1-confidence)/2
  function(x) dnorm(x, mean = mean, sd = moe/qnorm(confidence))
}

StatConfBand <- ggproto("StatConfBand", Stat,
  required_aes = c("mean", "moe"),                        
  default_aes = aes(fill = stat(ndensity)),
                        
  compute_group = function(data, scales, confidence = 0.95, xlim = NULL, n = 501) {
    # Check that confidence band parameters are constant within group
    params <- unique(data[c("mean", "moe")])
    if (nrow(params) > 1) {
      stop("Confidence band parameters can not vary within data groups", call. = FALSE)
    }
    params <- c(as.list(params), list(confidence = confidence))
    
    range <- xlim %||% scales$x$dimension()
    xseq <- seq(range[1], range[2], length.out = n)
    
    if (scales$x$is_discrete()) {
      x_trans <- xseq
    } else {
      # For continuous scales, need to back transform from transformed range
      # to original values
      x_trans <- scales$x$trans$inverse(xseq)
    }
    
    fun <- do.call(fit_normal, params)
    density <- fun(x_trans)
    
    data.frame(
      x = xseq,
      density = density,
      ndensity = density/max(density)
    )
  }
)

df_in <- data.frame(
  group = letters[1:3],
  mean = c(1, 3, 2),
  sd = c(.8, .4, .7)
)

df_data <- mutate(df_in, 
    value = purrr::map2(mean, sd, ~rnorm(250, .x, .y))
  ) %>%
  unnest()

df_out <- group_by(df_data, group) %>%
  summarize(
    mean = mean(value),
    sd = sd(value),
    moe = sd*1.96
  )

ggplot(df_out, aes(x = mean, y = group)) +
  stat_conf_band(aes(mean = mean, moe = sd), height = 0.8, confidence = 0.67) +
  geom_point(data = df_data, aes(x = value), position = position_jitter(width = 0), size = 1) +
  geom_errorbarh(aes(xmin = mean - sd, xmax = mean + sd), height = 0.2, color = "darkred", size = 1) +
  geom_point(size = 3, color = "darkred") +
  scale_fill_gradient(low = "#132B4300", high = "#56B1F7FF") +
  theme_minimal()
```

```{r fig.width = 5.5, fig.asp = 3/4}
sample_posterior_preds <- function(model, samples = 20, n = 100, unconditional = TRUE) {
  predictor <- model$pred.formula[[2]]
  response <- model$terms[[2]]
   
  # make a tibble with a sequence of predictor values
  pred_seq <- tibble(
    !!predictor := seq(
      min(model$model[[predictor]]),
      max(model$model[[predictor]]), 
      length.out = n
    )
  )
  
  # Get the linear prediction matrix
  pred_mat <- predict(
    model,
    newdata  = pred_seq,
    type = "lpmatrix",
    unconditional = unconditional
  )

  # Get the variance-covariance matrix of coefficients
  vcov_mat <- vcov(model, unconditional = unconditional)

  # Draw 20 samples from the posterior and make predictions from them
  coefs <- rmvnorm(samples, mean = coef(model), sigma = vcov_mat)
  preds <- pred_mat %*% t(coefs)
  pred_df <- as_tibble(preds) %>%
    set_names(as.character(1:samples)) %>%
    cbind(pred_seq) %>%
    gather(sample, !!response, -!!predictor)
  
  pred_df
}

# Get the smoothing-uncertainty corrected confidence intervals
confidence_band <- function(model, level = 0.95, n = 100, unconditional = TRUE) {
  predictor <- model$pred.formula[[2]]
  response <- model$terms[[2]]
   
  # make a tibble with a sequence of predictor values
  pred_seq <- tibble(
    !!predictor := seq(
      min(model$model[[predictor]]),
      max(model$model[[predictor]]), 
      length.out = n
    )
  )

  # normal quantile corresponding to confidence level
  std <- stats::qnorm(level / 2 + 0.5)

  # predict confidence band
  predict(
    model,
    newdata = pred_seq,
    se.fit = TRUE,
    unconditional = unconditional
  ) %>%
    as_tibble() %>%
    cbind(pred_seq) %>%
    rename(!!response := fit) %>%
    mutate(
      lo = !!response - std*se.fit,
      hi = !!response + std*se.fit
    )
}


blue_jays_male <- filter(blue_jays, KnownSex == "M")

fit <- gam(Head ~ Mass, data = blue_jays_male, method = "REML")
sample_df <- sample_posterior_preds(fit, 20)
ci_df <- confidence_band(fit)

ggplot(blue_jays_male, aes(Mass, Head)) + 
  geom_ribbon(data = ci_df, aes(ymin = lo, ymax = hi), fill="#80808080", color = NA) +
  geom_point(color = "#0072B2", size = 1.5) +
  geom_line(data = sample_df, aes(group = sample), color = "blue", size = 0.3) +
  geom_line(data = ci_df, color = "red", size = 0.5) +
  scale_x_continuous(
    limits = c(59, 82),
    expand = c(0, 0),
    name = "body mass (g)") +
  scale_y_continuous(
    limits = c(52, 61),
    expand = c(0, 0),
    name = "head length (mm)"
  ) +
  theme_dviz_open()

```

```{r tank-capacity-uncertain, fig.width = 8.5, fig.asp = 3/8, fig.cap='(ref:tank-capacity-uncertain)'}
cars93 <- MASS::Cars93

set.seed(8692282)

fit <- gam(Fuel.tank.capacity ~ s(Price, k = 5, bs = 'cr'), data = cars93)
#fit <- gam(Fuel.tank.capacity ~ s(Price, k = 6, bs = 'gp'), data=cars93)
#fit <- gam(Fuel.tank.capacity ~ s(Price), data = cars93, method = "REML")
sample_df <- sample_posterior_preds(fit, 10, unconditional = FALSE)
ci_df <- confidence_band(fit, unconditional = FALSE)

cars_base <- ggplot(cars93, aes(x = Price, y = Fuel.tank.capacity)) + 
  scale_x_continuous(
    name = "price (USD)",
    breaks = c(20, 40, 60),
    labels = c("$20,000", "$40,000", "$60,000")
  ) +
  scale_y_continuous(name = "fuel-tank capacity\n(US gallons)") +
  theme_minimal_grid(12)

p1 <- cars_base +
  geom_ribbon(data = ci_df, aes(ymin = lo, ymax = hi), fill="grey70", color = NA, alpha = 1/2) +
  geom_point(color = "grey60") +
  geom_line(data = ci_df, color = "#0072B2", size = 1)

p2 <- cars_base +
  geom_ribbon(data = ci_df, aes(ymin = lo, ymax = hi), fill="grey70", color = NA, alpha = 1/2) +
  geom_point(color = "grey60") +
  geom_line(data = sample_df, aes(group = sample), color = "#0072B2", size = 0.3)

plot_grid(
  p1, p2, align = 'hv',
  labels = 'auto'
)

```


## Methods of uncertainty estimates


Table: (\#tab:estimation-frameworks) Comparisons of different approaches to parameter and uncertainty estimation.

------------------------------------------------------------------------------------------
approach       interpretability    computational       complexity of     assumptions on
                of estimates         efficiency        modeling setup       data set
------------ ------------------ ------------------- ------------------- ------------------
Bayesian          high                 low                 high                few

frequentist       moderate              high              moderate            several
parametric  

bootstrap           low                moderate             low                none
------------------------------------------------------------------------------------------


## Hypothetical outcomes plots

Hypothetical outcomes plots: https://medium.com/hci-design-at-uw/hypothetical-outcomes-plots-experiencing-the-uncertain-b9ea60d7c740

## Value-suppressing uncertainty palette

