```{r echo = FALSE, message = FALSE, warning = FALSE}
# run setup script
source("_common.R")

library(forcats)
library(patchwork)
library(lubridate)
library(mgcv)
library(mvtnorm)
library(rlang)
library(tidyr)
```

# Visualizing uncertainty {#visualizing-uncertainty}

*Chapter under construction. Tentative outline:*

- *Error bars and confidence bands*
- *Distributions*
- *Hypothetical outcomes plots*
- *Value suppression*


## Error bars and confidence bands

```{r fig.asp = 1.2}
n <- c(10, 30, 100, 300, 1000)

df <- data.frame(n = unlist(lapply(n, function(x) rep(x, x))),
                 x = c(rnorm(sum(n))))

df %>% group_by(n) %>%
  summarize(mean = mean(x),
            se = sd(x)/sqrt(n())) -> df_mean

p1 <- ggplot(df, aes(x = factor(n), y = x)) + 
  geom_point(size = 0.5, position = position_jitter(width = 0.3)) +
  scale_x_discrete(breaks = NULL, name = NULL) +
  scale_y_continuous(breaks = c(-2, 0, 2)) +
  theme_dviz_hgrid() +
  theme(plot.margin = margin(3, 0, 14, 0))
p2 <- ggplot(df, aes(x = factor(n), y = x)) + 
  geom_boxplot(fill = "gray90") +
  scale_x_discrete(breaks = NULL, name = NULL) +
  scale_y_continuous(breaks = c(-2, 0, 2)) +
  theme_dviz_hgrid() +
  theme(plot.margin = margin(3, 0, 14, 0))
p3 <- ggplot(df_mean, aes(x = factor(n), y = mean)) + 
  geom_pointrange(aes(ymin = mean - se, ymax = mean + se)) +
  scale_x_discrete(name = "n") +
  scale_y_continuous(breaks = c(-.2, 0, .2), name = "mean(x)") +
  theme_dviz_hgrid()

p1 + p2 + p3 + plot_layout(ncol = 1)

```


```{r fig.width = 5}
cows %>% filter(breed != "Canadian") %>%
  group_by(breed) %>%
  summarize(mean = mean(butterfat),
            se = sd(butterfat)/sqrt(n())) %>%
  mutate(breed = fct_reorder(breed, desc(mean)))-> cow_means

ggplot(cow_means, aes(x = breed, y = mean, ymin = mean - se, ymax = mean + se)) +
  geom_pointrange() +
  scale_x_discrete(labels = c("Jersey", "Guernsey", "Ayrshire", "Holstein-\nFriesian"),
                   name = NULL) +
  scale_y_continuous(name = "mean % butterfat", expand = c(0, 0)) +
  theme_dviz_hgrid()
```

```{r fig.width = 5}
ggplot(cow_means, aes(x = breed, y = mean, ymin = mean - se, ymax = mean + se)) +
  geom_col(fill = "gray70") +
  geom_linerange() +
  scale_x_discrete(labels = c("Jersey", "Guernsey", "Ayrshire", "Holstein-\nFriesian"),
                   name = NULL) +
  scale_y_continuous(name = "mean % butterfat", expand = c(0, 0)) +
  theme_dviz_hgrid()
```

*Include a coefficient plot from a regression model, as in Cleveland's book?*

```{r}
# relevant materials: 
# Bowman "Graphs for Uncertainty": http://www.rss.org.uk/Images/PDF/events/2018/Bowman-5-Sept-2018.pdf
# R package denstrip: https://cran.r-project.org/package=denstrip

library(ggplot2)
library(rlang)
library(dplyr)
library(tidyr)

stat_conf_band <- function(mapping = NULL, data = NULL,
                           geom = "tile", position = "identity",
                           ...,
                           confidence = 0.95,
                           xlim = NULL,
                           n = 501,
                           na.rm = FALSE,
                           show.legend = FALSE,
                           inherit.aes = TRUE) {
  layer(
    data = data,
    mapping = mapping,
    stat = StatConfBand,
    geom = geom,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(
      confidence = confidence,
      n = n,
      na.rm = na.rm,
      xlim = xlim,
      ...
    )
  )
}

fit_normal <- function(mean, moe, confidence = 0.95) {
  # convert to two-tailed value
  confidence <- 1-(1-confidence)/2
  function(x) dnorm(x, mean = mean, sd = moe/qnorm(confidence))
}

StatConfBand <- ggproto("StatConfBand", Stat,
  required_aes = c("mean", "moe"),                        
  default_aes = aes(fill = stat(ndensity)),
                        
  compute_group = function(data, scales, confidence = 0.95, xlim = NULL, n = 501) {
    # Check that confidence band parameters are constant within group
    params <- unique(data[c("mean", "moe")])
    if (nrow(params) > 1) {
      stop("Confidence band parameters can not vary within data groups", call. = FALSE)
    }
    params <- c(as.list(params), list(confidence = confidence))
    
    range <- xlim %||% scales$x$dimension()
    xseq <- seq(range[1], range[2], length.out = n)
    
    if (scales$x$is_discrete()) {
      x_trans <- xseq
    } else {
      # For continuous scales, need to back transform from transformed range
      # to original values
      x_trans <- scales$x$trans$inverse(xseq)
    }
    
    fun <- do.call(fit_normal, params)
    density <- fun(x_trans)
    
    data.frame(
      x = xseq,
      density = density,
      ndensity = density/max(density)
    )
  }
)

df_in <- data.frame(
  group = letters[1:3],
  mean = c(1, 3, 2),
  sd = c(.8, .4, .7)
)

df_data <- mutate(df_in, 
    value = purrr::map2(mean, sd, ~rnorm(250, .x, .y))
  ) %>%
  unnest()

df_out <- group_by(df_data, group) %>%
  summarize(
    mean = mean(value),
    sd = sd(value),
    moe = sd*1.96
  )

ggplot(df_out, aes(x = mean, y = group)) +
  stat_conf_band(aes(mean = mean, moe = sd), height = 0.8, confidence = 0.67) +
  geom_point(data = df_data, aes(x = value), position = position_jitter(width = 0), size = 1) +
  geom_errorbarh(aes(xmin = mean - sd, xmax = mean + sd), height = 0.2, color = "darkred", size = 1) +
  geom_point(size = 3, color = "darkred") +
  scale_fill_gradient(low = "#132B4300", high = "#56B1F7FF") +
  theme_minimal()
```

```{r fig.width = 5.5, fig.asp = 3/4}
sample_posterior_preds <- function(model, samples = 20, n = 100, unconditional = TRUE) {
  predictor <- model$pred.formula[[2]]
  response <- model$terms[[2]]
   
  # make a tibble with a sequence of predictor values
  pred_seq <- tibble(
    !!predictor := seq(
      min(model$model[[predictor]]),
      max(model$model[[predictor]]), 
      length.out = n
    )
  )
  
  # Get the linear prediction matrix
  pred_mat <- predict(
    model,
    newdata  = pred_seq,
    type = "lpmatrix",
    unconditional = unconditional
  )

  # Get the variance-covariance matrix of coefficients
  vcov_mat <- vcov(model, unconditional = unconditional)

  # Draw 20 samples from the posterior and make predictions from them
  coefs <- rmvnorm(samples, mean = coef(model), sigma = vcov_mat)
  preds <- pred_mat %*% t(coefs)
  pred_df <- as_tibble(preds) %>%
    set_names(as.character(1:samples)) %>%
    cbind(pred_seq) %>%
    gather(sample, !!response, -!!predictor)
  
  pred_df
}

# Get the smoothing-uncertainty corrected confidence intervals
confidence_band <- function(model, level = 0.95, n = 100, unconditional = TRUE) {
  predictor <- model$pred.formula[[2]]
  response <- model$terms[[2]]
   
  # make a tibble with a sequence of predictor values
  pred_seq <- tibble(
    !!predictor := seq(
      min(model$model[[predictor]]),
      max(model$model[[predictor]]), 
      length.out = n
    )
  )

  # normal quantile corresponding to confidence level
  std <- stats::qnorm(level / 2 + 0.5)

  # predict confidence band
  predict(
    model,
    newdata = pred_seq,
    se.fit = TRUE,
    unconditional = unconditional
  ) %>%
    as_tibble() %>%
    cbind(pred_seq) %>%
    rename(!!response := fit) %>%
    mutate(
      lo = !!response - std*se.fit,
      hi = !!response + std*se.fit
    )
}


blue_jays_male <- filter(blue_jays, KnownSex == "M")

fit <- gam(Head ~ Mass, data = blue_jays_male, method = "REML")
sample_df <- sample_posterior_preds(fit, 20)
ci_df <- confidence_band(fit)

ggplot(blue_jays_male, aes(Mass, Head)) + 
  geom_ribbon(data = ci_df, aes(ymin = lo, ymax = hi), fill="#80808080", color = NA) +
  geom_point(color = "#0072B2", size = 1.5) +
  geom_line(data = sample_df, aes(group = sample), color = "blue", size = 0.3) +
  geom_line(data = ci_df, color = "red", size = 0.5) +
  scale_x_continuous(
    limits = c(59, 82),
    expand = c(0, 0),
    name = "body mass (g)") +
  scale_y_continuous(
    limits = c(52, 61),
    expand = c(0, 0),
    name = "head length (mm)"
  ) +
  theme_dviz_open()

```

```{r tank-capacity-uncertain, fig.width = 8.5, fig.asp = 3/8, fig.cap='(ref:tank-capacity-uncertain)'}
cars93 <- MASS::Cars93

set.seed(8692282)

fit <- gam(Fuel.tank.capacity ~ s(Price, k = 5, bs = 'cr'), data = cars93)
#fit <- gam(Fuel.tank.capacity ~ s(Price, k = 6, bs = 'gp'), data=cars93)
#fit <- gam(Fuel.tank.capacity ~ s(Price), data = cars93, method = "REML")
sample_df <- sample_posterior_preds(fit, 10, unconditional = FALSE)
ci_df <- confidence_band(fit, unconditional = FALSE)

cars_base <- ggplot(cars93, aes(x = Price, y = Fuel.tank.capacity)) + 
  scale_x_continuous(
    name = "price (USD)",
    breaks = c(20, 40, 60),
    labels = c("$20,000", "$40,000", "$60,000")
  ) +
  scale_y_continuous(name = "fuel-tank capacity\n(US gallons)") +
  theme_minimal_grid(12)

p1 <- cars_base +
  geom_ribbon(data = ci_df, aes(ymin = lo, ymax = hi), fill="grey70", color = NA, alpha = 1/2) +
  geom_point(color = "grey60") +
  geom_line(data = ci_df, color = "#0072B2", size = 1)

p2 <- cars_base +
  geom_ribbon(data = ci_df, aes(ymin = lo, ymax = hi), fill="grey70", color = NA, alpha = 1/2) +
  geom_point(color = "grey60") +
  geom_line(data = sample_df, aes(group = sample), color = "#0072B2", size = 0.3)

plot_grid(
  p1, p2, align = 'hv',
  labels = 'auto'
)

```

```{r}
pima <- rbind(MASS::Pima.te, MASS::Pima.tr)

fit <- glm(type ~ npreg + glu + bp + skin + bmi + age, data = pima, family = "binomial")
```

## Distributions


## Hypothetical outcomes plots

Hypothetical outcomes plots: https://medium.com/hci-design-at-uw/hypothetical-outcomes-plots-experiencing-the-uncertain-b9ea60d7c740

## Value-suppressing uncertainty palette


## Contour lines (Staging for overlapping points chapter)

Instead of binning data points into rectangles or hexagons, we can also estimate the point density across the plot area and indicate regions of different point densities with contour lines. This technique works well when the point density changes slowly across both the *x* and the *y* dimensions.

As an example for this approach, we consider the relationship between population number and area for counties in the Midwest. We have data for 1055 counties, and a scatter plot looks like a cloud of points (Figure \@ref(fig:midwest-scatter)). We can highlight the distribution of points more clearly by making them very small and partially transparent and plotting them on top of contour lines that delineate regions of comparable point density (Figure \@ref(fig:midwest-density-dots)). We can also plot just the contour lines, without the individual points (Figure \@ref(fig:midwest-density-smooth)). In this case, it can be helpful to add a trendline that shows the overall trend in the data. Here, there isn't much of a trend, and the shape of the trendline (approximately flat) reflects this lack of a trend.

(ref:midwest-scatter) Population versus area for counties in midwestern states. Data are taken from the 2010 US census and are shown for 1055 counties covering 12 states. Each dot represents one county. 

```{r midwest-scatter, fig.asp = 0.8, fig.cap = '(ref:midwest-scatter)', eval = FALSE}
county_midwest <- left_join(US_census, US_regions) %>%
  filter(region == "Midwest")

p_midwest_base <- ggplot(county_midwest, aes(area, pop2010)) + 
  scale_fill_gradient(low = "grey70", high = "grey30", guide = "none") +
  scale_y_log10(breaks = c(1000, 1e4, 1e5, 1e6, 1e7),
                limits = c(450, 5.2e6),
                labels = label_log10) +
  scale_x_log10(breaks = c(100, 300, 1000, 3000),
                limits = c(50, 7000), expand = c(0, 0)) +
  ylab("population total") +
  xlab("area (square miles)") +
  theme_dviz_grid()

p_midwest_base + geom_point(color = "navy", size = 1, alpha = .7)
```

(ref:blue-jays-contour)

```{r blue-jays-contour, fig.width = 6, fig.asp = 3/4, fig.cap='(ref:blue-jays-contour)'}

blue_jays_base <- ggplot(blue_jays, aes(Mass, Head)) + 
  scale_x_continuous(
    limits = c(57, 82),
    expand = c(0, 0),
    name = "body mass (g)") +
  scale_y_continuous(
    limits = c(49, 61),
    expand = c(0, 0),
    name = "head length (mm)"
  ) +
  theme_dviz_grid()

blue_jays_base + 
  stat_density_2d(color = "black", size = 0.4, binwidth = 0.004) +
  geom_point(color = "black", size = 1.5, alpha = 1/3)
```

```{r blue-jays-contour-filled, fig.width = 6, fig.asp = 3/4, fig.cap='(ref:blue-jays-contour-filled)'}
blue_jays_base + 
  stat_density_2d(aes(fill = ..level..), geom = "polygon", color = "black", size = 0.15, binwidth = 0.004) +
  geom_point(color = "black", size = 1.5, alpha = .4) +
  scale_fill_gradient(low = "grey95", high = "grey70", guide = "none")
```

```{r blue-jays-contour-by-sex, fig.width = 6, fig.asp = 3/4, fig.cap='(ref:blue-jays-contour-by-sex)'}
blue_jays_base + 
  aes(color = KnownSex) +
  stat_density_2d(size = 0.4, binwidth = 0.006) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 2, linetype = 0)
    )
  ) +
  theme_dviz_grid() +
  theme(
    legend.position = c(1, 0),
    legend.justification = c(1, 0),
    #legend.position = "top",
    #legend.justification = "right",
    #legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    #legend.key.width = unit(10, "pt")
    axis.ticks.length = unit(0, "pt"),
    axis.ticks = element_blank()
  )
```


(ref:midwest-density-dots) Population versus area for counties in midwestern states. Contour lines and shaded areas indicate the density of counties for that combination of population total and area. Individual counties are shown as light blue dots.

```{r midwest-density-dots, fig.asp = 0.8, fig.cap = '(ref:midwest-density-dots)', eval = FALSE}
p_midwest_base + 
  stat_density_2d(aes(fill = ..level..), geom = "polygon", color = "black", size = 0.2, alpha = 0.5) +
  geom_point(color = "navy", size = .5, alpha = .4)
```

(ref:midwest-density-smooth) Population versus area for counties in midwestern states. Contour lines and shaded areas indicate the density of counties for that combination of population total and area. Note that some counties lie outside the largest shaded area. The solid blue line highlights the mean relationship between population total and county area. It was obtained via least-square fitting of a general additive model with cubic spline base to the underlying data.

```{r midwest-density-smooth, fig.asp = 0.8, fig.cap = '(ref:midwest-density-smooth)', eval = FALSE}
p_midwest_base + 
  stat_density_2d(aes(fill = ..level..), geom = "polygon", color = "black", size = 0.2, alpha = 0.5) +
  geom_smooth(color = "navy", se = FALSE, size = 0.5)
```


```{r eval = FALSE}
# As an alternative plot with contour lines, show blue jays data as contours (no sex), then with sex (2 contour colours), then diamonds as contours (labelled bad), then contours faceted? And then maybe the following plot?

ggplot(diamonds, aes(carat, price)) + 
  geom_point(size = 0.1, color = "#E2E6CA", alpha = 0.5, na.rm = TRUE) + 
  geom_density2d(color = "#035F33", size = 0.25, na.rm = TRUE) +
  scale_x_continuous(limits = c(0, 4.2), expand = c(0, 0.1)) +
  scale_y_log10(limits = c(300, 20000),
                breaks = c(300, 1000, 3000, 10000)) +
  theme_dviz_grid()
```

```{r diamonds-points, fig.asp = 3/4, fig.cap = '(ref:diamonds-points)'}
p <- ggplot(diamonds, aes(carat, price, color = cut)) + 
  geom_point(size = .2, alpha = 1/5) +
  scale_x_continuous(
    limits = c(-1, 5.1)
  ) +
  scale_y_log10(
    breaks = c(300, 1000, 3000, 10000)
  ) +
  scale_color_discrete_sequential(
    palette = "Inferno",
    nmax = 6,
    order = 1:5,
    breaks = c("Ideal", "Premium", "Very Good", "Good", "Fair"),
    labels = c("ideal", "premium", "very good", "good", "fair"),
    guide = guide_legend(
      override.aes = list(size = 2, alpha = 1)
    )
  ) +
  coord_cartesian(xlim = c(-.1, 3.2), ylim = c(240, 25000), expand = FALSE) + 
  theme_dviz_grid() +
  panel_border() +
  theme(
    plot.margin = margin(18, 7, 1, 0),
    legend.key.width = unit(6, "pt"),
    legend.spacing.y = unit(3, "pt"),
    legend.title = element_text(hjust = 0.5, margin = margin(0, 0, 0, 0)),
    legend.position = c(.97, .3),
    legend.justification = c(1, 0.5),
    legend.box.margin = margin(7, 7, 7, 7),
    legend.box.background = element_rect(fill = "white", color = NA),
    axis.ticks.length = unit(0, "pt")
  )

stamp_bad(p)
```

```{r diamonds-contour-colors, fig.asp = 3/4}
p <- ggplot(diamonds, aes(carat, price, color = cut)) + 
  geom_density2d(size = .35, binwidth = 0.8) +
  scale_x_continuous(
    limits = c(-1, 5.1)
  ) +
  scale_y_log10(
    breaks = c(300, 1000, 3000, 10000)
  ) +
  scale_color_discrete_sequential(
    palette = "Inferno",
    nmax = 6,
    order = 1:5,
    breaks = c("Ideal", "Premium", "Very Good", "Good", "Fair"),
    labels = c("ideal", "premium", "very good", "good", "fair"),
    guide = guide_legend(
      override.aes = list(size = 0.5)
    )
  ) +
  coord_cartesian(xlim = c(-.1, 2.3), ylim = c(240, 25000), expand = FALSE) + 
  theme_dviz_grid() +
  panel_border() +
  theme(
    plot.margin = margin(18, 7, 1, 0),
    legend.spacing.y = unit(3, "pt"),
    legend.title = element_text(hjust = 0.5, margin = margin(0, 0, 0, 0)),
    legend.position = c(.97, .3),
    legend.justification = c(1, 0.5),
    legend.box.margin = margin(7, 7, 7, 7),
    legend.box.background = element_rect(fill = "white", color = NA),
    axis.ticks.length = unit(0, "pt")
  )

stamp_bad(p)
```

```{r diamonds-contour-facets, fig.asp = 3/4}
ggplot(diamonds, aes(carat, price)) + 
  geom_density2d(color = darken("#0072B2", .2), size = .3, binwidth = 0.8) +
  scale_x_continuous(
    limits = c(-1, 5.1)
  ) +
  scale_y_log10(
    breaks = c(300, 1000, 3000, 10000)
  ) +
  coord_cartesian(xlim = c(-.1, 2.3), ylim = c(200, 25000), expand = FALSE) + 
  facet_wrap(~cut, scales = "free_x", labeller = labeller(cut = tolower)) +
  theme_dviz_grid() +
  panel_border() +
  theme(
    legend.title = element_text(hjust = 0.5),
    legend.position = c(.95, .05),
    legend.justification = c(1, 0),
    axis.ticks.length = unit(0, "pt")
  )



# ggplot(diamonds, aes(carat, price)) + 
#   geom_bin2d(bins = 50) +
#   scale_x_continuous(limits = c(0, 4.2), expand = c(0, 0.1)) +
#   scale_y_log10(limits = c(300, 20000),
#                 breaks = c(300, 1000, 3000, 10000)) +
#   scale_fill_continuous_sequential("Green-Yellow", rev = TRUE, begin = 0) +
#   theme_dviz_grid() +
#   theme(legend.position = c(0.9, 0.4),
#         legend.justification = c(1, 0.5),
#         legend.box.background = element_rect(fill = "white",
#                                              color = NA),
#         legend.box.margin = margin(3.5, 3.5, 3.5, 3.5))
```



