```{r echo = FALSE, message = FALSE}
# run setup script
source("_common.R")
```

# Visualizing trends {#visualizing-trends}

Topics to cover:

- linear fits
- non-linear fits
- exponential growth
- averaging/detrending? Use `CO2` dataset?

(ref:blue-jays-scatter-line) Head length versus body mass for 123 blue jays. The birds' sex is indicated by color. At the same body mass, male birds tend to have longer heads (and specifically, longer bills) than female birds. Data source: Keith Tarvin, Oberlin College

```{r blue-jays-scatter-line, fig.width = 5, fig.asp = 3/4, fig.cap='(ref:blue-jays-scatter-line)'}
ggplot(blue_jays, aes(Mass, Head, color = KnownSex, fill = KnownSex)) + 
  geom_point(pch = 21, color = "white", size = 2.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(name = "body mass (g)") +
  scale_y_continuous(name = "head length (mm)") +
  scale_fill_manual(
    values = c(F = "#D55E00B0", M = "#0072B2B0"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(direction = "horizontal")) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(direction = "horizontal")) +
  theme_dviz_grid() +
  theme(
    legend.position = c(1, 0.01),
    legend.justification = c(1, 0),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(20, "pt")
  )
```



(ref:tank-capacity-smoothers) **Figure title.** (a) LOESS smoother. (b) Cubic regression splines with 5 knots. (c) Thin-plate regression spline with 3 knots. (d)  Gaussian process spline with 6 knots.

```{r tank-capacity-smoothers, fig.width=8.5, fig.asp = 0.75, fig.cap='(ref:tank-capacity-smoothers)'}
cars93 <- MASS::Cars93

cars_base <- ggplot(cars93, aes(x = Price, y = Fuel.tank.capacity)) + geom_point(color = "grey60") + 
  scale_x_continuous(name = "price (USD)",
                     breaks = c(20, 40, 60),
                     labels = c("$20,000", "$40,000", "$60,000")) +
  scale_y_continuous(name = "fuel tank capacity\n(US gallons)") +
  theme_minimal_grid(12)  
  
p1 <- cars_base + geom_smooth(se = FALSE, method = "loess", formula = y ~ x, color = "#0072B2")
p2 <- cars_base + geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 5, bs = 'cr'), color = "#0072B2")
p3 <- cars_base + geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 3), color = "#0072B2")
p4 <- cars_base + geom_smooth(se = FALSE, method = "gam", formula = y ~ s(x, k = 6, bs = 'gp'), color = "#0072B2")

plot_grid(p1, p2, p3, p4, align = 'hv',
          labels = 'auto', label_fontface = "plain")

# See also: https://altaf-ali.github.io/ggplot_tutorial/smoother.html
```


(ref:tank-capacity-model) **Figure title.** The solid line corresponds to the formula $y = 68.20 - 46.37 x/(x - 1,560)$.

```{r tank-capacity-model, fig.asp = 0.75, fig.cap = '(ref:tank-capacity-model)'}
fit.out <- nls(Fuel.tank.capacity ~ a*Price/(Price + b) + c, data = cars93, start = c(a = -45, b = -1, c = 70))
fit.df <- data.frame(Price = 7:62,
                     Fuel.tank.capacity = predict(fit.out, data.frame(Price = 7:62)))

ggplot(cars93, aes(x = Price, y = Fuel.tank.capacity)) + geom_point(color = "grey60") + 
  geom_line(data = fit.df, size = 1, color = "#0072B2") +
  scale_x_continuous(name = "price (USD)",
                     breaks = c(20, 40, 60),
                     labels = c("$20,000", "$40,000", "$60,000")) +
  scale_y_continuous(name = "fuel tank capacity\n(US gallons)") +
  theme_minimal_grid()
```
